---
title: "Coral Photosynthetic Efficiency — GAMM Analysis"
author: "Michael Studivan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# -----------------------------------------------------------------------------
# Global chunk options
# echo = TRUE:     show code in rendered output
# warning = FALSE: suppress warnings in rendered document
#                  (warnings still print to console during knitting)
# message = FALSE: suppress package loading messages
# -----------------------------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 300
)
```

```{r packages}
# -----------------------------------------------------------------------------
# Load all required packages
# -----------------------------------------------------------------------------
library(mgcv)      # bam() for generalized additive mixed models
library(dplyr)     # data wrangling
library(ggplot2)   # plotting
library(gratia)    # ggplot2-based visualization of mgcv models
library(emmeans)   # estimated marginal means and post-hoc contrasts
library(tibble)    # rownames_to_column() for tidy stats export
library(writexl)   # export stats tables to Excel
library(patchwork) # combining multiple ggplot panels (used by gratia::draw)
library(ggpubr)    # theme_pubr() for publication-ready plot styling
```

---

# Data import and preparation

```{r data-import}
# -----------------------------------------------------------------------------
# Import master iPAM dataset and genotype lookup table
# -----------------------------------------------------------------------------
ipam <- read.csv(
  file = "../../data/photosynthesis/urban ipam.csv",
  head = TRUE
)

genotypes <- read.csv(
  file = "../../data/genotype lookup table.csv",
  head = TRUE
)

# Join genotype lookup — makes genotype labels distinct across sites and species
ipam <- ipam %>%
  inner_join(genotypes, by = "sampleID")
```

```{r data-prep}
# -----------------------------------------------------------------------------
# Remove baseline timepoints.
# The model response is raw FvFm, not the ratio column — ratio is used here
# only as a marker to identify baseline rows, since ratio == 1 exclusively at
# the initial measurement (the timepoint all subsequent ratios are expressed
# relative to).
#
# "Initial" is excluded rather than treated as an early point in the time
# series: it was measured over a month before the first post-treatment
# timepoint (time == 0), even though fragments were already assigned to
# treatment tanks by then. That gap is too large to represent as a
# continuous point on the same time axis without fabricating an elapsed-time
# value, so the model is fit on post-baseline (time >= 0) observations only.
# -----------------------------------------------------------------------------
ipam <- subset(ipam, ratio != "1")

# -----------------------------------------------------------------------------
# Code factors with explicit reference level ordering.
# The first level listed becomes the reference in all model contrasts:
#   site: "Emerald"      — expected non-resilient reference site
#   pH:   "control pH"   — ambient pH reference
#   temp: "control temp" — ambient temperature reference
#   treatment: "CC"      — contemporary pH, ambient temperature reference
#
# geno_id:  unique genotype identifier nested within site and species
# exp_unit: unique coral fragment identifier and the repeated measures subject.
#           Combines sampleID with temperature treatment because the same
#           physical fragment may appear in both temperature conditions.
# -----------------------------------------------------------------------------
ipam <- ipam %>%
  mutate(
    site     = factor(site,      levels = c("Emerald", "Rainbow", "Star", "MacN")),
    pH       = factor(pH,        levels = c("control pH",   "low pH")),
    temp     = factor(temp,      levels = c("control temp", "high temp")),
    treatment= factor(treatment, levels = c("CC", "LC", "CH", "LH")),
    time     = as.numeric(time),
    geno_id  = interaction(site, geno,     drop = TRUE),
    exp_unit = interaction(sampleID, temp, drop = TRUE)
  )

str(ipam)
```

```{r species-subsets}
# -----------------------------------------------------------------------------
# Subset by species
# -----------------------------------------------------------------------------
ipam_ofav <- subset(ipam, species == "Ofav")
ipam_ssid <- subset(ipam, species == "Ssid")

str(ipam_ofav)
str(ipam_ssid)
```

```{r treatment-subsets}
# -----------------------------------------------------------------------------
# Subset by experimental phase
#
# Control datasets: timepoints < 21 days (pre-heat-stress phase)
#   Both pH treatments present; control temperature treatment present
#
# Heat datasets: high temperature treatment only, all timepoints
#   Tests whether pH effects on FvFm differ under thermal stress
# -----------------------------------------------------------------------------

# O. faveolata
ipam_ofav %>% filter(time < 21)              -> ipam_ofav_control
ipam_ofav %>% filter(temp != "control temp") -> ipam_ofav_heat

# S. siderea
ipam_ssid %>% filter(time < 21)              -> ipam_ssid_control
ipam_ssid %>% filter(temp != "control temp") -> ipam_ssid_heat
```

```{r sanity-checks}
# -----------------------------------------------------------------------------
# Sanity checks — confirm data structure is correct before modelling
#
# Check 1: No exp_unit appears more than once at the same timepoint.
#          Duplicates indicate a data entry or merge error. Should be FALSE.
#
# Check 2: No sampleID maps to more than one geno_id.
#          Duplicates indicate a genotype lookup join error. Should be 0 rows.
# -----------------------------------------------------------------------------

cat("--- Duplicate exp_unit x time checks (should all be FALSE) ---\n")
cat("ofav_control:", any(duplicated(ipam_ofav_control %>% dplyr::select(exp_unit, time))), "\n")
cat("ssid_control:", any(duplicated(ipam_ssid_control %>% dplyr::select(exp_unit, time))), "\n")
cat("ofav_heat:   ", any(duplicated(ipam_ofav_heat    %>% dplyr::select(exp_unit, time))), "\n")
cat("ssid_heat:   ", any(duplicated(ipam_ssid_heat    %>% dplyr::select(exp_unit, time))), "\n")

cat("\n--- Duplicate sampleID -> geno_id checks (should all be 0 rows) ---\n")

check_geno_dups <- function(dat, label) {
  dupes <- dat %>%
    distinct(sampleID, geno_id) %>%
    count(sampleID) %>%
    filter(n > 1)
  cat(label, ":", nrow(dupes), "duplicated sampleIDs\n")
  if (nrow(dupes) > 0) print(dupes)
}

check_geno_dups(ipam_ofav_control, "ofav_control")
check_geno_dups(ipam_ssid_control, "ssid_control")
check_geno_dups(ipam_ofav_heat,    "ofav_heat")
check_geno_dups(ipam_ssid_heat,    "ssid_heat")
```

---

# Exploratory data analysis

Exploratory plots are examined before any modelling decisions to assess:

1. Whether the time trajectory is linear or nonlinear
2. Whether trajectories diverge across treatments and sites
3. Whether variance is homogeneous across groups and time
4. Whether site baseline differences justify treating site as a fixed effect
5. Whether residuals from a simple linear model show systematic patterns
   that motivate a GAMM

```{r eda-trajectories, fig.width=14, fig.height=8}
# -----------------------------------------------------------------------------
# Plot 1: Mean trajectories by species, site, and treatment
#
# Ribbons show ±1 SE across fragments. Overlaid loess smooths (dashed)
# show the approximate trajectory shape without imposing a parametric form.
# Curvature in loess lines = nonlinearity a linear model cannot capture.
# Diverging lines across treatment groups = treatment x time interactions.
# -----------------------------------------------------------------------------

p_eda_traj <- ipam %>%
  group_by(species, site, pH, temp, time) %>%
  summarise(mean_FvFm = mean(FvFm, na.rm = TRUE),
            se_FvFm   = sd(FvFm,   na.rm = TRUE) / sqrt(n()),
            .groups   = "drop") %>%
  ggplot(aes(x      = time,
             y      = mean_FvFm,
             colour = interaction(pH, temp, sep = " / "),
             fill   = interaction(pH, temp, sep = " / "),
             group  = interaction(pH, temp))) +
  geom_ribbon(aes(ymin = mean_FvFm - se_FvFm,
                  ymax = mean_FvFm + se_FvFm),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE,
              linewidth = 0.5, linetype = "dashed", alpha = 0.1) +
  facet_grid(species ~ site) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette   = "Dark2") +
  labs(x        = "Time (days)",
       y        = "Mean Fv/Fm \u00b1 SE",
       colour   = "pH / Temperature",
       fill     = "pH / Temperature",
       title    = "Raw FvFm trajectories by species, site, and treatment",
       subtitle = "Dashed loess curves show trajectory shape — curvature indicates nonlinearity") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_eda_traj

ggsave("../../outputs/photosynthesis/EDA_raw_trajectories.pdf",
       plot   = p_eda_traj,
       width  = 14,
       height = 8,
       device = pdf)
```

```{r eda-individual, fig.width=14, fig.height=8}
# -----------------------------------------------------------------------------
# Plot 2: Individual fragment trajectories
#
# Every exp_unit plotted as a thin line, overlaid with the group mean.
# High spread of thin lines = high between-fragment variance, which justifies
# the factor-smooth for exp_unit. If fragments follow very different trajectory
# shapes, a random intercept alone is insufficient — the GAMM factor-smooth
# is needed to capture individual-level trajectory variation.
# -----------------------------------------------------------------------------

p_eda_indiv <- ipam %>%
  ggplot(aes(x      = time,
             y      = FvFm,
             group  = exp_unit,
             colour = interaction(pH, temp, sep = " / "))) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  stat_summary(aes(group = interaction(pH, temp, sep = " / ")),
               fun       = mean,
               geom      = "line",
               linewidth = 1.2) +
  facet_grid(species ~ site) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x        = "Time (days)",
       y        = "Fv/Fm",
       colour   = "pH / Temperature",
       title    = "Individual fragment trajectories",
       subtitle = "Thin lines = individual exp_units; thick lines = group means") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_eda_indiv

ggsave("../../outputs/photosynthesis/EDA_individual_trajectories.pdf",
       plot   = p_eda_indiv,
       width  = 14,
       height = 8,
       device = pdf)
```

```{r eda-boxplots, fig.width=14, fig.height=8}
# -----------------------------------------------------------------------------
# Plot 3: FvFm distribution by timepoint and treatment
#
# Boxplots at each timepoint reveal whether variance changes over time
# (heteroscedasticity) or differs between treatment groups. Expanding
# box widths toward later timepoints, or markedly different spreads
# between treatments, would violate the Gaussian GAMM assumption of
# homogeneous residual variance.
# -----------------------------------------------------------------------------

p_eda_box <- ipam %>%
  ggplot(aes(x    = factor(time),
             y    = FvFm,
             fill = interaction(pH, temp, sep = " / "))) +
  geom_boxplot(position     = position_dodge(0.8),
               outlier.size = 1,
               alpha        = 0.7) +
  facet_grid(species ~ site) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x        = "Time (days)",
       y        = "Fv/Fm",
       fill     = "pH / Temperature",
       title    = "FvFm distribution by timepoint and treatment",
       subtitle = "Expanding variance over time would indicate heteroscedasticity") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_eda_box

ggsave("../../outputs/photosynthesis/EDA_boxplots_by_timepoint.pdf",
       plot   = p_eda_box,
       width  = 14,
       height = 8,
       device = pdf)
```

```{r eda-site-baseline, fig.width=10, fig.height=5}
# -----------------------------------------------------------------------------
# Plot 4: Site differences in starting FvFm
#
# Distribution of FvFm at the first post-baseline timepoint across sites.
# Differences here are biological signal reflecting inherent site resilience —
# they are not confounds to adjust away with a baseline covariate.
# This justifies treating site as a fixed effect rather than including
# baseline FvFm as a covariate, which would partial out the effect of interest.
# -----------------------------------------------------------------------------

p_eda_site <- ipam %>%
  group_by(species, site, exp_unit) %>%
  filter(time == min(time)) %>%
  ungroup() %>%
  ggplot(aes(x = site, y = FvFm, fill = site)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ species) +
  scale_fill_brewer(palette = "Set2") +
  labs(x        = "Site",
       y        = "Fv/Fm at first post-baseline timepoint",
       title    = "Site differences in starting FvFm",
       subtitle = "Differences reflect biological resilience signal — site is treated as a fixed effect") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

p_eda_site

ggsave("../../outputs/photosynthesis/EDA_site_baseline_differences.pdf",
       plot   = p_eda_site,
       width  = 10,
       height = 5,
       device = pdf)
```

```{r eda-lm-residuals, fig.width=14, fig.height=8}
# -----------------------------------------------------------------------------
# Plot 5: Residuals from a simple linear model
#
# Fits a basic lm ignoring repeated measures and plots residuals vs time.
# Systematic patterns — curvature in loess lines, fanning of points, or
# group-level trends — directly demonstrate why the linear model is
# misspecified and motivate the GAMM. This plot provides methods
# justification and is suitable for supplementary materials.
# -----------------------------------------------------------------------------

lm_simple <- lm(FvFm ~ treatment * site * time, data = ipam)

p_eda_resid <- ipam %>%
  mutate(resid_lm = resid(lm_simple)) %>%
  ggplot(aes(x      = time,
             y      = resid_lm,
             colour = interaction(pH, temp, sep = " / "))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  facet_grid(species ~ site) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x        = "Time (days)",
       y        = "Residual (observed \u2212 fitted)",
       colour   = "pH / Temperature",
       title    = "Residuals from a simple linear model",
       subtitle = "Non-random patterns confirm the linear model is misspecified") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_eda_resid

ggsave("../../outputs/photosynthesis/EDA_linear_model_residuals.pdf",
       plot   = p_eda_resid,
       width  = 14,
       height = 8,
       device = pdf)
```

---

# Helper functions

```{r helper-functions}
# -----------------------------------------------------------------------------
# All helper functions defined here and reused across all four datasets.
# Any change to model structure, plot aesthetics, or export settings
# propagates automatically to all four analyses.
# -----------------------------------------------------------------------------


# --- add_ar_start() ----------------------------------------------------------
# Creates the AR.start indicator required by bam() for AR(1) correction.
# Must be called AFTER subsetting because the first timepoint must be
# identified within each subset independently — a fragment's first timepoint
# in the heat subset may differ from its first in the full dataset.
#
# The combined treatment factor (CC, LC, CH, LH) is assumed to already exist
# as a column in the data from the ipam pipeline — it is not created here.
#
# Prints the treatment levels present in this subset to console as a
# sanity check — control subsets should show CC, LC, CH, LH; heat subsets
# should show only CH and LH.
#
# Arguments:
#   dat: subsetted data frame for one species/treatment combination
# Returns:
#   same data frame with ar_start column added (TRUE at first timepoint
#   per exp_unit, FALSE at all subsequent timepoints)

add_ar_start <- function(dat) {

  # Confirm which treatment levels are present in this subset.
  # droplevels() removes empty factor levels left over from subsetting
  # so only levels actually present in this subset are printed.
  cat("  Treatment levels in this subset:\n")
  print(levels(droplevels(dat$treatment)))

  dat %>%
    arrange(exp_unit, time) %>%
    group_by(exp_unit) %>%
    mutate(ar_start = row_number() == 1) %>%   # TRUE only at each unit's first timepoint
    ungroup()
}


# --- estimate_rho() ----------------------------------------------------------
# Estimates the AR(1) autocorrelation parameter (rho) by fitting a simplified
# bam() model without AR(1) correction, then extracting the lag-1 value from
# the residual autocorrelation function (ACF).
#
# The lag-0 ACF value is always 1 (perfect self-correlation), so the lag-1
# value is at index [2] of the $acf vector.
#
# Uses the combined treatment factor (CC, LC, CH, LH) rather than separate
# pH and temp terms. This is necessary because heat subsets contain only one
# level of temp after subsetting, which makes pH * temp * site unidentifiable
# and causes model failure.
#
# k is calculated dynamically from the number of unique timepoints in the
# subset rather than hardcoded. This is necessary because not all four
# datasets have the same number of timepoints — k cannot exceed the number
# of unique time values in the data being modelled, and setting it too high
# will cause an error.
#
# rho interpretation:
#   < 0.2:   weak — AR(1) correction omitted (negligible effect on inference)
#   0.2-0.5: moderate — correction applied
#   > 0.5:   strong — correction important for valid standard errors and p-values
#
# Arguments:
#   dat: data frame with ar_start column (output of add_ar_start())
# Returns:
#   numeric scalar: estimated AR(1) rho

estimate_rho <- function(dat) {

  # k ceiling from the number of unique timepoints in this subset.
  # Printed to console so you can confirm the correct value is being
  # used for each dataset.
  k_val <- length(unique(dat$time))
  cat("  Unique timepoints:", k_val, "-> using k =", k_val, "\n")

  mod_temp <- bam(
    FvFm ~ treatment * site +
      s(time, k = k_val) +
      s(exp_unit, bs = "re"),   # simple random intercept to absorb fragment variance
    data   = dat,
    method = "fREML"
  )

  rho <- acf(resid(mod_temp), plot = FALSE)$acf[2]
  cat("  Estimated rho:", round(rho, 3), "\n")

  if      (rho < 0.2) cat("  -> Weak autocorrelation; AR(1) correction will be omitted\n")
  else if (rho < 0.5) cat("  -> Moderate autocorrelation; AR(1) correction applied\n")
  else                cat("  -> Strong autocorrelation; AR(1) correction important\n")

  return(rho)
}


# --- fit_bam() ---------------------------------------------------------------
# Fits the full GAMM with:
#
# Fixed effects:
#   treatment * site — all main effects and interaction. treatment is the
#   combined pH x temp factor (CC, LC, CH, LH). Tests whether treatment
#   effects on mean FvFm differ across sites at the reference timepoint
#   (parametric interpretation). For control subsets all four treatment
#   levels are present; for heat subsets only CH and LH are present.
#   Using the combined factor rather than pH * temp * site avoids model
#   failure in heat subsets where temp has only one level.
#
# Smooth terms:
#   s(time):                    global nonlinear time trajectory shared
#                               across all groups
#   s(time, by = treatment):    treatment-specific deviation from the global
#                               trajectory — captures whether the shape of
#                               the decline differs between treatment
#                               combinations over time
#   s(time, by = site):         site-specific trajectory deviation — key
#                               term for testing whether resilient sites
#                               decline differently over time
#
# Random smooth:
#   s(time, exp_unit, bs = "fs"): factor-smooth fitting a smooth trajectory
#     deviation per fragment. Simultaneously acts as a repeated measures
#     random effect AND allows individual fragments to follow different
#     curve shapes. Penalty shrinks all fragment smooths toward zero
#     (the population curve).
#
# k is calculated dynamically from the number of unique timepoints in the
# subset. Each dataset gets its own correct k ceiling automatically. With
# fewer timepoints, the smooth has less complexity available, but the
# penalty will still shrink effective df further below the ceiling.
# If k.check() flags a problem after fitting, the solution is NOT to
# increase k (it is already at the maximum possible for the data) —
# instead revisit the AR(1) correction or random effects structure.
#
# Factor levels are preserved from the ipam pipeline — not recoded here.
#
# Arguments:
#   dat:     data frame with ar_start column (output of add_ar_start())
#   rho_est: numeric scalar from estimate_rho()
# Returns:
#   fitted bam model object

fit_bam <- function(dat, rho_est) {

  # k ceiling from the number of unique timepoints in this subset.
  # Printed to console alongside rho for verification.
  k_val <- length(unique(dat$time))
  cat("  Using k =", k_val, "\n")

  formula <- FvFm ~ treatment * site +
    s(time, k = k_val) +                          # global smooth decline
    s(time, by = treatment, k = k_val) +          # treatment-specific trajectory
                                                   # deviation (combined pH x temp —
                                                   # avoids single-level temp failure
                                                   # in heat subsets)
    s(time, by = site,      k = k_val) +          # site-specific trajectory deviation
    s(time, exp_unit, bs = "fs", k = k_val)       # fragment-level smooth deviations
                                                   # (repeated measures random effect)

  if (rho_est >= 0.2) {
    cat("  Fitting with AR(1) correction (rho =", round(rho_est, 3), ")\n")
    mod <- bam(formula,
               data     = dat,
               method   = "fREML",
               rho      = rho_est,
               AR.start = dat$ar_start)
  } else {
    cat("  Fitting without AR(1) correction (rho =", round(rho_est, 3), ")\n")
    mod <- bam(formula,
               data   = dat,
               method = "fREML")
  }

  return(mod)
}


# --- extract_stats() ---------------------------------------------------------
# Extracts parametric coefficients and smooth term significance tables
# into tidy data frames.
#
# Parametric table: coefficients relative to reference levels (CC treatment,
#   Emerald site). Reflect mean differences at the reference timepoint only —
#   not overall treatment effects across time. The treatment smooth terms
#   below are where the temporal treatment effects live.
#
# Smooth table: edf (effective df) and p-value per smooth term.
#   edf = 1: approximately linear. Higher edf = more curvature.
#   Significant p-values for s(time):treatment terms indicate that those
#   treatment combinations follow a trajectory shape that differs from the
#   global trend. Significant s(time):site terms indicate site-specific
#   trajectory differences — the key signal for your resilience hypothesis.
#
# Arguments:
#   mod:      fitted bam model object
#   mod_name: character label added as "model" column
# Returns:
#   list with $parametric and $smooths data frames

extract_stats <- function(mod, mod_name) {
  s <- summary(mod)

  parametric <- as.data.frame(s$p.table) %>%
    rownames_to_column("term") %>%
    mutate(model = mod_name) %>%
    rename(estimate  = Estimate,
           std_error = `Std. Error`,
           t_value   = `t value`,
           p_value   = `Pr(>|t|)`)

  smooths <- as.data.frame(s$s.table) %>%
    rownames_to_column("term") %>%
    mutate(model = mod_name) %>%
    rename(edf     = edf,
           ref_df  = Ref.df,
           F_value = F,
           p_value = `p-value`)

  # Overall model fit statistics
  # dev_expl: proportion of deviance explained by the full model —
  #           the GAMM equivalent of R-squared. Printed in summary()
  #           as "Deviance explained".
  # r_sq_adj: adjusted R-squared, also from summary(). May differ
  #           from dev_expl when the response distribution is non-Gaussian.
  fit <- tibble(
    model    = mod_name,
    dev_expl = summary(mod)$dev.expl,
    r_sq_adj = summary(mod)$r.sq
  )

  list(parametric = parametric, smooths = smooths, fit = fit)
}


# --- extract_kcheck() --------------------------------------------------------
# Runs k.check() and returns results as a tidy data frame.
#
# k-index near 1 = smooth basis dimension is adequate.
# k-index substantially below 1 with small p = potential underfitting.
# Because k is set to the number of unique timepoints (the maximum possible),
# this cannot be resolved by increasing k — if flagged, revisit the AR(1)
# correction or random effects structure instead.
#
# Arguments:
#   mod:      fitted bam model object
#   mod_name: character label
# Returns:
#   data frame of k.check() output

extract_kcheck <- function(mod, mod_name) {
  kc <- k.check(mod)
  as.data.frame(kc) %>%
    rownames_to_column("term") %>%
    mutate(model = mod_name)
}


# --- extract_emmeans() -------------------------------------------------------
# Computes estimated marginal means for all treatment x site combinations,
# evaluated across all observed timepoints, and pairwise contrasts among
# treatment combinations within each site (Tukey-adjusted).
#
# time is a smooth term in the GAMM — it cannot appear in the by argument
# of contrast(), which only accepts factor variables that exist as discrete
# grid dimensions. Timepoints are instead specified via the at argument
# when computing the emmeans grid, so marginal means and contrasts reflect
# overall treatment differences averaged across the full time course rather
# than at any specific moment. This is appropriate when the question is
# about general trajectory differences rather than differences at specific
# timepoints.
#
# Arguments:
#   mod:      fitted bam model object
#   dat:      data frame (used to extract observed timepoints)
#   mod_name: character label
# Returns:
#   list with $means and $contrasts data frames

extract_emmeans <- function(mod, dat, mod_name) {

  emm <- emmeans(mod,
                 specs = ~ treatment * site,
                 at    = list(time = unique(dat$time)),
                 pbkrtest.limit = 10000)

  # Marginal means averaged across all observed timepoints —
  # one row per treatment x site combination
  means <- as.data.frame(emm) %>%
    mutate(model = mod_name)

  # Pairwise contrasts among treatment combinations within each site,
  # averaged across timepoints. Reflects overall trajectory differences
  # rather than differences at specific moments. Tukey adjustment controls
  # for multiple comparisons within each site.
  contrasts_out <- as.data.frame(
    contrast(emm,
             method = "pairwise",
             by     = "site",
             adjust = "tukey")
  ) %>%
    mutate(model = mod_name)

  list(means = means, contrasts = contrasts_out)
}


# --- extract_joint_tests() ---------------------------------------------------
# Computes omnibus F-tests for each fixed factor (treatment, site, and their
# interaction) using emmeans::joint_tests(). This produces a single F-statistic
# and p-value per factor, analogous to a Type III ANOVA, averaged across all
# observed timepoints.
#
# This addresses a key limitation of the parametric table, which only reports
# individual contrast coefficients relative to the reference level rather than
# an overall factor effect. joint_tests() is preferred over car::Anova() here
# because it operates on the emmeans grid (averaged over time) rather than
# requiring model refitting with ML, and it handles the GAMM structure cleanly.
#
# Interpretation:
#   treatment:      Is there any overall difference in mean FvFm between
#                   treatment combinations, averaged across sites and timepoints?
#   site:           Is there any overall difference in mean FvFm between sites,
#                   averaged across treatments and timepoints?
#   treatment:site: Does the treatment effect differ across sites?
#
# Arguments:
#   mod:      fitted bam model object
#   dat:      data frame (used to extract observed timepoints)
#   mod_name: character label
# Returns:
#   data frame of joint_tests() output with model column added

extract_joint_tests <- function(mod, dat, mod_name) {

  emm <- emmeans(mod,
                 specs = ~ treatment * site,
                 at    = list(time = unique(dat$time)),
                 pbkrtest.limit = 10000)

  jt <- as.data.frame(joint_tests(emm)) %>%
    mutate(model = mod_name) %>%
    rename(df1     = df1,
           df2     = df2,
           F_value = `F.ratio`,
           p_value = `p.value`) %>%
    mutate(
      # Partial eta-squared: proportion of variance explained by each factor
      # after accounting for all other terms in the model.
      # Calculated from F and degrees of freedom as:
      #   eta_p^2 = (F * df1) / (F * df1 + df2)
      # Interpretation: 0.01 = small, 0.06 = medium, 0.14 = large effect
      partial_R2 = (F_value * df1) / (F_value * df1 + df2)
    )

  return(jt)
}


# --- get_smooth_stat() --------------------------------------------------------
# Looks up edf, Ref.df, F, and p-value for one smooth term from a
# results$stats$smooths table (output of extract_stats()).
#
# Arguments:
#   smooths_tbl: data frame from results$stats$smooths
#   term:        exact term label as it appears in the smooth table,
#                e.g. "s(time)" for the global temporal smooth
# Returns:
#   named list: df1 (edf), df2 (Ref.df), F_value, p_value

get_smooth_stat <- function(smooths_tbl, term) {
  row <- smooths_tbl %>% filter(term == !!term)
  if (nrow(row) != 1) stop("Expected exactly one smooth row for '", term, "', found ", nrow(row))
  list(df1 = row$edf, df2 = row$ref_df, F_value = row$F_value, p_value = row$p_value)
}


# --- get_joint_stat() ----------------------------------------------------------
# Looks up df1, df2, F-value, and p-value for one fixed-effect term from a
# results$joint_tests table (output of extract_joint_tests()).
#
# emmeans::joint_tests() sometimes reports a higher-order term as
# "(confounded)" instead of the requested term name — this happens when its
# contrast subspace overlaps with a lower-order term already tested. In this
# GAMM design the treatment:site interaction is frequently confounded because
# the treatment- and site-specific time smooths already absorb part of the
# interaction contrast space. When that happens the requested term row does
# not exist, and NULL is returned so the caller can fall back to a
# "not testable" label instead of erroring.
#
# Arguments:
#   joint_tbl: data frame from results$joint_tests
#   term:      value to match in the `model term` column, e.g. "treatment",
#              "site", or "treatment:site"
# Returns:
#   named list (df1, df2, F_value, p_value), or NULL if the term is absent

get_joint_stat <- function(joint_tbl, term) {
  row <- joint_tbl %>% filter(`model term` == term)
  if (nrow(row) == 0) return(NULL)
  list(df1 = row$df1, df2 = row$df2, F_value = row$F_value, p_value = row$p_value)
}


# --- format_stat_label() --------------------------------------------------------
# Formats a stat block into the "term: F(df1,df2)=value, p=value*" label text
# used to annotate the summary line plots. Asterisk flags p < 0.05.
# stat = NULL (term not testable, e.g. "(confounded)" from get_joint_stat())
# produces a "not testable" label instead of erroring.
#
# Arguments:
#   prefix: label prefix, e.g. "time", "treat", "site", "interaction"
#   stat:   named list from get_smooth_stat()/get_joint_stat(), or NULL
# Returns:
#   character string

format_stat_label <- function(prefix, stat) {
  if (is.null(stat)) {
    return(paste0(prefix, ": not testable (confounded)"))
  }
  p_str <- if (stat$p_value < 0.001) {
    "p<0.001"
  } else {
    paste0("p=", formatC(stat$p_value, digits = 3, format = "f"))
  }
  star <- if (stat$p_value < 0.05) "*" else ""
  sprintf("%s: F(%.1f,%.1f)=%.1f, %s%s",
          prefix, stat$df1, stat$df2, stat$F_value, p_str, star)
}


# --- build_stat_labels() --------------------------------------------------------
# Builds the three stat annotation rows (time, treat, site) placed on one
# facet panel of a summary line plot. Values are pulled live from a
# run_full_analysis() result rather than hardcoded, so the labels always
# reflect the current GAMM fit.
#
# "time" comes from the global s(time) smooth (results$stats$smooths) —
# joint_tests() only covers the parametric factors, not the temporal trend.
# "treat" and "site" come from the omnibus joint_tests table. The
# treatment:site interaction is omitted — emmeans::joint_tests() reports it
# as "(confounded)" for this GAMM design (see get_joint_stat()), so it isn't
# a usable annotation.
#
# label_ys is supplied by the caller (see the shared FvFm_label_ys computed
# in plot-ylim) rather than hardcoded, because raw FvFm tops out well below
# 1 — fixed positions like 1.000/0.925/... were tuned for the old 0-1 ratio
# scale and would sit off the top of an FvFm-scaled plot.
#
# Arguments:
#   result:   named list from run_full_analysis() (e.g. results_ofav_control)
#   pH_val:   pH facet value to place the labels in, e.g. "low pH"
#   temp_val: temp facet value to place the labels in, e.g. "control temp"
#   label_ys: numeric vector of length 3 — y-position for time/treat/site
#             labels, in that order
#   time_val: x-position for all three labels. Defaults to Inf so labels
#             anchor to the panel's right edge regardless of the time axis
#             range — paired with hjust in geom_text() for right-justified
#             text that reads inward from the edge.
# Returns:
#   data frame with one row per label (time, FvFm, lab, pH, temp)

build_stat_labels <- function(result, pH_val, temp_val, label_ys, time_val = Inf) {
  data.frame(
    time = time_val,
    FvFm = label_ys,
    lab  = c(
      format_stat_label("time",  get_smooth_stat(result$stats$smooths, "s(time)")),
      format_stat_label("treat", get_joint_stat(result$joint_tests, "treatment")),
      format_stat_label("site",  get_joint_stat(result$joint_tests, "site"))
    ),
    pH   = pH_val,
    temp = temp_val
  )
}


# --- make_trajectory_plot() --------------------------------------------------
# Plots model-fitted FvFm trajectories over time with 95% CI ribbons,
# one panel per site, coloured by pH x temp treatment combination.
# Uses gratia::fitted_values() to extract predictions with uncertainty.
# Colour and group mappings use interaction(pH, temp) rather than the
# combined treatment factor so that axis labels remain human-readable.
#
# Arguments:
#   mod:      fitted bam model object
#   dat:      data frame
#   mod_name: character label used in plot title
# Returns:
#   ggplot object

make_trajectory_plot <- function(mod, dat, mod_name) {
  fitted_vals <- fitted_values(mod, data = dat)

  ggplot(fitted_vals,
         aes(x      = time,
             y      = .fitted,
             colour = interaction(pH, temp, sep = " / "),
             fill   = interaction(pH, temp, sep = " / "),
             group  = interaction(pH, temp))) +
    geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci),
                alpha = 0.15, colour = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    facet_wrap(~ site, labeller = label_both) +
    scale_colour_brewer(palette = "Dark2") +
    scale_fill_brewer(palette   = "Dark2") +
    labs(x      = "Time (days)",
         y      = "Fv/Fm",
         colour = "pH / Temperature",
         fill   = "pH / Temperature",
         title  = paste("Fitted Fv/Fm trajectories —", mod_name)) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")
}


# --- make_smooth_plot() ------------------------------------------------------
# Plots smooth terms using gratia::draw(). The factor-smooth for exp_unit
# is excluded (too many levels to plot usefully). edf on each panel
# indicates nonlinearity: edf = 1 is linear, higher values = more curvature.
#
# Arguments:
#   mod:      fitted bam model object
#   mod_name: character label
# Returns:
#   patchwork ggplot object

make_smooth_plot <- function(mod, mod_name) {
  draw(mod,
       residuals = FALSE,
       select    = seq_len(length(mod$smooth) - 1)) +  # exclude fs smooth
    plot_annotation(title = paste("Smooth terms —", mod_name))
}


# --- make_diff_smooth_plot() -------------------------------------------------
# Plots the treatment-specific smooth terms to show how each treatment
# trajectory deviates from the global time trend.
#
# difference_smooths() is not used here because it produces non-conformable
# array errors when the treatment factor has different numbers of levels
# across datasets (4 in control subsets, 2 in heat subsets), causing
# mismatched prediction matrix dimensions.
#
# Instead, the treatment-specific smooth terms are plotted directly using
# gratia::draw(). Smooths that deviate substantially from zero indicate
# treatment trajectories that differ meaningfully from the global trend.
# The emmeans pairwise contrasts provide the formal statistical comparison.
#
# Arguments:
#   mod:      fitted bam model object
#   mod_name: character label
# Returns:
#   patchwork ggplot object

make_diff_smooth_plot <- function(mod, mod_name) {

  # Identify which smooths correspond to the treatment by= terms.
  # smooth labels follow the pattern "s(time):treatmentXX" —
  # select these by index to pass to draw()
  smooth_labels <- sapply(mod$smooth, function(s) s$label)
  treatment_idx <- grep("treatment", smooth_labels)

  draw(mod,
       residuals = FALSE,
       select    = treatment_idx) +
    plot_annotation(
      title    = paste("Treatment-specific smooth deviations —", mod_name),
      subtitle = "Deviation from global time trend; values near zero follow the global trajectory"
    )
}


# --- save_plot() -------------------------------------------------------------
# Saves a ggplot or patchwork object as a PDF.
# Use device = pdf (default PDF device, no dependencies) or
# device = cairo_pdf (better Unicode and transparency handling, requires
# XQuartz on macOS — install from xquartz.org if needed).
#
# Arguments:
#   plot:     ggplot or patchwork object
#   filename: name without extension — saved to outputs/photosynthesis/
#   w, h:     width and height in inches

save_plot <- function(plot, filename, w = 10, h = 6) {
  path <- paste0("../../outputs/photosynthesis/", filename, ".pdf")
  ggsave(filename = path,
         plot     = plot,
         width    = w,
         height   = h,
         device   = grDevices::pdf)
  cat("  Saved:", path, "\n")
}


# --- run_full_analysis() -----------------------------------------------------
# Master wrapper — runs the complete pipeline for one dataset in sequence:
#   1. Add AR.start indicator and print treatment levels (after subsetting)
#   2. Estimate rho from simplified model residuals
#   3. Fit full GAMM with or without AR(1) depending on rho threshold
#   4. Print model summary and k.check to console
#   5. Extract tidy stats tables (parametric, smooths, kcheck, emmeans,
#      joint tests)
#   6. Build trajectory, smooth term, and difference smooth plots
#   7. Save all three plots as PDFs
#
# Arguments:
#   dat:      subsetted data frame (e.g. ipam_ofav_control)
#   mod_name: character label (e.g. "ofav_control")
# Returns:
#   named list: model, rho, stats, kcheck, emmeans, joint_tests, plots

run_full_analysis <- function(dat, mod_name) {
  cat("\n========================================\n")
  cat("Analysis:", mod_name, "\n")
  cat("========================================\n")

  cat("Step 1: Adding AR.start indicator...\n")
  dat <- add_ar_start(dat)

  cat("Step 2: Estimating rho...\n")
  rho_est <- estimate_rho(dat)

  cat("Step 3: Fitting GAMM...\n")
  mod <- fit_bam(dat, rho_est)

  cat("Step 4: Model summary\n")
  print(summary(mod))

  cat("Step 5: k.check\n")
  print(k.check(mod))

  cat("Step 6: Extracting stats tables...\n")
  stats   <- extract_stats(mod, mod_name)
  kcheck  <- extract_kcheck(mod, mod_name)
  emm_out <- extract_emmeans(mod, dat, mod_name)
  jt_out  <- extract_joint_tests(mod, dat, mod_name)

  cat("Step 7: Building and saving plots...\n")
  p_traj   <- make_trajectory_plot(mod, dat, mod_name)
  p_smooth <- make_smooth_plot(mod, mod_name)
  p_diff   <- make_diff_smooth_plot(mod, mod_name)

  save_plot(p_traj,   paste0(mod_name, "_trajectories"))
  save_plot(p_smooth, paste0(mod_name, "_smooths"), w = 14, h = 10)
  save_plot(p_diff,   paste0(mod_name, "_treatment_smooths"), w = 14, h = 8)

  cat("Done:", mod_name, "\n")

  list(
    model       = mod,
    rho         = rho_est,
    stats       = stats,
    kcheck      = kcheck,
    emmeans     = emm_out,
    joint_tests = jt_out,
    plots       = list(trajectory = p_traj,
                       smooths    = p_smooth,
                       diff       = p_diff)
  )
}
```

---

# *Orbicella faveolata* — Control {.tabset}

Control temperature treatment, timepoints < 21 days. Tests whether pH affects
photosynthetic efficiency prior to heat stress, and whether this differs across sites.

```{r ofav-control, cache=TRUE}
results_ofav_control <- run_full_analysis(
  dat      = ipam_ofav_control,
  mod_name = "ofav_control"
)
```

## Trajectories

```{r ofav-control-traj}
results_ofav_control$plots$trajectory
```

## Smooth terms

```{r ofav-control-smooth, fig.width=14, fig.height=10}
results_ofav_control$plots$smooths
```

## pH difference

```{r ofav-control-diff, fig.width=10, fig.height=5}
results_ofav_control$plots$diff
```

## Joint tests

```{r ofav-control-joint}
knitr::kable(
  results_ofav_control$joint_tests,
  digits  = 4,
  caption = "Omnibus F-tests — O. faveolata control"
)
```

## Parametric coefficients

```{r ofav-control-params}
knitr::kable(
  results_ofav_control$stats$parametric,
  digits  = 4,
  caption = "Fixed effect estimates — O. faveolata control"
)
```

## Smooth significance

```{r ofav-control-smoothsig}
knitr::kable(
  results_ofav_control$stats$smooths,
  digits  = 4,
  caption = "Smooth term significance — O. faveolata control"
)
```

## k.check

```{r ofav-control-kcheck}
knitr::kable(
  results_ofav_control$kcheck,
  digits  = 4,
  caption = "Smooth adequacy — O. faveolata control"
)
```

## Marginal means

```{r ofav-control-emm}
knitr::kable(
  results_ofav_control$emmeans$means,
  digits  = 4,
  caption = "Estimated marginal means — O. faveolata control"
)
```

## Pairwise contrasts

```{r ofav-control-contrasts}
knitr::kable(
  results_ofav_control$emmeans$contrasts,
  digits  = 4,
  caption = "Pairwise contrasts — O. faveolata control"
)
```

---

# *Orbicella faveolata* — Heat {.tabset}

High temperature treatment only, all timepoints. Tests whether pH effects on
photosynthetic efficiency are amplified or modulated under thermal stress.

```{r ofav-heat, cache=TRUE}
results_ofav_heat <- run_full_analysis(
  dat      = ipam_ofav_heat,
  mod_name = "ofav_heat"
)
```

## Trajectories

```{r ofav-heat-traj}
results_ofav_heat$plots$trajectory
```

## Smooth terms

```{r ofav-heat-smooth, fig.width=14, fig.height=10}
results_ofav_heat$plots$smooths
```

## pH difference

```{r ofav-heat-diff, fig.width=10, fig.height=5}
results_ofav_heat$plots$diff
```

## Joint tests

```{r ofav-heat-joint}
knitr::kable(
  results_ofav_heat$joint_tests,
  digits  = 4,
  caption = "Omnibus F-tests — O. faveolata heat"
)
```

## Parametric coefficients

```{r ofav-heat-params}
knitr::kable(
  results_ofav_heat$stats$parametric,
  digits  = 4,
  caption = "Fixed effect estimates — O. faveolata heat"
)
```

## Smooth significance

```{r ofav-heat-smoothsig}
knitr::kable(
  results_ofav_heat$stats$smooths,
  digits  = 4,
  caption = "Smooth term significance — O. faveolata heat"
)
```

## k.check

```{r ofav-heat-kcheck}
knitr::kable(
  results_ofav_heat$kcheck,
  digits  = 4,
  caption = "Smooth adequacy — O. faveolata heat"
)
```

## Marginal means

```{r ofav-heat-emm}
knitr::kable(
  results_ofav_heat$emmeans$means,
  digits  = 4,
  caption = "Estimated marginal means — O. faveolata heat"
)
```

## Pairwise contrasts

```{r ofav-heat-contrasts}
knitr::kable(
  results_ofav_heat$emmeans$contrasts,
  digits  = 4,
  caption = "Pairwise contrasts — O. faveolata heat"
)
```

---

# *Siderastrea siderea* — Control {.tabset}

Control temperature treatment, timepoints < 21 days.

```{r ssid-control, cache=TRUE}
results_ssid_control <- run_full_analysis(
  dat      = ipam_ssid_control,
  mod_name = "ssid_control"
)
```

## Trajectories

```{r ssid-control-traj}
results_ssid_control$plots$trajectory
```

## Smooth terms

```{r ssid-control-smooth, fig.width=14, fig.height=10}
results_ssid_control$plots$smooths
```

## pH difference

```{r ssid-control-diff, fig.width=10, fig.height=5}
results_ssid_control$plots$diff
```

## Joint tests

```{r ssid-control-joint}
knitr::kable(
  results_ssid_control$joint_tests,
  digits  = 4,
  caption = "Omnibus F-tests — S. siderea control"
)
```

## Parametric coefficients

```{r ssid-control-params}
knitr::kable(
  results_ssid_control$stats$parametric,
  digits  = 4,
  caption = "Fixed effect estimates — S. siderea control"
)
```

## Smooth significance

```{r ssid-control-smoothsig}
knitr::kable(
  results_ssid_control$stats$smooths,
  digits  = 4,
  caption = "Smooth term significance — S. siderea control"
)
```

## k.check

```{r ssid-control-kcheck}
knitr::kable(
  results_ssid_control$kcheck,
  digits  = 4,
  caption = "Smooth adequacy — S. siderea control"
)
```

## Marginal means

```{r ssid-control-emm}
knitr::kable(
  results_ssid_control$emmeans$means,
  digits  = 4,
  caption = "Estimated marginal means — S. siderea control"
)
```

## Pairwise contrasts

```{r ssid-control-contrasts}
knitr::kable(
  results_ssid_control$emmeans$contrasts,
  digits  = 4,
  caption = "Pairwise contrasts — S. siderea control"
)
```

---

# *Siderastrea siderea* — Heat {.tabset}

High temperature treatment only, all timepoints.

```{r ssid-heat, cache=TRUE}
results_ssid_heat <- run_full_analysis(
  dat      = ipam_ssid_heat,
  mod_name = "ssid_heat"
)
```

## Trajectories

```{r ssid-heat-traj}
results_ssid_heat$plots$trajectory
```

## Smooth terms

```{r ssid-heat-smooth, fig.width=14, fig.height=10}
results_ssid_heat$plots$smooths
```

## pH difference

```{r ssid-heat-diff, fig.width=10, fig.height=5}
results_ssid_heat$plots$diff
```

## Joint tests

```{r ssid-heat-joint}
knitr::kable(
  results_ssid_heat$joint_tests,
  digits  = 4,
  caption = "Omnibus F-tests — S. siderea heat"
)
```

## Parametric coefficients

```{r ssid-heat-params}
knitr::kable(
  results_ssid_heat$stats$parametric,
  digits  = 4,
  caption = "Fixed effect estimates — S. siderea heat"
)
```

## Smooth significance

```{r ssid-heat-smoothsig}
knitr::kable(
  results_ssid_heat$stats$smooths,
  digits  = 4,
  caption = "Smooth term significance — S. siderea heat"
)
```

## k.check

```{r ssid-heat-kcheck}
knitr::kable(
  results_ssid_heat$kcheck,
  digits  = 4,
  caption = "Smooth adequacy — S. siderea heat"
)
```

## Marginal means

```{r ssid-heat-emm}
knitr::kable(
  results_ssid_heat$emmeans$means,
  digits  = 4,
  caption = "Estimated marginal means — S. siderea heat"
)
```

## Pairwise contrasts

```{r ssid-heat-contrasts}
knitr::kable(
  results_ssid_heat$emmeans$contrasts,
  digits  = 4,
  caption = "Pairwise contrasts — S. siderea heat"
)
```

---

# Export stats

```{r export-stats}
# -----------------------------------------------------------------------------
# Combine stats tables across all four datasets and export to a single Excel
# workbook with one sheet per table type. The "model" column identifies the
# source dataset within each sheet.
# -----------------------------------------------------------------------------

all_results <- list(
  results_ofav_control,
  results_ofav_heat,
  results_ssid_control,
  results_ssid_heat
)

# Stack each table type across all four datasets
all_fit         <- bind_rows(lapply(all_results, function(r) r$stats$fit))
all_parametric  <- bind_rows(lapply(all_results, function(r) r$stats$parametric))
all_smooths     <- bind_rows(lapply(all_results, function(r) r$stats$smooths))
all_kcheck      <- bind_rows(lapply(all_results, function(r) r$kcheck))
all_means       <- bind_rows(lapply(all_results, function(r) r$emmeans$means))
all_contrasts   <- bind_rows(lapply(all_results, function(r) r$emmeans$contrasts))
all_joint_tests <- bind_rows(lapply(all_results, function(r) r$joint_tests))

# Rho summary — report in methods to justify AR(1) decisions per dataset
rho_summary <- tibble(
  model       = c("ofav_control", "ofav_heat", "ssid_control", "ssid_heat"),
  rho         = c(results_ofav_control$rho,
                  results_ofav_heat$rho,
                  results_ssid_control$rho,
                  results_ssid_heat$rho),
  AR1_applied = rho >= 0.2
)

# Export to a single Excel workbook
write_xlsx(
  list(
    "Model fit"                = all_fit,
    "Rho estimates"            = rho_summary,
    "Joint tests"              = all_joint_tests,
    "Parametric coefficients"  = all_parametric,
    "Smooth significance"      = all_smooths,
    "k.check"                  = all_kcheck,
    "Estimated marginal means" = all_means,
    "Pairwise contrasts"       = all_contrasts
  ),
  path = "../../outputs/photosynthesis/coral_FvFm_GAMM_stats.xlsx"
)

cat("Stats exported to: ../../outputs/photosynthesis/coral_FvFm_GAMM_stats.xlsx\n")

knitr::kable(rho_summary,
             digits  = 3,
             caption = "AR(1) autocorrelation estimates — all models")
```

---

# Summary line plots

Mean FvFm ± SE over time for each site, faceted by pH and temperature
treatment. Stats labels are pulled live from the GAMM output (`build_stat_labels()`)
computed in the four analysis sections above, so they always reflect the
current model fit rather than hardcoded values.

```{r plot-colors}
# -----------------------------------------------------------------------------
# Site color palette — applied in factor level order: Emerald, Rainbow,
# Star, MacN
# -----------------------------------------------------------------------------
fill.color <- c("#018571", "#80cdc1", "#dfc27d", "#a6611a")
```

```{r plot-ylim}
# -----------------------------------------------------------------------------
# Shared y-axis range for the Ofav and Ssid summary line plots.
#
# Computed from the mean ± SE actually drawn by stat_summary() (grouped by
# site, pH, temp, time — matching the plot's aes/facet grouping), not the raw
# FvFm scatter, so the range matches what's visually on the plot. The same
# range is applied to both species plots so panels are directly comparable.
#
# A small pad is added below the min and above the max — without it, a point
# or errorbar cap sitting exactly at the min/max gets its marker clipped by
# the panel edge (coord_cartesian() clips hard at the exact ylim value, and a
# marker's drawn radius extends past its data coordinate). The bottom pad is
# larger than the top pad since there's no headroom band competing for space
# below the data.
#
# The three stacked stat labels (time, treat, site) are NOT given a
# dedicated headroom band above the data — that was inflating ymax every
# time label spacing increased. Instead they anchor just under the plot's
# top edge and stack downward into the existing range, right-justified
# (see geom_text(hjust = ...) in the plot chunks) so they sit in the
# upper-right corner, away from the peak values which occur early in the
# time series.
# -----------------------------------------------------------------------------

summarise_mean_se_range <- function(dat) {
  dat %>%
    group_by(site, pH, temp, time) %>%
    summarise(mean_FvFm = mean(FvFm, na.rm = TRUE),
              se_FvFm   = sd(FvFm,   na.rm = TRUE) / sqrt(n()),
              .groups   = "drop") %>%
    summarise(min_val = min(mean_FvFm - se_FvFm, na.rm = TRUE),
              max_val = max(mean_FvFm + se_FvFm, na.rm = TRUE))
}

FvFm_range <- bind_rows(summarise_mean_se_range(ipam_ofav),
                         summarise_mean_se_range(ipam_ssid)) %>%
  summarise(min_val = min(min_val), max_val = max(max_val))

FvFm_span        <- FvFm_range$max_val - FvFm_range$min_val
FvFm_pad_bottom  <- 0.08 * FvFm_span
FvFm_pad_top     <- 0.03 * FvFm_span
FvFm_data_max    <- FvFm_range$max_val + FvFm_pad_top
FvFm_ylim        <- c(FvFm_range$min_val - FvFm_pad_bottom, FvFm_data_max)

FvFm_label_margin <- 0.03 * FvFm_span   # gap below ymax so the top label's text isn't clipped
FvFm_label_step   <- 0.09 * FvFm_span   # vertical gap between stacked labels (~50% more than before)
FvFm_label_top    <- FvFm_data_max - FvFm_label_margin
FvFm_label_ys     <- FvFm_label_top - FvFm_label_step * c(0, 1, 2)
```

```{r ofav-line-plot, fig.width=12, fig.height=6}
# -----------------------------------------------------------------------------
# Stats label data frames for O. faveolata
#
# build_stat_labels() (see Helper functions) pulls the time/treat/site/
# interaction F-stats and p-values directly from the GAMM output of the
# corresponding analysis above, one call per facet panel. Label y-positions
# and the plot's y-axis range come from plot-ylim, shared with the Ssid plot.
#
# x = Inf (set by build_stat_labels()'s time_val default) anchors labels to
# the panel's right edge; hjust = 1.1 right-justifies the text and nudges it
# slightly inward so it doesn't touch the plot border.
# -----------------------------------------------------------------------------

ofav_control_labs <- build_stat_labels(results_ofav_control, "low pH", "control temp", FvFm_label_ys)
ofav_heat_labs    <- build_stat_labels(results_ofav_heat,    "low pH", "high temp",    FvFm_label_ys)

ofav <- ggplot(ipam_ofav,
               aes(x = time, y = FvFm, color = site, group = site)) +
  stat_summary(fun      = mean,
               geom     = "line",
               size     = 0.75) +
  stat_summary(fun  = mean,
               geom = "point",
               size = 2) +
  stat_summary(fun.data = mean_se,
               geom     = "errorbar",
               width    = 0.2) +
  facet_grid(temp ~ pH) +
  scale_color_manual(values = fill.color) +
  coord_cartesian(ylim = FvFm_ylim) +
  labs(title = "Ofav",
       x     = "Day",
       y     = "Fv/Fm",
       color = "Site") +
  theme_pubr(legend = "bottom") +
  geom_text(data  = ofav_control_labs,
            aes(x = time, y = FvFm, label = lab),
            hjust = 1.1,
            inherit.aes = FALSE) +
  geom_text(data  = ofav_heat_labs,
            aes(x = time, y = FvFm, label = lab),
            hjust = 1.1,
            inherit.aes = FALSE)

ofav

ggsave("../../outputs/photosynthesis/urban ipam ofav.pdf",
       plot   = ofav,
       width  = 12,
       height = 6,
       device = pdf)
```

```{r ssid-line-plot, fig.width=12, fig.height=6}
# -----------------------------------------------------------------------------
# Stats label data frames for S. siderea
#
# build_stat_labels() (see Helper functions) pulls the time/treat/site/
# interaction F-stats and p-values directly from the GAMM output of the
# corresponding analysis above, one call per facet panel. Label y-positions
# and the plot's y-axis range come from plot-ylim, shared with the Ofav plot.
#
# x = Inf (set by build_stat_labels()'s time_val default) anchors labels to
# the panel's right edge; hjust = 1.1 right-justifies the text and nudges it
# slightly inward so it doesn't touch the plot border.
# -----------------------------------------------------------------------------

ssid_control_labs <- build_stat_labels(results_ssid_control, "low pH", "control temp", FvFm_label_ys)
ssid_heat_labs    <- build_stat_labels(results_ssid_heat,    "low pH", "high temp",    FvFm_label_ys)

ssid <- ggplot(ipam_ssid,
               aes(x = time, y = FvFm, color = site, group = site)) +
  stat_summary(fun      = mean,
               geom     = "line",
               size     = 0.75) +
  stat_summary(fun  = mean,
               geom = "point",
               size = 2) +
  stat_summary(fun.data = mean_se,
               geom     = "errorbar",
               width    = 0.2) +
  facet_grid(temp ~ pH) +
  scale_color_manual(values = fill.color) +
  coord_cartesian(ylim = FvFm_ylim) +
  labs(title = "Ssid",
       x     = "Day",
       y     = "Fv/Fm",
       color = "Site") +
  theme_pubr(legend = "bottom") +
  geom_text(data  = ssid_control_labs,
            aes(x = time, y = FvFm, label = lab),
            hjust = 1.1,
            inherit.aes = FALSE) +
  geom_text(data  = ssid_heat_labs,
            aes(x = time, y = FvFm, label = lab),
            hjust = 1.1,
            inherit.aes = FALSE)

ssid

ggsave("../../outputs/photosynthesis/urban ipam ssid.pdf",
       plot   = ssid,
       width  = 12,
       height = 6,
       device = pdf)
```

---

# Session info

```{r session-info}
# -----------------------------------------------------------------------------
# Record R and package versions for reproducibility.
# Include this output in supplementary materials when submitting.
# -----------------------------------------------------------------------------
sessionInfo()
```
