---
title: "O. faveolata Gene Expression — Differential Expression (dream)"
author: "Michael Studivan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# -----------------------------------------------------------------------------
# Global chunk options
# echo = TRUE:     show code in rendered output
# warning = FALSE: suppress warnings in rendered document
#                  (warnings still print to console during knitting)
# message = FALSE: suppress package loading messages
# -----------------------------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 300
)

# save_plot(): writes a figure to PDF (outputs folder) and shows the same figure in the knitted HTML.
# The plotting code is run once for the PDF and once for a PNG copy at the same size, which is embedded in
# the HTML; when run interactively (not knitting) the plot is also drawn in the plot pane.
save_plot <- function(file, width, height, expr) {
  expr <- substitute(expr); env <- parent.frame()
  # draw on a newly opened device and close exactly that device (never whichever device happens to be current).
  # Some plotting functions move drawing away from that device: pheatmap() opens and closes a temporary device
  # (pdf(NULL) ... dev.off()) before drawing, and R then makes some other open device current, which left PDFs blank or
  # missing from the HTML. RStudio/knitr graphics hooks can do the same. While the file is drawn, those hooks are
  # replaced by one that switches back to the file's device each time a new plot page starts (plot.new / grid.newpage).
  hooks <- c("before.plot.new", "plot.new", "before.grid.newpage", "grid.newpage")
  draw_on <- function(open_device) {
    saved <- lapply(setNames(hooks, hooks), getHook)
    open_device(); d <- dev.cur()
    back_to_d <- function() if (d %in% dev.list() && dev.cur() != d) dev.set(d)
    for (h in hooks) setHook(h, NULL, "replace")
    setHook("before.plot.new", back_to_d); setHook("before.grid.newpage", back_to_d)
    on.exit({ if (d %in% dev.list()) dev.off(d)
              for (h in hooks) setHook(h, saved[[h]], "replace") })
    eval(expr, env)
  }
  draw_on(function() pdf(file = file, width = width, height = height))
  if (!isTRUE(getOption("knitr.in.progress"))) { # interactive: also draw in the plot pane (skipped if the pane is too small)
    ok <- try(eval(expr, env), silent = TRUE)
    if (inherits(ok, "try-error") && dev.cur() > 1)
      try(dev.off(dev.cur()), silent = TRUE) # reset the screen device left in a broken state (the PDF is already closed)
    return(invisible(file)) }
  png_file <- file.path(tempdir(), sub("\\.pdf$", ".png", basename(file)))
  draw_on(function() png(png_file, width = width, height = height, units = "in", res = 150))
  if (!file.exists(png_file)) { warning("nothing was drawn for ", basename(file)); return(invisible(file)) }
  knitr::include_graphics(png_file, rel_path = FALSE, dpi = 150) # displayed at its true size
}

# -----------------------------------------------------------------------------
# Sensitivity sections (chunks named sensitivity-...): FALSE = skipped when knitting (routine knits);
# TRUE = run in the knit, for the final knit of record. They refit the models several times and are slow
# (O. faveolata: LH vs CH robustness alone refits the Kenward-Roger model seven times).
# -----------------------------------------------------------------------------
run_sensitivity <- TRUE
```


# PACKAGES

```{r packages}
# run these once, then comment out
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.16")
# BiocManager::install("DESeq2",dependencies=T)
# BiocManager::install("arrayQualityMetrics",dependencies=T)  # requires Xquartz, xquartz.org
# BiocManager::install("BiocParallel")
# BiocManager::install("edgeR")  # also installs limma
# BiocManager::install("variancePartition")  # dream: mixed model for each gene (differential expression)

# install.packages("pheatmap")
# install.packages("VennDiagram")
# install.packages("gplots")
# install.packages("vegan")
# install.packages("plotrix")
# install.packages("ape")
# install.packages("ggplot2")
# install.packages("rgl")
# install.packages("adegenet")
```


# DATA IMPORT

```{r data-import}
# assembling data, running outlier detection, and fitting models
# (skip this section if you don't need to remake models)

library(DESeq2)
library(arrayQualityMetrics)
library(dplyr)

#read in counts
counts = read.table("../../data/transcriptomics/ofav/allcounts_host.txt")

# how many genes we have total?
nrow(counts) 
ncol(counts)

# how does the data look? 
head(counts)

# importing a design .csv file
design = read.csv("../../data/transcriptomics/ofav/design_ofav.csv", head=TRUE)
design
# count columns and design rows must be in the same sample order before any subsetting
stopifnot(identical(colnames(counts), design$id))
design$site <- as.factor(design$site)
# replicate as a factor (as in the S. siderea script)
design$replicate <- as.factor(design$replicate)
design$tank <- as.factor(design$tank)
design$ph <- as.factor(design$ph)
design$temp <- as.factor(design$temp)
design$treatment <- as.factor(design$treatment)
design$treat <- as.factor(design$treat)
# parent colony (genotype): fragments cut from the same colony share a genotype ID (unique across sites)
design$genotype <- as.factor(design$genotype)
# habitat: urban (Star Island, MacArthur North) vs reef (Emerald, Rainbow)
design$habitat <- factor(ifelse(design$site %in% c("Star", "MacN"), "urban", "reef"), levels = c("reef", "urban"))

# fragments removed from the survivorship analysis are excluded from all analyses (reasons in the survivorship
# data file's `exclude` column): here 7 samples from 6 O. faveolata fragments with partial mortality unrelated to
# heat stress (mostly a ciliate infection): Eo1-3, Eo3-3, Ro2-3, Ro4-3, So1-3, Mo5-3 (LC), and Mo5-3's heat-stress
# resample (LH). Removed by sample ID before outlier detection,
# so the outlier tests and the transformation behind them use only fragments that are analysed.
# only O. faveolata exclusions (fragment IDs are matched within species)
surv_exclude <- read.csv("../../data/survivorship/urban survivorship.csv") %>% filter(species == "Ofav", exclude != "")
surv_excl <- design$sampleid %in% surv_exclude$sample
stopifnot(sum(surv_excl) == 7) # 7 sequenced samples excluded (see above)
design[surv_excl, c("id", "sampleid", "treat")] # samples excluded
design=design[!surv_excl,]

# PRIMARY ANALYSIS = FIRST COHORT ONLY (decided 2026-09-30).
# The control-temperature tanks were heated later (second cohort), and 10 of their O. faveolata fragments were
# resampled as LH on 14 May; every other sample was taken on 23 April. Within the same colonies these resamples
# differ from first-cohort LH samples (~2,300 DEGs; SENSITIVITY: COHORT EFFECT WITHIN LH), so pooling them would
# confound the LH treatment with sampling date, tank and repeat biopsy. They are removed from the primary analysis
# here, before the QC gene filter and outlier detection, so the outlier tests use only primary-analysis samples;
# the pooled model is refit in SENSITIVITY: POOLED MODEL. (In the earlier report that included them, none of the
# resamples was flagged as an outlier.)
# resampled = heat-treatment sample whose fragment ID also appears under a control-temperature treatment in the
# full design sheet (so pairs whose LC sample was excluded are still caught)
full_design <- read.csv("../../data/transcriptomics/ofav/design_ofav.csv", head=TRUE)
resampled_ids <- full_design$id[full_design$treat %in% c("CH","LH") &
                                full_design$sampleid %in% full_design$sampleid[full_design$treat %in% c("CC","LC")]]
resampled <- design$id %in% resampled_ids
design[resampled, c("id","sampleid","treat","tank")] # second-cohort resamples (LH, tank 4)
stopifnot(sum(resampled) == 10, all(design$treat[resampled] == "LH"))
design_pooled <- design          # kept for the pooled sensitivity model (outliers removed below)
design <- design[!resampled, ]
table(treat = design$treat)

# gene filter on the retained samples (exclusions and second-cohort resamples removed), used only for outlier detection (arrayQualityMetrics) below;
# countData and counts4wgcna are recomputed (and exported) on the analysed samples after outlier removal
counts_qc <- counts[, design$id]
stopifnot(identical(colnames(counts_qc), design$id)) # count columns and design rows in the same order
keep <- rowSums(counts_qc) >= 10
countData <- counts_qc[keep,]
nrow(countData)
ncol(countData)

str(design)
```


# MODEL DESIGN and OUTLIERS

```{r model-design-and-outliers}
# make big dataframe including all factors and interaction, getting normalized data for outlier detection
# parent colony (genotype) is included so treatment effects are estimated within colonies;
# (this design is only used for the variance-stabilising transformation; differential expression is tested with
# the site model in DIFFERENTIAL EXPRESSION)
dds = DESeqDataSetFromMatrix(countData=countData, colData=design, design=~ genotype+treat)

# reorders treatment factor according to "control" vs "treatment" levels
dds$site <- factor(dds$site, levels = c("Emerald", "Rainbow", "Star", "MacN"))
dds$treat <- factor(dds$treat, levels = c("CC", "LC", "CH", "LH"))

# for large datasets, rlog may take too much time, especially for an unfiltered dataframe
# vsd is much faster and still works for outlier detection
Vsd=varianceStabilizingTransformation(dds)

library(Biobase)
e=ExpressionSet(assay(Vsd), AnnotatedDataFrame(as.data.frame(colData(Vsd))))

# running outlier detection
devs_before <- dev.list() # arrayQualityMetrics leaves graphics devices open; close only the ones it opened
arrayQualityMetrics(e, outdir="../../outputs/transcriptomics/ofav/limma/arrayQualityMetrics", intgroup=c("site"), force=T)
for (dd in setdiff(dev.list(), devs_before)) dev.off(dd)
```

Open the directory "outputs/transcriptomics/ofav/limma/arrayQualityMetrics" and open index.html  
Array metadata and outlier detection overview gives a report of all samples, and which are likely outliers according to the 3 methods tested. I typically remove the samples that violate *1 (distance between arrays).  
Figure 2 shows a bar plot of array-to-array distances and an outlier detection threshold based on your samples. Samples above the threshold are considered outliers  
Under Figure 3: Principal Components Analyses, look for any points far away from the rest of the sample cluster  
Use the sample ID (`id`) of each outlier array for removal in the following section

```{r model-design-and-outliers-cont}
# if there were outliers:
# outliers are removed by sample ID (not array position), so removal does not depend on sample order
# IDs from the arrayQualityMetrics report built on the primary samples (survivorship exclusions and second-cohort
# resamples removed; distance-between-arrays, *1): Ofav_125 (So4-3, LC) and Ofav_131 (Eo2-4, LH).
# (Earlier reports that still included the excluded fragments and/or resamples also flagged Ofav_109, 127 and 135.)
# Ofav_135 (So5-4, LH) is also removed (Michael, 2026-09-30): it was flagged as an array outlier in both earlier
# reports, is the most distant sample after removing colony differences (~2x the typical distance), and alone
# changes the LH results on its own: with the primary model (SENSITIVITY: LH vs CH ROBUSTNESS, 2026-10-06), adding it
# back gives 84 LH vs CH DEGs instead of 0 and doubles LH vs CC (3,959 vs 1,914 DEGs); in the earlier DESeq2 colony
# model it was 2,089 vs 19 LH vs CH DEGs. Report this removal and its reason in the Methods.
outlier_ids <- c("Ofav_125","Ofav_131","Ofav_135")
outs <- which(design$id %in% outlier_ids)
stopifnot(length(outs) == length(outlier_ids))
design[outs, c("id","sampleid","treat")] # samples removed as outliers
design=design[-outs,]
design_pooled <- design_pooled[!(design_pooled$id %in% outlier_ids), ] # same outliers removed from the pooled sample set

# gene filters recomputed on the analysed samples only (exclusions, outliers and second-cohort resamples removed), from the raw counts
counts_kept <- counts[, design$id]
stopifnot(identical(colnames(counts_kept), design$id)) # count columns and design rows in the same order
keep <- rowSums(counts_kept) >= 10
countData <- counts_kept[keep,]
nrow(countData)
ncol(countData)
write.csv(countData, file="../../outputs/transcriptomics/ofav/limma/countData.csv")

# for WGCNA: removing all genes with counts of <10 in more than 90 % of samples (analysed samples only)
counts4wgcna = counts_kept[apply(counts_kept,1,function(x) sum(x<10))<ncol(counts_kept)*0.9,]
nrow(counts4wgcna)
ncol(counts4wgcna)
write.csv(counts4wgcna, file="../../outputs/transcriptomics/ofav/limma/counts4wgcna.csv")

# remaking model with outliers removed from dataset
dds = DESeqDataSetFromMatrix(countData=countData, colData=design, design=~ genotype+treat)
dds$site <- factor(dds$site, levels = c("Emerald", "Rainbow", "Star", "MacN"))
dds$treat <- factor(dds$treat, levels = c("CC", "LC", "CH", "LH"))

# variance-stabilized data recomputed on the analysed samples (the Vsd above, before outlier removal, was for outlier detection only)
Vsd=varianceStabilizingTransformation(dds)

# save all these dataframes as an Rdata package so you don't need to rerun each time
save(dds,design,design_pooled,countData,Vsd,counts4wgcna,file="../../outputs/transcriptomics/ofav/limma/initial.RData")

# generating normalized variance-stabilized data for PCoA, heatmaps, etc
vsd=assay(Vsd)
# takes the sample IDs and factor levels from the design to create new column names for the dataframe
# (site and treat by name rather than column position)
snames=paste(colnames(countData),design$site,design$treat,sep=".")
# renames the column names
colnames(vsd)=snames

save(vsd,design,file="../../outputs/transcriptomics/ofav/limma/vsd.RData")

# more reduced stabilized dataset for WGCNA
wg = DESeqDataSetFromMatrix(countData=counts4wgcna, colData=design, design=~ site+treat)
vsd.wg=assay(varianceStabilizingTransformation(wg, blind=TRUE))
# vsd.wg=assay(rlog(wg), blind=TRUE)
head(vsd.wg)
colnames(vsd.wg)=snames
save(vsd.wg,design,file="../../outputs/transcriptomics/ofav/limma/data4wgcna.RData")
```


# PCOA and PERMANOVA

```{r pcoa-and-permanova}
# heatmap and hierarchical clustering:
load("../../outputs/transcriptomics/ofav/limma/vsd.RData")
library(pheatmap)
# similarity among samples
save_plot(file="../../outputs/transcriptomics/ofav/limma/heatmap_ofav_host.pdf", width=15, height=15, {
pheatmap(cor(vsd))
})

# Principal coordinates analysis
library(vegan)
library(ape)

conditions=design
conditions$site <- factor(conditions$site, levels = c("Emerald", "Rainbow", "Star", "MacN"))
conditions$treat <- factor(conditions$treat, levels = c("CC", "LC", "CH", "LH"))

# creating a PCoA eigenvalue matrix
dds.pcoa=pcoa(dist(t(vsd),method="manhattan")/1000)
scores=dds.pcoa$vectors
# % variation explained by each axis, computed from the PCoA (used for axis labels)
ve <- sprintf("%.1f", 100 * dds.pcoa$values$Relative_eig[1:4])
# export this table for % variation explained by each axis (Relative_eig column)
write.csv(dds.pcoa$values, file = "../../outputs/transcriptomics/ofav/limma/PCoA_variance.csv")

# how many good PC's do we have? Compared to random ("broken stick") model
# plotting PCoA eigenvalues 
save_plot(file="../../outputs/transcriptomics/ofav/limma/PCoA_Manhattan.pdf", width=6, height=6, {
plot(dds.pcoa$values$Relative_eig)
points(dds.pcoa$values$Broken_stick,col="red",pch=3)
}) # the number of black points above the line of red crosses (random model) corresponds to the number of good PC's

# plotting PCoA by site and treatment
save_plot(file="../../outputs/transcriptomics/ofav/limma/PCoA_ofav_host.pdf", width=12, height=6, {
par(mfrow=c(1,2))
plot(scores[,1], scores[,2],col=c("#018571","#80cdc1","#dfc27d","#a6611a")[as.numeric(as.factor(conditions$site))],pch=c(6,25,2,17)[as.numeric(as.factor(conditions$treat))], bg= c("#018571","#80cdc1","#dfc27d","#a6611a")[as.numeric(as.factor(conditions$site))], xlab=paste0("Coordinate 1 (", ve[1], "%)"), ylab=paste0("Coordinate 2 (", ve[2], "%)"), main="Site")
ordiellipse(scores, conditions$site, label=F, col=c("#018571","#80cdc1","#dfc27d","#a6611a"))
legend("topright", legend=c("Emerald", "Rainbow", "Star", "MacN"), fill = c("#018571","#80cdc1","#dfc27d","#a6611a"), bty="n")
# legend symbols are drawn in black; the LC symbol (pch 25) is filled to match the filled (site-coloured) LC points,
# which distinguishes it from the open CC symbol (pch 6)
legend("topleft", legend=c("CC", "CH", "LC", "LH"), pch=c(6,2,25,17), bty="n", pt.bg = c(NA,NA,"black",NA))
plot(scores[,1], scores[,2],col=c("#92c5de","#0571b0","#f4a582","#ca0020")[as.numeric(as.factor(conditions$treat))],pch=c(15,0,1,16)[as.numeric((as.factor(conditions$site)))], xlab=paste0("Coordinate 1 (", ve[1], "%)"), ylab=paste0("Coordinate 2 (", ve[2], "%)"), main="Treatment")
ordiellipse(scores, conditions$treat, label=F, col=c("#92c5de","#0571b0","#f4a582","#ca0020"))
legend("topleft", legend=c("CC", "CH", "LC", "LH"), fill = c("#92c5de","#f4a582","#0571b0","#ca0020"), bty="n")
legend("topright", legend=c("Emerald", "Rainbow", "Star", "MacN"), pch=c(15,0,1,16), bty="n")
})

# PCo's 2 and 3
save_plot(file="../../outputs/transcriptomics/ofav/limma/PCoA_ofav_host_PC3.pdf", width=12, height=6, {
par(mfrow=c(1,2))
plot(scores[,2], scores[,3],col=c("#018571","#80cdc1","#dfc27d","#a6611a")[as.numeric(as.factor(conditions$site))],pch=c(6,25,2,17)[as.numeric(as.factor(conditions$treat))], bg= c("#018571","#80cdc1","#dfc27d","#a6611a")[as.numeric(as.factor(conditions$site))], xlab=paste0("Coordinate 2 (", ve[2], "%)"), ylab=paste0("Coordinate 3 (", ve[3], "%)"), main="Site")
ordiellipse(scores, conditions$site, label=F, col=c("#018571","#80cdc1","#dfc27d","#a6611a"))
legend("topright", legend=c("Emerald", "Rainbow", "Star", "MacN"), fill = c("#018571","#80cdc1","#dfc27d","#a6611a"), bty="n")
# legend symbols are drawn in black; the LC symbol (pch 25) is filled to match the filled (site-coloured) LC points,
# which distinguishes it from the open CC symbol (pch 6)
legend("topleft", legend=c("CC", "CH", "LC", "LH"), pch=c(6,2,25,17), bty="n", pt.bg = c(NA,NA,"black",NA))
plot(scores[,2], scores[,3],col=c("#92c5de","#0571b0","#f4a582","#ca0020")[as.numeric(as.factor(conditions$treat))],pch=c(15,0,1,16)[as.numeric((as.factor(conditions$site)))], xlab=paste0("Coordinate 2 (", ve[2], "%)"), ylab=paste0("Coordinate 3 (", ve[3], "%)"), main="Treatment")
ordiellipse(scores, conditions$treat, label=F, col=c("#92c5de","#0571b0","#f4a582","#ca0020"))
legend("topleft", legend=c("CC", "CH", "LC", "LH"), fill = c("#92c5de","#f4a582","#0571b0","#ca0020"), bty="n")
legend("topright", legend=c("Emerald", "Rainbow", "Star", "MacN"), pch=c(15,0,1,16), bty="n")
})

# PCo's 3 and 4
save_plot(file="../../outputs/transcriptomics/ofav/limma/PCoA_ofav_host_PC4.pdf", width=12, height=6, {
par(mfrow=c(1,2))
plot(scores[,3], scores[,4],col=c("#018571","#80cdc1","#dfc27d","#a6611a")[as.numeric(as.factor(conditions$site))],pch=c(6,25,2,17)[as.numeric(as.factor(conditions$treat))], bg= c("#018571","#80cdc1","#dfc27d","#a6611a")[as.numeric(as.factor(conditions$site))], xlab=paste0("Coordinate 3 (", ve[3], "%)"), ylab=paste0("Coordinate 4 (", ve[4], "%)"), main="Site")
ordiellipse(scores, conditions$site, label=F, col=c("#018571","#80cdc1","#dfc27d","#a6611a"))
legend("topright", legend=c("Emerald", "Rainbow", "Star", "MacN"), fill = c("#018571","#80cdc1","#dfc27d","#a6611a"), bty="n")
# legend symbols are drawn in black; the LC symbol (pch 25) is filled to match the filled (site-coloured) LC points,
# which distinguishes it from the open CC symbol (pch 6)
legend("topleft", legend=c("CC", "CH", "LC", "LH"), pch=c(6,2,25,17), bty="n", pt.bg = c(NA,NA,"black",NA))
plot(scores[,3], scores[,4],col=c("#92c5de","#0571b0","#f4a582","#ca0020")[as.numeric(as.factor(conditions$treat))],pch=c(15,0,1,16)[as.numeric((as.factor(conditions$site)))], xlab=paste0("Coordinate 3 (", ve[3], "%)"), ylab=paste0("Coordinate 4 (", ve[4], "%)"), main="Treatment")
ordiellipse(scores, conditions$treat, label=F, col=c("#92c5de","#0571b0","#f4a582","#ca0020"))
legend("topleft", legend=c("CC", "CH", "LC", "LH"), fill = c("#92c5de","#f4a582","#0571b0","#ca0020"), bty="n")
legend("topright", legend=c("Emerald", "Rainbow", "Star", "MacN"), pch=c(15,0,1,16), bty="n")
})

# neighbor-joining tree of samples (based on significant PCo's):
save_plot(file="../../outputs/transcriptomics/ofav/limma/PCoA_tree.pdf", width=10, height=10, {
tre=nj(dist(scores[,1:4]))
plot(tre,cex=0.8)
})

# formal analysis of variance in distance matrices:
# fragments cut from the same parent colony (genotype) are not independent, so
# 1) treatment is tested WITHIN colonies (permutations restricted to within each genotype)
# 2) urban vs reef is tested at the COLONY level (colony = replicate), after removing treatment effects
library(permute)
conditions$genotype <- factor(conditions$genotype)
set.seed(12935)
ad=adonis2(t(vsd)~genotype+treat, data=conditions, method="manhattan", by="terms",
           permutations=how(nperm=1e6, blocks=conditions$genotype))
# genotype is constant within its own permutation blocks, so its p-value is not meaningful here;
# its R2 is the variance explained by parent colony (which includes site/habitat)
ad$`Pr(>F)`[1] <- NA
ad
write.csv(ad, file = "../../outputs/transcriptomics/ofav/limma/PERMANOVA_output.csv")

# urban vs reef: remove treatment effects, average fragments within each colony, then test colonies
# treatment effects are estimated WITHIN colonies (~ genotype + treat): colonies are not equally represented in
# every treatment (outliers/exclusions), so a treat-only fit would absorb colony (and habitat) differences into
# the treatment estimates. Only the fitted treatment contribution is subtracted, leaving colony differences intact.
X_adj <- model.matrix(~ genotype + treat, data=conditions)
fit_adj <- lm(t(vsd) ~ genotype + treat, data=conditions)
treat_terms <- grep("^treat", colnames(X_adj), value=TRUE)
vsd_adj <- vsd - t(X_adj[, treat_terms, drop=FALSE] %*% coef(fit_adj)[treat_terms, , drop=FALSE])
colony <- droplevels(conditions$genotype)
colony_means <- t(sapply(levels(colony), function(g) rowMeans(vsd_adj[, colony==g, drop=FALSE])))
colony_habitat <- factor(as.character(conditions$habitat[match(rownames(colony_means), as.character(colony))]), levels=c("reef","urban"))
table(colony_habitat) # number of colonies per habitat
set.seed(12935)
ad_habitat=adonis2(colony_means~colony_habitat, method="manhattan", permutations=99999)
ad_habitat
write.csv(ad_habitat, file = "../../outputs/transcriptomics/ofav/limma/PERMANOVA_habitat_colony.csv")

# overall site test at the colony level (four sites; urban vs reef above is its planned 1-df comparison)
colony_site <- factor(as.character(conditions$site[match(rownames(colony_means), as.character(colony))]), levels=c("Emerald","Rainbow","Star","MacN"))
table(colony_site) # number of colonies per site
set.seed(12935)
ad_site=adonis2(colony_means~colony_site, method="manhattan", permutations=99999)
ad_site
write.csv(ad_site, file = "../../outputs/transcriptomics/ofav/limma/PERMANOVA_site_colony.csv")

# creating pie chart to represent PERMANOVA results
cols=c("grey60","orange","white")
save_plot(file="../../outputs/transcriptomics/ofav/limma/PERMANOVA_pie.pdf", width=6, height=6, {
pie(ad$R2[1:3],labels=c("colony (genotype)","treatment","residual"),col=cols,main="colony vs treatment")
})
```


# SITE MODEL HELPERS

```{r site-model-helpers}
# Functions used by DIFFERENTIAL EXPRESSION, FRONT-LOADING and the sensitivity sections (run this chunk first when
# running those by hand). Install once: BiocManager::install("variancePartition")
library(limma)
library(edgeR)
library(variancePartition) # dream; loaded after limma
library(BiocParallel)
site_model_param <- MulticoreParam(max(1, parallel::detectCores() - 2))
# degrees of freedom for the dream tests: "Kenward-Roger" (more exact with few colonies; the method emmeans uses for
# the lmer physiology contrasts) or "Satterthwaite" (several times faster, slightly liberal for the site contrasts).
# The two are compared in SENSITIVITY: SATTERTHWAITE. Kenward-Roger makes this script slow to knit.
dream_ddf <- "Kenward-Roger"

# fit_site_model(): mixed model for each gene with site (four levels) and treatment as fixed effects and parent
# colony (genotype) as a random effect (voom precision weights + dream, variancePartition; Kenward-Roger degrees of freedom), i.e. the same structure
# as the lmer physiology models. Each gene has its own colony variance, and the degrees of freedom of each contrast
# reflect its replication (colonies for site contrasts, fragments for treatment contrasts).
# Returns one table per contrast.
#   counts: raw count matrix (genes x samples); meta: one row per sample with id, site, treat and genotype
#   genes:  NULL = filter genes with edgeR filterByExpr (minimum group size from the fixed-effects design);
#           otherwise the gene IDs to test (used by the sensitivity fits so they test the primary gene set)
fit_site_model <- function(counts, meta, genes = NULL, ddf = dream_ddf) {
  stopifnot(identical(colnames(counts), as.character(meta$id)))
  meta <- data.frame(id = as.character(meta$id),
                     site = factor(as.character(meta$site), levels = c("Emerald", "Rainbow", "Star", "MacN")),
                     treat = factor(as.character(meta$treat), levels = c("CC", "LC", "CH", "LH")),
                     genotype = droplevels(factor(meta$genotype)))
  rownames(meta) <- meta$id
  dge <- DGEList(as.matrix(counts))
  keep <- if (is.null(genes)) filterByExpr(dge, design = model.matrix(~ 0 + site + treat, meta)) else rownames(dge) %in% genes
  dge <- normLibSizes(dge[keep, , keep.lib.sizes=FALSE])
  form <- ~ 0 + site + treat + (1 | genotype)
  v <- voomWithDreamWeights(dge, form, meta, BPPARAM = site_model_param)
  # LC, CH and LH vs CC are the model's own treatment coefficients (CC is the reference level), so they are read
  # directly from the fit; the other contrasts are combinations of coefficients
  own_coef <- c(LC_CC = "treatLC", CH_CC = "treatCH", LH_CC = "treatLH")
  L <- makeContrastsDream(form, meta, contrasts = c(
    CH_LC = "treatCH - treatLC", LH_CH = "treatLH - treatCH", LH_LC = "treatLH - treatLC",
    urban_reef      = "(siteStar + siteMacN)/2 - (siteEmerald + siteRainbow)/2",
    MacN_reef       = "siteMacN - (siteEmerald + siteRainbow)/2",
    Star_reef       = "siteStar - (siteEmerald + siteRainbow)/2",
    MacN_Star       = "siteMacN - siteStar",
    Rainbow_Emerald = "siteRainbow - siteEmerald",
    Star_Emerald    = "siteStar - siteEmerald",
    Star_Rainbow    = "siteStar - siteRainbow",
    MacN_Emerald    = "siteMacN - siteEmerald",
    MacN_Rainbow    = "siteMacN - siteRainbow"))
  fit <- dream(v, form, meta, L = L, ddf = ddf, BPPARAM = site_model_param)
  fit <- variancePartition::eBayes(fit)
  all_contrasts <- c(treat_contrasts, site_contrasts)
  tt <- lapply(setNames(all_contrasts, all_contrasts), function(k) {
    x <- variancePartition::topTable(fit, coef = if (k %in% names(own_coef)) own_coef[[k]] else k, number = Inf, sort.by = "none")
    stopifnot(setequal(rownames(x), rownames(dge)))
    if (is.null(x$t)) x$t <- x$z.std # moderated t (named z.std in some variancePartition versions)
    x[rownames(dge), ]
  })
  list(tt = tt, genes_tested = nrow(dge), samples = table(meta$site, meta$treat), ddf = ddf)
}
treat_contrasts <- c("LC_CC", "CH_CC", "LH_CC", "CH_LC", "LH_CH", "LH_LC")
# urban vs reef, each urban site vs the reef sites, and all six pairwise site comparisons
site_contrasts  <- c("urban_reef", "MacN_reef", "Star_reef", "MacN_Star", "Rainbow_Emerald",
                     "Star_Emerald", "Star_Rainbow", "MacN_Emerald", "MacN_Rainbow")

# as_res(): contrast table with the column names used by the sections below
# (log2 fold change, moderated t, p and BH-adjusted p)
as_res <- function(x) data.frame(AveExpr = x$AveExpr, log2FoldChange = x$logFC, stat = x$t, pvalue = x$P.Value,
                                 padj = x$adj.P.Val, row.names = rownames(x))
```


# DIFFERENTIAL EXPRESSION (SITE MODEL)

```{r differential-expression}
# One model for all differential expression tests, with the same structure as the physiology models:
# site (four levels) and treatment as fixed effects and parent colony (genotype) as a random effect, fitted for each
# gene with dream (see SITE MODEL HELPERS), so colonies, not fragments, are the replicates for site.
# - treatment: the six pairwise contrasts among CC, LC, CH and LH
# - urban vs reef: planned contrast (Star + MacN)/2 - (Emerald + Rainbow)/2
# - sites: each urban site vs the reef sites, and all six pairwise site comparisons
# History: until 2026-10-05 treatments were tested with DESeq2 (~ genotype + treat) and urban vs reef with limma
# (duplicateCorrelation). The limma colony blocking uses one within-colony correlation for all genes and fragment-
# based degrees of freedom, which overstated the evidence for between-colony (site, urban vs reef) differences;
# both earlier models are compared with this one in SENSITIVITY: MODEL CHOICE.
load("../../outputs/transcriptomics/ofav/limma/initial.RData")
library(DESeq2)
meta <- as.data.frame(colData(dds))
table(site = tapply(as.character(meta$site), meta$genotype, `[`, 1)) # colonies per site
de <- fit_site_model(counts(dds), meta)
de$samples      # samples per site and treatment
de$genes_tested # genes passing the expression filter
de_tt <- de$tt

# DEGs (adjusted p < 0.1) per contrast
de_summary <- do.call(rbind, lapply(names(de_tt), function(k) {
  x <- de_tt[[k]]; d <- x$adj.P.Val < 0.1
  data.frame(contrast = k, genes_tested = nrow(x), DEGs = sum(d), up = sum(d & x$logFC > 0), down = sum(d & x$logFC < 0))
}))
de_summary
write.csv(de_summary, file = "../../outputs/transcriptomics/ofav/limma/DE_summary.csv", row.names = FALSE)

# treatment contrasts: one table per contrast (log2 fold change, moderated t, p and BH-adjusted p; the column names
# log2FoldChange/stat/pvalue/padj are those used by the sections below)
for (k in treat_contrasts) {
  r <- as_res(de_tt[[k]])
  assign(k, r)
  assign(paste0("degs_", k), rownames(r)[r$padj < 0.1])
}
save(LC_CC, CH_CC, LH_CC, CH_LC, LH_CH, LH_LC,
     degs_LC_CC, degs_CH_CC, degs_LH_CC, degs_CH_LC, degs_LH_CH, degs_LH_LC, file="../../outputs/transcriptomics/ofav/limma/pvals.RData")

# urban vs reef (planned contrast) and the site contrasts
urban_reef <- de_tt$urban_reef
degs_urban_reef <- row.names(urban_reef)[urban_reef$adj.P.Val < 0.1]
length(degs_urban_reef)                             # DEGs, padj < 0.1
table(direction = ifelse(urban_reef[degs_urban_reef, "logFC"] > 0, "higher in urban", "lower in urban"))
site_tt <- de_tt[site_contrasts]
save(urban_reef, degs_urban_reef, site_tt, file = "../../outputs/transcriptomics/ofav/limma/urban_reef.RData")

# are the urban vs reef DEGs shared by both urban sites? (each urban site vs the mean of the reef sites)
hab <- degs_urban_reef
m <- site_tt$MacN_reef[hab, ]; s <- site_tt$Star_reef[hab, ]; h <- sign(urban_reef[hab, "logFC"])
urban_site_split <- data.frame(
  urban_reef_DEGs = length(hab),
  both_sites_same_direction = sum(sign(m$logFC) == h & sign(s$logFC) == h),
  both_sites_p_below_0.05 = sum(sign(m$logFC) == h & sign(s$logFC) == h & m$P.Value < 0.05 & s$P.Value < 0.05),
  MacN_only_p_below_0.05 = sum(sign(m$logFC) == h & m$P.Value < 0.05 & !(sign(s$logFC) == h & s$P.Value < 0.05)),
  Star_only_p_below_0.05 = sum(sign(s$logFC) == h & s$P.Value < 0.05 & !(sign(m$logFC) == h & m$P.Value < 0.05)),
  neither_p_below_0.05 = sum(!(sign(m$logFC) == h & m$P.Value < 0.05) & !(sign(s$logFC) == h & s$P.Value < 0.05)),
  MacN_vs_Star_logFC_r = round(cor(m$logFC, s$logFC), 3),
  median_abs_logFC_MacN = round(median(abs(m$logFC)), 3),
  median_abs_logFC_Star = round(median(abs(s$logFC)), 3),
  MacN_effect_larger = sum(abs(m$logFC) > abs(s$logFC)))
t(urban_site_split)
write.csv(urban_site_split, file = "../../outputs/transcriptomics/ofav/limma/urban_reef_by_urban_site.csv", row.names = FALSE)

# per-gene table of the site contrasts (genes with adjusted p < 0.1 in any of them)
site_genes <- data.frame(gene = rownames(urban_reef))
for (k in site_contrasts) {
  site_genes[[paste0(k, "_logFC")]] <- site_tt[[k]]$logFC
  site_genes[[paste0(k, "_P")]]     <- site_tt[[k]]$P.Value
  site_genes[[paste0(k, "_padj")]]  <- site_tt[[k]]$adj.P.Val
}
site_genes <- site_genes[apply(site_genes[, paste0(site_contrasts, "_padj")] < 0.1, 1, any), ]
nrow(site_genes)
write.csv(site_genes, file = "../../outputs/transcriptomics/ofav/limma/site_contrasts_genes.csv", row.names = FALSE)

# size-factor normalised counts for the gene boxplots in the orthofinder script (no DESeq2 model is fitted)
dds <- estimateSizeFactors(dds)
save(dds, file="../../outputs/transcriptomics/ofav/limma/realModels.RData")
```


# SENSITIVITY: MODEL CHOICE

```{r sensitivity-model-choice, eval=run_sensitivity}
# Sensitivity section: runs in the knit only when run_sensitivity is TRUE (setup chunk); can also be run by hand
# after SITE MODEL HELPERS once the script has been knit.
# The primary model (dream) is compared with the two earlier approaches:
#   1) DESeq2 (~ genotype + treat): treatments, parent colony as a fixed term
#   2) limma-voom site model with duplicateCorrelation (one within-colony correlation for all genes): all contrasts
# and 3) the share of each gene's variance due to parent colony is summarised for the genes the methods disagree on.
library(DESeq2)
load("../../outputs/transcriptomics/ofav/limma/initial.RData")
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData")
primary_res <- c(mget(treat_contrasts), lapply(site_tt, as_res))
genes <- rownames(urban_reef)

# 1) treatments: DESeq2 with parent colony as a fixed blocking term
dds_deseq <- DESeq(dds, parallel=TRUE)
model_treat <- do.call(rbind, lapply(treat_contrasts, function(k) {
  pri <- primary_res[[k]]
  des <- as.data.frame(results(dds_deseq, contrast = c("treat", sub("_.*", "", k), sub(".*_", "", k))))
  des_degs <- rownames(des)[!is.na(des$padj) & des$padj < 0.1]
  pri_degs <- rownames(pri)[pri$padj < 0.1]
  common <- intersect(rownames(pri), rownames(des)[!is.na(des$log2FoldChange)])
  both <- intersect(pri_degs, des_degs)
  data.frame(contrast = k,
             dream_DEGs = length(pri_degs),
             DESeq2_DEGs = length(des_degs),
             DESeq2_DEGs_in_dream_gene_set = length(intersect(des_degs, rownames(pri))),
             shared = length(both),
             shared_same_direction = sum(sign(pri[both, "log2FoldChange"]) == sign(des[both, "log2FoldChange"])),
             logFC_r = round(cor(pri[common, "log2FoldChange"], des[common, "log2FoldChange"]), 3))
}))
model_treat
write.csv(model_treat, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_model_treatment_DESeq2.csv", row.names = FALSE)

# 2) all contrasts: limma-voom site model with duplicateCorrelation, same genes
meta_l <- as.data.frame(colData(dds))
meta_l$site     <- factor(as.character(meta_l$site), levels = c("Emerald", "Rainbow", "Star", "MacN"))
meta_l$treat    <- factor(as.character(meta_l$treat), levels = c("CC", "LC", "CH", "LH"))
meta_l$genotype <- droplevels(factor(meta_l$genotype))
rownames(meta_l) <- as.character(meta_l$id)
des_l <- model.matrix(~ 0 + site + treat, meta_l)
colnames(des_l) <- sub("^site", "", colnames(des_l))
dge_l <- normLibSizes(DGEList(counts(dds)[genes, ]))
v_l  <- voom(dge_l, des_l)
dc_l <- duplicateCorrelation(v_l, des_l, block = meta_l$genotype)
v_l  <- voom(dge_l, des_l, block = meta_l$genotype, correlation = dc_l$consensus)
dc_l <- duplicateCorrelation(v_l, des_l, block = meta_l$genotype)
dc_l$consensus # the single within-colony correlation used for every gene
fit_l <- lmFit(v_l, des_l, block = meta_l$genotype, correlation = dc_l$consensus)
con_l <- makeContrasts(LC_CC = treatLC, CH_CC = treatCH, LH_CC = treatLH,
                       CH_LC = treatCH - treatLC, LH_CH = treatLH - treatCH, LH_LC = treatLH - treatLC,
                       urban_reef      = (Star + MacN)/2 - (Emerald + Rainbow)/2,
                       MacN_reef       = MacN - (Emerald + Rainbow)/2,
                       Star_reef       = Star - (Emerald + Rainbow)/2,
                       MacN_Star       = MacN - Star,
                       Rainbow_Emerald = Rainbow - Emerald,
                       Star_Emerald    = Star - Emerald,
                       Star_Rainbow    = Star - Rainbow,
                       MacN_Emerald    = MacN - Emerald,
                       MacN_Rainbow    = MacN - Rainbow,
                       levels = des_l)
fit_l <- limma::eBayes(contrasts.fit(fit_l, con_l))
limma_res <- lapply(setNames(names(primary_res), names(primary_res)),
                    function(k) as_res(limma::topTable(fit_l, coef = k, number = Inf, sort.by = "none")))
model_limma <- do.call(rbind, lapply(names(primary_res), function(k) {
  pri <- primary_res[[k]]; lim <- limma_res[[k]][rownames(pri), ]
  pri_degs <- rownames(pri)[pri$padj < 0.1]; lim_degs <- rownames(lim)[lim$padj < 0.1]
  both <- intersect(pri_degs, lim_degs)
  data.frame(contrast = k, genes_tested = nrow(pri), dream_DEGs = length(pri_degs), limma_DEGs = length(lim_degs),
             shared = length(both),
             shared_same_direction = sum(sign(pri[both, "log2FoldChange"]) == sign(lim[both, "log2FoldChange"])),
             dream_only = length(setdiff(pri_degs, lim_degs)), limma_only = length(setdiff(lim_degs, pri_degs)),
             logFC_r = round(cor(pri$log2FoldChange, lim$log2FoldChange), 3))
}))
model_limma
write.csv(model_limma, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_model_limma_duplicateCorrelation.csv", row.names = FALSE)

# 3) share of variance explained by parent colony (alongside site and treatment; variancePartition needs all
# categorical terms as random here) for the genes found by both methods, by dream only and by limma only
vp <- fitExtractVarPartModel(v_l, ~ (1 | site) + (1 | treat) + (1 | genotype), meta_l, BPPARAM = site_model_param)
colony_var <- do.call(rbind, lapply(c("LC_CC", "CH_CC", "urban_reef", "MacN_reef"), function(k) {
  pri <- primary_res[[k]]; lim <- limma_res[[k]]
  pri_degs <- rownames(pri)[pri$padj < 0.1]; lim_degs <- rownames(lim)[lim$padj < 0.1]
  sets <- list(all_genes = genes, both = intersect(pri_degs, lim_degs),
               dream_only = setdiff(pri_degs, lim_degs), limma_only = setdiff(lim_degs, pri_degs))
  do.call(rbind, lapply(names(sets), function(n)
    data.frame(contrast = k, gene_set = n, genes = length(sets[[n]]),
               median_colony_variance_share = if (length(sets[[n]])) round(median(vp[sets[[n]], "genotype"]), 3) else NA)))
}))
colony_var
write.csv(colony_var, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_model_colony_variance.csv", row.names = FALSE)
```


# SENSITIVITY: SATTERTHWAITE

```{r sensitivity-satterthwaite, eval=run_sensitivity}
# Sensitivity section: runs in the knit only when run_sensitivity is TRUE (setup chunk); can also be run by hand
# after SITE MODEL HELPERS once the script has been knit.
# The primary model uses Kenward-Roger degrees of freedom (as emmeans does for the lmer physiology contrasts).
# The faster Satterthwaite approximation is fitted here for comparison; it is slightly liberal for the site
# contrasts, which rest on about five colonies per site. Same samples, genes and contrasts.
library(DESeq2)
load("../../outputs/transcriptomics/ofav/limma/initial.RData")
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData")
primary_res <- c(mget(treat_contrasts), lapply(site_tt, as_res))
de_sat <- fit_site_model(counts(dds), as.data.frame(colData(dds)), genes = rownames(urban_reef), ddf = "Satterthwaite")
model_sat <- do.call(rbind, lapply(names(primary_res), function(k) {
  pri <- primary_res[[k]]; sat <- as_res(de_sat$tt[[k]])[rownames(pri), ]
  pri_degs <- rownames(pri)[pri$padj < 0.1]; sat_degs <- rownames(sat)[sat$padj < 0.1]
  data.frame(contrast = k, Kenward_Roger_DEGs = length(pri_degs), Satterthwaite_DEGs = length(sat_degs),
             shared = length(intersect(pri_degs, sat_degs)),
             Kenward_Roger_only = length(setdiff(pri_degs, sat_degs)), Satterthwaite_only = length(setdiff(sat_degs, pri_degs)),
             p_value_r_log10 = round(cor(log10(pri$pvalue), log10(sat$pvalue)), 4))
}))
model_sat
write.csv(model_sat, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_model_Satterthwaite.csv", row.names = FALSE)
```


# SENSITIVITY: POOLED MODEL (SECOND-COHORT RESAMPLES INCLUDED)

```{r sensitivity-pooled, eval=run_sensitivity}
# Sensitivity section: runs in the knit only when run_sensitivity is TRUE (setup chunk); can also be run by hand
# after SITE MODEL HELPERS once the script has been knit.
# The primary analysis uses first-cohort samples only (see MODEL DESIGN and OUTLIERS). Here the site model (dream) and the
# PERMANOVAs are refit with the 10 second-cohort LH resamples added back (same outlier removals, same genes as the
# primary analysis) and compared with the primary results.
library(DESeq2)
library(vegan)
library(permute)
load("../../outputs/transcriptomics/ofav/limma/initial.RData")    # design (primary), design_pooled
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")      # primary treatment contrasts
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData") # primary urban vs reef and site contrasts
table(treat = design_pooled$treat, resample = !(design_pooled$id %in% design$id))

# pooled dataset: raw counts for the pooled samples
counts <- read.table("../../data/transcriptomics/ofav/allcounts_host.txt")
counts_p <- counts[, design_pooled$id]
stopifnot(identical(colnames(counts_p), design_pooled$id))
countData_p <- counts_p[rowSums(counts_p) >= 10, ]
# variance-stabilised counts for the pooled PERMANOVAs below
dds_pooled <- DESeqDataSetFromMatrix(countData=countData_p, colData=design_pooled, design=~ genotype+treat)
vsd_p <- assay(varianceStabilizingTransformation(dds_pooled, blind=TRUE))

# 1) site model (treatment, urban vs reef and site contrasts) on the pooled samples
de_pooled <- fit_site_model(counts_p, design_pooled, genes = rownames(urban_reef))
res_pooled <- lapply(de_pooled$tt, as_res)
primary_res <- c(mget(treat_contrasts), lapply(site_tt, as_res))
compare <- function(n, primary, pooled) {
  s_1 <- row.names(primary)[primary$padj < 0.1]
  s_p <- row.names(pooled)[pooled$padj < 0.1]
  data.frame(contrast = n, primary_DEGs = length(s_1), pooled_DEGs = length(s_p),
             shared = length(intersect(s_1, s_p)),
             primary_only = length(setdiff(s_1, s_p)), pooled_only = length(setdiff(s_p, s_1)),
             LFC_r = round(cor(primary$log2FoldChange, pooled$log2FoldChange[match(row.names(primary), row.names(pooled))],
                               use="complete.obs"), 3))
}
sensitivity <- do.call(rbind, lapply(names(res_pooled), function(n) compare(n, primary_res[[n]], res_pooled[[n]])))
sensitivity
write.csv(sensitivity, file="../../outputs/transcriptomics/ofav/limma/sensitivity_pooled_DEGs.csv", row.names=FALSE)

# 2) PERMANOVAs: treatment within colonies, and urban vs reef and site on colony means (as in the primary analysis)
cond_p <- design_pooled
cond_p$treat    <- factor(cond_p$treat, levels = c("CC", "LC", "CH", "LH"))
cond_p$genotype <- droplevels(factor(cond_p$genotype))
set.seed(12935)
ad_pooled <- adonis2(t(vsd_p)~genotype+treat, data=cond_p, method="manhattan", by="terms",
                     permutations=how(nperm=1e6, blocks=cond_p$genotype))
ad_pooled$`Pr(>F)`[1] <- NA
ad_pooled
write.csv(ad_pooled, file="../../outputs/transcriptomics/ofav/limma/sensitivity_pooled_PERMANOVA.csv")

X_p <- model.matrix(~ genotype + treat, data=cond_p)
fit_adj_p <- lm(t(vsd_p) ~ genotype + treat, data=cond_p)
tt_p <- grep("^treat", colnames(X_p), value=TRUE)
vsd_adj_p <- vsd_p - t(X_p[, tt_p, drop=FALSE] %*% coef(fit_adj_p)[tt_p, , drop=FALSE])
colony_p <- cond_p$genotype
colony_means_p <- t(sapply(levels(colony_p), function(g) rowMeans(vsd_adj_p[, colony_p==g, drop=FALSE])))
colony_habitat_p <- factor(as.character(cond_p$habitat[match(rownames(colony_means_p), as.character(colony_p))]), levels=c("reef","urban"))
set.seed(12935)
ad_habitat_pooled <- adonis2(colony_means_p~colony_habitat_p, method="manhattan", permutations=99999)
ad_habitat_pooled
write.csv(ad_habitat_pooled, file="../../outputs/transcriptomics/ofav/limma/sensitivity_pooled_PERMANOVA_habitat.csv")
colony_site_p <- factor(as.character(cond_p$site[match(rownames(colony_means_p), as.character(colony_p))]), levels=c("Emerald","Rainbow","Star","MacN"))
set.seed(12935)
ad_site_pooled <- adonis2(colony_means_p~colony_site_p, method="manhattan", permutations=99999)
ad_site_pooled
write.csv(ad_site_pooled, file="../../outputs/transcriptomics/ofav/limma/sensitivity_pooled_PERMANOVA_site.csv")

save(design_pooled, res_pooled, sensitivity, ad_pooled, ad_habitat_pooled, ad_site_pooled,
     file="../../outputs/transcriptomics/ofav/limma/sensitivity_pooled.RData")
```


# SENSITIVITY: COHORT EFFECT WITHIN LH

```{r sensitivity-cohort, eval=run_sensitivity}
# Sensitivity section: runs in the knit only when run_sensitivity is TRUE (setup chunk); by hand, run after
# SENSITIVITY: POOLED MODEL.
# Direct test of the cohort effect (sampling date 23 Apr vs 14 May, tank 8 vs tank 4, first vs repeat biopsy)
# within the LH treatment. Only colonies with LH samples from BOTH cohorts are used, and colony is in the model
# (limma-voom, ~ genotype + cohort; site is constant within a colony, so the site model does not apply here),
# so the cohort term compares the two cohorts within the same parent colonies.
# Result of the earlier DESeq2 run (2026-09-30), which led to the first-cohort primary analysis: 10 samples from
# 5 colonies, ~2,300 cohort DEGs (padj < 0.1) with 4 residual df; 12.5% of genes p < 0.05. LH vs CC DEGs found only
# in the first-cohort model carried strong cohort effects (55% p < 0.05), i.e. pooling dilutes first-cohort LH
# responses. limma run (2026-10-06 knit): 1,970 cohort DEGs, 22% of genes p < 0.05; LH vs CC DEGs found only in the
# first-cohort model: 69% with cohort p < 0.05 (17% for genes found in both models).
# Note: the LFC-shift correlation is largely mechanical (pooling only adds cohort-2 samples).
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")              # primary (first-cohort) contrasts
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData")         # primary gene set
load("../../outputs/transcriptomics/ofav/limma/sensitivity_pooled.RData") # design_pooled, res_pooled
counts <- read.table("../../data/transcriptomics/ofav/allcounts_host.txt")

# LH samples (both cohorts) from the pooled dataset, restricted to colonies sampled in both cohorts
lh_cd <- design_pooled[design_pooled$treat == "LH", ]
lh_cd$cohort <- factor(ifelse(lh_cd$tank == "8", "first", "second"), levels = c("first", "second"))
shared_col <- intersect(lh_cd$genotype[lh_cd$cohort == "first"], lh_cd$genotype[lh_cd$cohort == "second"])
lh_cd <- lh_cd[lh_cd$genotype %in% shared_col, ]
lh_cd$genotype <- droplevels(factor(lh_cd$genotype))
table(colony = lh_cd$genotype, cohort = lh_cd$cohort) # every colony should have one sample per cohort
lh_cd[, c("id", "sampleid", "site", "cohort")]
stopifnot(all(table(lh_cd$genotype, lh_cd$cohort) >= 1))

des_c <- model.matrix(~ genotype + cohort, lh_cd)
dge_c <- normLibSizes(DGEList(as.matrix(counts[rownames(urban_reef), as.character(lh_cd$id)])))
fit_c <- limma::eBayes(lmFit(voom(dge_c, des_c), des_c))
res_cohort <- as_res(limma::topTable(fit_c, coef = "cohortsecond", number = Inf, sort.by = "none"))
cohort_degs <- rownames(res_cohort)[res_cohort$padj < 0.1]
length(cohort_degs)

# proportion of genes with no cohort effect (Storey-type estimate from the p-value distribution)
p_c <- res_cohort$pvalue
pi0 <- min(1, mean(p_c > 0.5) * 2)
pi0
hist(p_c, breaks = 40, main = "LH cohort 2 vs cohort 1 (within colony)", xlab = "p-value")

# do the LH vs CC DEGs found only when the resamples are pooled carry cohort effects?
sig <- function(r) rownames(r)[!is.na(r$padj) & r$padj < 0.1]
lhcc_pooled <- res_pooled$LH_CC
lhcc_first  <- LH_CC
gene_sets <- list(all_genes          = rownames(res_cohort),
                  LH_CC_both         = intersect(sig(lhcc_pooled), sig(lhcc_first)),
                  LH_CC_pooled_only  = setdiff(sig(lhcc_pooled), sig(lhcc_first)),
                  LH_CC_first_only   = setdiff(sig(lhcc_first), sig(lhcc_pooled)))
cohort_summary <- do.call(rbind, lapply(names(gene_sets), function(n) {
  r <- res_cohort[gene_sets[[n]], ]
  data.frame(gene_set = n, genes = length(gene_sets[[n]]),
             frac_cohort_p05 = round(mean(r$pvalue < 0.05, na.rm = TRUE), 3),
             median_abs_cohort_LFC = round(median(abs(r$log2FoldChange), na.rm = TRUE), 3),
             cohort_DEGs = sum(gene_sets[[n]] %in% cohort_degs))
}))
cohort_summary

# does the cohort difference line up with how pooling shifts the LH vs CC fold changes?
shift <- lhcc_pooled$log2FoldChange - lhcc_first$log2FoldChange[match(rownames(lhcc_pooled), rownames(lhcc_first))]
cohort_lfc <- res_cohort$log2FoldChange[match(rownames(lhcc_pooled), rownames(res_cohort))]
lfc_shift_r <- cor(cohort_lfc, shift, use = "complete.obs")
lfc_shift_r

cohort_test <- data.frame(samples = nrow(lh_cd), colonies = nlevels(lh_cd$genotype), cohort_DEGs = length(cohort_degs),
                          frac_p05 = round(mean(p_c < 0.05), 3), pi0 = round(pi0, 3), LFC_shift_r = round(lfc_shift_r, 3))
cohort_test
write.csv(cohort_test, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_cohort_LH.csv", row.names = FALSE)
write.csv(cohort_summary, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_cohort_LH_genesets.csv", row.names = FALSE)
```


# SENSITIVITY: LH vs CH ROBUSTNESS

```{r sensitivity-lh-ch, eval=run_sensitivity}
# Sensitivity section: runs in the knit only when run_sensitivity is TRUE (setup chunk); can also be run by hand
# after SITE MODEL HELPERS once the script has been knit.
# Ofav_135 (So5-4, the only Star Island LH sample) was removed as an outlier because it alone drove the LH vs CH
# contrast (earlier DESeq2 colony model, 2026-09-30: 2,089 DEGs with it, 19 without). The same checks with the
# site model, each on the primary gene set. Result (2026-10-06 knit): with Ofav_135 added back, LH vs CH 84 DEGs
# (primary 0) and LH vs CC 3,959 (primary 1,914). Leave-one-out: LH vs CH stays at 0-2 DEGs except without Ofav_137
# (Mo5-4): 293; LH vs CC ranges from 994 to 3,234 DEGs. With six LH samples the LH contrasts depend on which
# samples are included, so LH results should be reported with that caution:
#   1) Ofav_135 added back to the primary samples
#   2) leave-one-out: each remaining LH sample dropped in turn
# If LH vs CH stays about the same in every leave-one-out fit and changes only when Ofav_135 is added, the
# contrast depends on that sample alone.
library(DESeq2)
load("../../outputs/transcriptomics/ofav/limma/initial.RData")    # dds, design (primary)
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")      # primary contrasts
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData") # primary gene set
genes <- rownames(urban_reef)
sig <- function(r) rownames(r)[!is.na(r$padj) & r$padj < 0.1]
summarise_fit <- function(label, de_fit, n_lh) {
  out <- lapply(c("LH_CH", "LH_CC", "LH_LC"), function(k) {
    r <- as_res(de_fit$tt[[k]])
    primary <- get(k)
    data.frame(DEGs = length(sig(r)), shared_with_primary = length(intersect(sig(r), sig(primary))),
               LFC_r = round(cor(primary$log2FoldChange, r$log2FoldChange[match(rownames(primary), rownames(r))],
                                 use = "complete.obs"), 3))
  })
  names(out) <- c("LH_CH", "LH_CC", "LH_LC")
  data.frame(fit = label, LH_n = n_lh, do.call(cbind, lapply(names(out), function(k) setNames(out[[k]], paste(k, names(out[[k]]), sep = "_")))))
}
meta <- data.frame(id = as.character(design$id), sampleid = as.character(design$sampleid), site = as.character(design$site),
                   treat = as.character(design$treat), genotype = as.character(design$genotype), stringsAsFactors = FALSE)
stopifnot(identical(colnames(counts(dds)), meta$id))
lh_ids <- meta$id[meta$treat == "LH"]
robust <- data.frame(fit = "primary", LH_n = length(lh_ids),
                     LH_CH_DEGs = length(sig(LH_CH)), LH_CH_shared_with_primary = length(sig(LH_CH)), LH_CH_LFC_r = 1,
                     LH_CC_DEGs = length(sig(LH_CC)), LH_CC_shared_with_primary = length(sig(LH_CC)), LH_CC_LFC_r = 1,
                     LH_LC_DEGs = length(sig(LH_LC)), LH_LC_shared_with_primary = length(sig(LH_LC)), LH_LC_LFC_r = 1)

# 1) Ofav_135 added back
counts_all  <- read.table("../../data/transcriptomics/ofav/allcounts_host.txt")
full_design <- read.csv("../../data/transcriptomics/ofav/design_ofav.csv", head=TRUE)
d135 <- full_design[full_design$id == "Ofav_135", ]
stopifnot(nrow(d135) == 1, d135$treat == "LH")
meta_135 <- rbind(meta, data.frame(id = "Ofav_135", sampleid = as.character(d135$sampleid), site = as.character(d135$site),
                                   treat = as.character(d135$treat), genotype = as.character(d135$genotype)))
counts_135 <- cbind(counts(dds)[genes, ], Ofav_135 = counts_all[genes, "Ofav_135"])
robust <- rbind(robust, summarise_fit(paste0("with Ofav_135 (", d135$sampleid, ") added back"),
                                      fit_site_model(counts_135, meta_135, genes = genes), length(lh_ids) + 1))

# 2) leave one LH sample out
for (i in lh_ids) {
  fit_i <- fit_site_model(counts(dds)[genes, meta$id != i], meta[meta$id != i, ], genes = genes)
  robust <- rbind(robust, summarise_fit(paste0("without ", i, " (", meta$sampleid[meta$id == i], ")"), fit_i, length(lh_ids) - 1))
}
robust
write.csv(robust, file = "../../outputs/transcriptomics/ofav/limma/sensitivity_LH_CH_robustness.csv", row.names = FALSE)
```


# FRONT-LOADING TEST

```{r front-loading}
# Do urban colonies hold heat-response genes at a higher baseline, and respond less to heat (front-loading,
# Barshis et al. 2013)? The site model above has no habitat x treatment term and its urban vs reef contrast is
# averaged over treatments, so this is tested with a separate dream model with one mean per habitat x treatment
# group (parent colony as a random effect; same samples and genes):
#   baseline        = urban - reef in the control treatment (CC)
#   heat_reef       = CH - CC in reef colonies;  heat_urban = CH - CC in urban colonies
#   heat_by_habitat = heat_urban - heat_reef (negative = smaller increase in urban colonies)
# Heat-response genes are taken from the primary model (CH vs CC over all colonies, adjusted p < 0.1), so the gene
# set is not chosen from the reef controls alone. Front-loading predicts that, among genes that go UP under heat,
# urban colonies have higher control expression (baseline > 0) and a smaller response (heat_by_habitat < 0), more
# often than for genes in general. Note that baseline and heat_by_habitat share the control groups with opposite
# signs, so they are negatively related for every gene; the comparison with all genes (and with heat-down genes)
# is what carries the result, not the two signs agreeing within a gene.
load("../../outputs/transcriptomics/ofav/limma/initial.RData")
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")
library(DESeq2)
meta_f <- as.data.frame(colData(dds))
meta_f <- data.frame(id = as.character(meta_f$id),
                     habitat = ifelse(meta_f$site %in% c("Star", "MacN"), "urban", "reef"),
                     treat = as.character(meta_f$treat), genotype = droplevels(factor(meta_f$genotype)))
rownames(meta_f) <- meta_f$id
stopifnot(identical(colnames(counts(dds)), meta_f$id))
meta_f$group <- factor(paste(meta_f$habitat, meta_f$treat, sep = "_"),
                       levels = as.vector(t(outer(c("reef", "urban"), c("CC", "LC", "CH", "LH"), paste, sep = "_"))))
table(meta_f$habitat, meta_f$treat) # samples per habitat and treatment
stopifnot(all(table(meta_f$group) > 0))
dge_f  <- normLibSizes(DGEList(counts(dds)[rownames(CH_CC), ]))
form_f <- ~ 0 + group + (1 | genotype)
v_f <- voomWithDreamWeights(dge_f, form_f, meta_f, BPPARAM = site_model_param)
L_f <- makeContrastsDream(form_f, meta_f, contrasts = c(
  baseline        = "groupurban_CC - groupreef_CC",
  heat_reef       = "groupreef_CH - groupreef_CC",
  heat_urban      = "groupurban_CH - groupurban_CC",
  heat_by_habitat = "(groupurban_CH - groupurban_CC) - (groupreef_CH - groupreef_CC)"))
fit_f <- variancePartition::eBayes(dream(v_f, form_f, meta_f, L = L_f, ddf = dream_ddf, BPPARAM = site_model_param))
fl <- lapply(setNames(colnames(L_f), colnames(L_f)), function(k)
  as_res(variancePartition::topTable(fit_f, coef = k, number = Inf, sort.by = "none")[rownames(CH_CC), ]))

# DEGs (adjusted p < 0.1) per contrast
fl_degs <- do.call(rbind, lapply(names(fl), function(k) {
  d <- fl[[k]]$padj < 0.1
  data.frame(contrast = k, DEGs = sum(d), up = sum(d & fl[[k]]$log2FoldChange > 0), down = sum(d & fl[[k]]$log2FoldChange < 0))
}))
fl_degs
write.csv(fl_degs, file = "../../outputs/transcriptomics/ofav/limma/frontloading_DEGs.csv", row.names = FALSE)

# baseline and response differences within heat-response gene sets, compared with all genes
heat_up   <- rownames(CH_CC)[CH_CC$padj < 0.1 & CH_CC$log2FoldChange > 0]
heat_down <- rownames(CH_CC)[CH_CC$padj < 0.1 & CH_CC$log2FoldChange < 0]
fl_sets <- list(all_genes = rownames(CH_CC), heat_up = heat_up, heat_down = heat_down)
fl_summary <- do.call(rbind, lapply(names(fl_sets), function(n) {
  b <- fl$baseline[fl_sets[[n]], ]; i <- fl$heat_by_habitat[fl_sets[[n]], ]
  data.frame(gene_set = n, genes = nrow(b),
             share_higher_in_urban_at_control = round(mean(b$log2FoldChange > 0), 3),
             median_baseline_logFC = round(median(b$log2FoldChange), 3),
             baseline_DEGs_higher_in_urban = sum(b$padj < 0.1 & b$log2FoldChange > 0),
             baseline_DEGs_lower_in_urban = sum(b$padj < 0.1 & b$log2FoldChange < 0),
             share_smaller_increase_in_urban = round(mean(i$log2FoldChange < 0), 3),
             median_heat_by_habitat_logFC = round(median(i$log2FoldChange), 3),
             heat_by_habitat_DEGs_negative = sum(i$padj < 0.1 & i$log2FoldChange < 0),
             heat_by_habitat_DEGs_positive = sum(i$padj < 0.1 & i$log2FoldChange > 0),
             # candidates: higher at control in urban (p < 0.05) and a smaller increase under heat in urban
             frontloaded_pattern = sum(b$log2FoldChange > 0 & b$pvalue < 0.05 & i$log2FoldChange < 0))
}))
fl_summary
write.csv(fl_summary, file = "../../outputs/transcriptomics/ofav/limma/frontloading_summary.csv", row.names = FALSE)

# per-gene table for the heat-up genes
fl_genes <- data.frame(gene = heat_up, heat_logFC_all_colonies = CH_CC[heat_up, "log2FoldChange"],
                       baseline_logFC = fl$baseline[heat_up, "log2FoldChange"], baseline_p = fl$baseline[heat_up, "pvalue"],
                       baseline_padj = fl$baseline[heat_up, "padj"],
                       heat_reef_logFC = fl$heat_reef[heat_up, "log2FoldChange"],
                       heat_urban_logFC = fl$heat_urban[heat_up, "log2FoldChange"],
                       heat_by_habitat_logFC = fl$heat_by_habitat[heat_up, "log2FoldChange"],
                       heat_by_habitat_p = fl$heat_by_habitat[heat_up, "pvalue"],
                       heat_by_habitat_padj = fl$heat_by_habitat[heat_up, "padj"])
write.csv(fl_genes, file = "../../outputs/transcriptomics/ofav/limma/frontloading_heat_up_genes.csv", row.names = FALSE)

# permutation test of the baseline shift: genes are correlated, so the shares above cannot be tested gene by gene.
# Control (CC) samples only: variance-stabilised counts averaged per parent colony, urban minus reef difference in
# colony means per gene, and the share of heat-up (and heat-down) genes that are higher in urban colonies. The
# habitat labels are then shuffled among colonies (5,000 times) to give the distribution of that share when habitat
# has no effect. p (two-sided) = how often a shuffled share is at least as far from 0.5 as the observed one;
# p_vs_all_genes does the same for the difference between the gene set and all genes.
load("../../outputs/transcriptomics/ofav/limma/vsd.RData") # vsd, design (same samples, in the same order)
stopifnot(identical(sub("\\..*$", "", colnames(vsd)), as.character(design$id)))
cc_fl   <- design$treat == "CC"
geno_fl <- droplevels(factor(design$genotype[cc_fl]))
cm_fl   <- t(rowsum(t(vsd[rownames(CH_CC), cc_fl]), geno_fl) / as.vector(table(geno_fl))) # genes x colonies
hab_fl  <- ifelse(design$site[cc_fl] %in% c("Star", "MacN"), "urban", "reef")[match(colnames(cm_fl), geno_fl)]
table(hab_fl) # colonies with a control sample, per habitat
i_up <- match(heat_up, rownames(cm_fl)); i_down <- match(heat_down, rownames(cm_fl))
share_fl <- function(h) {
  d <- rowMeans(cm_fl[, h == "urban", drop = FALSE]) - rowMeans(cm_fl[, h == "reef", drop = FALSE])
  c(heat_up = mean(d[i_up] > 0), heat_down = mean(d[i_down] > 0), all_genes = mean(d > 0))
}
obs_fl <- share_fl(hab_fl)
set.seed(12935)
perm_fl <- replicate(5000, share_fl(sample(hab_fl)))
fl_perm <- do.call(rbind, lapply(c("heat_up", "heat_down"), function(s) data.frame(
  gene_set = s, genes = length(fl_sets[[s]]),
  colonies_reef = sum(hab_fl == "reef"), colonies_urban = sum(hab_fl == "urban"),
  share_higher_in_urban_at_control = round(obs_fl[[s]], 3),
  share_all_genes = round(obs_fl[["all_genes"]], 3),
  null_2.5 = round(quantile(perm_fl[s, ], 0.025)[[1]], 3), null_97.5 = round(quantile(perm_fl[s, ], 0.975)[[1]], 3),
  p = mean(abs(perm_fl[s, ] - 0.5) >= abs(obs_fl[[s]] - 0.5)),
  p_vs_all_genes = mean(abs(perm_fl[s, ] - perm_fl["all_genes", ]) >= abs(obs_fl[[s]] - obs_fl[["all_genes"]])))))
fl_perm
write.csv(fl_perm, file = "../../outputs/transcriptomics/ofav/limma/frontloading_permutation.csv", row.names = FALSE)
```


# VENN DIAGRAMS

```{r venn-diagrams}
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")
library(DESeq2)

# install.packages("VennDiagram")
library(VennDiagram)
invisible(futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")) # no VennDiagram .log file per plot

pairwise_treat=list("LC_CC"=degs_LC_CC, "CH_CC"=degs_CH_CC, "LH_CC"=degs_LH_CC)
# pairwise comparisons among treatments
venn_treat=venn.diagram(
  x = pairwise_treat,
  filename=NULL,
  col = "transparent",
  fill = c("#0571b0","#f4a582","#ca0020"),
  alpha = 0.5,
  label.col = c("#0571b0","white","#f4a582","white","black","white","#ca0020"),
  cex = 3.5,
  fontfamily = "sans",
  fontface = "bold",
  cat.default.pos = "text",
  cat.col =c("#0571b0","#f4a582","#ca0020"),
  cat.cex = 3.5,
  cat.fontfamily = "sans",
  cat.just = list(c(0.25,0),c(0.75,0),c(0.5,1))
)
save_plot(file="../../outputs/transcriptomics/ofav/limma/Venn_treatment.pdf", height=12, width=12, {
grid.draw(venn_treat)
})

# urban vs reef DEGs compared with the stress-response DEGs (does habitat differ in the same genes that respond to
# OA and/or heat?). Overlap tested with Fisher's exact test within the genes tested in BOTH analyses
# (all contrasts come from the same site model and gene set); same_direction = share of
# overlapping genes whose urban-minus-reef difference has the same sign as the stress response.
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData")
overlap_test <- function(k) {
  r <- get(k)
  tested <- intersect(rownames(r)[!is.na(r$padj)], rownames(urban_reef))
  a <- intersect(degs_urban_reef, tested)
  b <- intersect(get(paste0("degs_", k)), tested)
  ov <- intersect(a, b)
  tab <- matrix(c(length(ov), length(a) - length(ov), length(b) - length(ov),
                  length(tested) - length(a) - length(b) + length(ov)), nrow = 2)
  ft <- fisher.test(tab)
  data.frame(contrast = k, genes_tested = length(tested), urban_reef_DEGs = length(a), contrast_DEGs = length(b),
             overlap = length(ov), expected = round(length(a) * length(b) / length(tested), 1),
             odds_ratio = round(unname(ft$estimate), 2), p = signif(ft$p.value, 3),
             same_direction = if (length(ov)) round(mean(sign(urban_reef[ov, "logFC"]) == sign(r[ov, "log2FoldChange"])), 2) else NA)
}
habitat_overlap <- do.call(rbind, lapply(c("LC_CC", "CH_CC", "LH_CC"), overlap_test))
habitat_overlap
write.csv(habitat_overlap, file="../../outputs/transcriptomics/ofav/limma/urban_reef_vs_treatment_overlap.csv", row.names=FALSE)

venn_habitat=venn.diagram(
  x = list("Urban vs reef"=degs_urban_reef, "LC_CC"=degs_LC_CC, "CH_CC"=degs_CH_CC, "LH_CC"=degs_LH_CC),
  filename=NULL,
  col = "transparent",
  fill = c("#636363","#0571b0","#f4a582","#ca0020"),
  alpha = 0.5,
  label.col = "black",
  cex = 2.5,
  fontfamily = "sans",
  fontface = "bold",
  cat.col = c("#636363","#0571b0","#f4a582","#ca0020"),
  cat.cex = 2.5,
  cat.fontfamily = "sans"
)
save_plot(file="../../outputs/transcriptomics/ofav/limma/Venn_habitat_treatment.pdf", height=12, width=12, {
grid.draw(venn_habitat)
})
```



# GO/KOG EXPORT

```{r go-kog-export}
load("../../outputs/transcriptomics/ofav/limma/realModels.RData")
load("../../outputs/transcriptomics/ofav/limma/pvals.RData")

# urban vs reef (planned contrast from the site model)
load("../../outputs/transcriptomics/ofav/limma/urban_reef.RData")
# signed log p-values: -log(padj)* direction:
urban_reef.p=data.frame("gene"=row.names(urban_reef))
urban_reef.p$lpv=-log(urban_reef$adj.P.Val,10)
urban_reef.p$lpv[urban_reef$t<0]=urban_reef.p$lpv[urban_reef$t<0]*-1
urban_reef.p=urban_reef.p[!is.na(urban_reef.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(urban_reef.p)
write.csv(urban_reef.p,file="../../outputs/transcriptomics/ofav/limma/urban_reef_lpv.csv",row.names=F,quote=F)
save(urban_reef.p,file="../../outputs/transcriptomics/ofav/limma/urban_reef_lpv.RData")

# site contrasts (each urban site vs the reef sites, and the six pairwise site comparisons): signed log p-values
# for the KOG script (first site minus second, e.g. MacN_Star > 0 = higher at MacArthur North than at Star Island)
for (k in setdiff(names(site_tt), "urban_reef")) {
  x <- site_tt[[k]]
  site_lpv <- data.frame(gene = row.names(x), lpv = -log(x$adj.P.Val, 10) * ifelse(x$t < 0, -1, 1))
  site_lpv <- site_lpv[!is.na(site_lpv$lpv), ]
  write.csv(site_lpv, file = paste0("../../outputs/transcriptomics/ofav/limma/", k, "_lpv.csv"), row.names = FALSE, quote = FALSE)
}

# treatment contrasts

# signed log p-values: -log(padj)* direction:
source=LC_CC[!is.na(LC_CC$pvalue),]
LC_CC.p=data.frame("gene"=row.names(source))
LC_CC.p$lpv=-log(source[,"padj"],10)
LC_CC.p$lpv[source$stat<0]=LC_CC.p$lpv[source$stat<0]*-1
LC_CC.p=LC_CC.p[!is.na(LC_CC.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(LC_CC.p)
write.csv(LC_CC.p,file="../../outputs/transcriptomics/ofav/limma/LC_CC_lpv.csv",row.names=F,quote=F)
save(LC_CC.p,file="../../outputs/transcriptomics/ofav/limma/LC_CC_lpv.RData")

source=CH_CC[!is.na(CH_CC$pvalue),]
CH_CC.p=data.frame("gene"=row.names(source))
CH_CC.p$lpv=-log(source[,"padj"],10)
CH_CC.p$lpv[source$stat<0]=CH_CC.p$lpv[source$stat<0]*-1
CH_CC.p=CH_CC.p[!is.na(CH_CC.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(CH_CC.p)
write.csv(CH_CC.p,file="../../outputs/transcriptomics/ofav/limma/CH_CC_lpv.csv",row.names=F,quote=F)
save(CH_CC.p,file="../../outputs/transcriptomics/ofav/limma/CH_CC_lpv.RData")

source=LH_CC[!is.na(LH_CC$pvalue),]
LH_CC.p=data.frame("gene"=row.names(source))
LH_CC.p$lpv=-log(source[,"padj"],10)
LH_CC.p$lpv[source$stat<0]=LH_CC.p$lpv[source$stat<0]*-1
LH_CC.p=LH_CC.p[!is.na(LH_CC.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(LH_CC.p)
write.csv(LH_CC.p,file="../../outputs/transcriptomics/ofav/limma/LH_CC_lpv.csv",row.names=F,quote=F)
save(LH_CC.p,file="../../outputs/transcriptomics/ofav/limma/LH_CC_lpv.RData")

source=CH_LC[!is.na(CH_LC$pvalue),]
CH_LC.p=data.frame("gene"=row.names(source))
CH_LC.p$lpv=-log(source[,"padj"],10)
CH_LC.p$lpv[source$stat<0]=CH_LC.p$lpv[source$stat<0]*-1
CH_LC.p=CH_LC.p[!is.na(CH_LC.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(CH_LC.p)
write.csv(CH_LC.p,file="../../outputs/transcriptomics/ofav/limma/CH_LC_lpv.csv",row.names=F,quote=F)
save(CH_LC.p,file="../../outputs/transcriptomics/ofav/limma/CH_LC_lpv.RData")

source=LH_LC[!is.na(LH_LC$pvalue),]
LH_LC.p=data.frame("gene"=row.names(source))
LH_LC.p$lpv=-log(source[,"padj"],10)
LH_LC.p$lpv[source$stat<0]=LH_LC.p$lpv[source$stat<0]*-1
LH_LC.p=LH_LC.p[!is.na(LH_LC.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(LH_LC.p)
write.csv(LH_LC.p,file="../../outputs/transcriptomics/ofav/limma/LH_LC_lpv.csv",row.names=F,quote=F)
save(LH_LC.p,file="../../outputs/transcriptomics/ofav/limma/LH_LC_lpv.RData")

source=LH_CH[!is.na(LH_CH$pvalue),]
LH_CH.p=data.frame("gene"=row.names(source))
LH_CH.p$lpv=-log(source[,"padj"],10)
LH_CH.p$lpv[source$stat<0]=LH_CH.p$lpv[source$stat<0]*-1
LH_CH.p=LH_CH.p[!is.na(LH_CH.p$lpv),] # drop genes with no adjusted p-value (NA padj, e.g. independent filtering)
head(LH_CH.p)
write.csv(LH_CH.p,file="../../outputs/transcriptomics/ofav/limma/LH_CH_lpv.csv",row.names=F,quote=F)
save(LH_CH.p,file="../../outputs/transcriptomics/ofav/limma/LH_CH_lpv.RData")

save(urban_reef.p,
     LC_CC.p, CH_CC.p, LH_CC.p, CH_LC.p, LH_CH.p, LH_LC.p, file="../../outputs/transcriptomics/ofav/limma/exports.RData")
```


# DEG MATCHING

```{r deg-matching}
library(DESeq2)
library(tidyverse)
library(reshape2)
library(RColorBrewer)
library(ggplot2)
library(patchwork)
load("../../outputs/transcriptomics/ofav/limma/exports.RData")

# This section of code does several things: 1) join -log10(pval) across treatment comparisons, 2) filter by 0.1 pval cutoff (log10(0.1)=1), 3) adds gene annotations, and 4) then pulls on corresponding KOG classes

# Treatment
# stress treatments (LC, CH, and LH) versus control treatment (CC)
LC_CC.p %>%
  dplyr::inner_join(CH_CC.p, by = "gene", suffix = c(".LC_CC", ".CH_CC")) %>%
  dplyr::inner_join(LH_CC.p %>% dplyr::rename(lpv.LH_CC = lpv), by = "gene") %>%
  filter(abs(lpv.LC_CC) >= 1 & abs(lpv.CH_CC) >= 1 & abs(lpv.LH_CC) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  mutate(comparison="stress_control", .before="gene") -> commongenes_treatment

# exporting all DEGs matching across stress vs control treatments
write.csv(commongenes_treatment, file="../../outputs/transcriptomics/ofav/limma/commongenes_treatment.csv")

# pairwise treatment comparisons
LC_CC.p %>%
  dplyr::inner_join(CH_CC.p, by = "gene", suffix = c(".LC_CC", ".CH_CC")) %>%
  filter(abs(lpv.LC_CC) >= 1 & abs(lpv.CH_CC) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) -> bleaching_oa

LC_CC.p %>%
  dplyr::inner_join(LH_CC.p, by = "gene", suffix = c(".LC_CC", ".LH_CC")) %>%
  filter(abs(lpv.LC_CC) >= 1 & abs(lpv.LH_CC) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) -> oableaching_oa

CH_CC.p %>%
  dplyr::inner_join(LH_CC.p, by = "gene", suffix = c(".CH_CC", ".LH_CC")) %>%
  filter(abs(lpv.CH_CC) >= 1 & abs(lpv.LH_CC) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) -> bleaching_oableaching

# joining all matching DEGs into a single dataframe
commongenes_pairwise_treatment <- bind_rows(bleaching_oa,oableaching_oa,bleaching_oableaching)
write.csv(commongenes_pairwise_treatment, file="../../outputs/transcriptomics/ofav/limma/commongenes_pairwise_treatment.csv")

# Site
# urban vs reef DEGs (colony-level test; |lpv| >= 1 means padj < 0.1), with gene annotations and KOG classes
urban_reef.p %>%
  dplyr::rename(lpv.urban_reef = lpv) %>%
  filter(abs(lpv.urban_reef) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG = V2) %>%
              dplyr::select(-V1, -V2), by = c("gene" = "gene")) %>%
  mutate(comparison="urban_reef", .before="gene") -> urban_reef_degs

# exporting all urban vs reef DEGs
write.csv(urban_reef_degs, file="../../outputs/transcriptomics/ofav/limma/urban_reef_DEGs.csv")
```


# CORRELATION PLOT HELPERS

```{r correlation-plot-helpers}
# --- Unicode-safe symbols (avoid warnings on some devices) ---
use_unicode <- l10n_info()[["UTF-8"]] && capabilities("cairo")
LTE   <- if (use_unicode) "\u2264" else "<="
MINUS <- if (use_unicode) "\u2212" else "-"

# --- Parameters ---
inf_cap  <- 20
REL_NAME <- "Relationship"
SIG_NAME <- "p Value"

# Discordant filter knobs (used by both dataframe sections)
delta_cut_dd <- 2.0
ratio_cut_dd <- 100
high_cut_dd  <- 2.0   # ~ p <= 0.01
low_cut_dd   <- 1.3   # ~ p <= 0.05
min_sig      <- 1

# --- Relationship colors (colorblind-friendly) ---
REL_LEVELS <- c("Direct", "Inverse")
rel_cols   <- setNames(c("#009E73", "#D55E00"), REL_LEVELS)

# --- Significance tiers (shared by all plots) ---
SIG_LEVELS <- c("0.05 < p \u2264 0.1", "p \u2264 0.05", "p \u2264 1e\u22126")
sig_alpha  <- setNames(c(0.15, 0.90, 0.90), SIG_LEVELS)
sig_size   <- setNames(c(1.0,  1.2,  1.2),  SIG_LEVELS)
sig_stroke <- setNames(c(0.00, 0.20, 0.00), SIG_LEVELS)

# --- Shared helper functions ---
# Normalize any two named lpv columns into lpv_x / lpv_y, cap infinities, drop non-finite rows
clean_lpv_df <- function(df, col_x, col_y, cap = inf_cap) {
  if (is.null(df) || !nrow(df)) stop("Input data frame is empty.", call. = FALSE)
  if (!all(c(col_x, col_y) %in% names(df)))
    stop(sprintf("Columns not found: %s",
                 paste(setdiff(c(col_x, col_y), names(df)), collapse = ", ")), call. = FALSE)
  d <- df
  d$lpv_x <- as.numeric(d[[col_x]])
  d$lpv_y <- as.numeric(d[[col_y]])
  d$lpv_x[ is.infinite(d$lpv_x) & d$lpv_x > 0 ] <- cap
  d$lpv_x[ is.infinite(d$lpv_x) & d$lpv_x < 0 ] <- -cap
  d$lpv_y[ is.infinite(d$lpv_y) & d$lpv_y > 0 ] <- cap
  d$lpv_y[ is.infinite(d$lpv_y) & d$lpv_y < 0 ] <- -cap
  d <- d[is.finite(d$lpv_x) & is.finite(d$lpv_y), , drop = FALSE]
  d
}

safe_stats <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]; y <- y[ok]
  n <- sum(complete.cases(x, y))
  if (n >= 3 && sd(x) > 0 && sd(y) > 0) {
    ct <- suppressWarnings(cor.test(x, y, method = "pearson"))
    r  <- unname(ct$estimate); p <- ct$p.value
    list(r = r, p = p, n = n)
  } else list(r = NA_real_, p = NA_real_, n = n)
}

fmt_num <- function(x) {
  if (is.na(x)) "NA" else if (abs(x) < 1e-3) format(x, digits = 2, scientific = TRUE) else sprintf("%.3f", x)
}

sig_category3 <- function(lpv_x, lpv_y, levels_vec = SIG_LEVELS) {
  lv <- pmin(abs(lpv_x), abs(lpv_y))
  p  <- 10^(-lv)
  out <- dplyr::case_when(
    p <= 1e-6 ~ levels_vec[3],
    p <= 0.05 ~ levels_vec[2],
    TRUE      ~ levels_vec[1]
  )
  factor(out, levels = levels_vec)
}

compute_global_limits <- function(comparisons, min_span = 0.6, pad_frac = 0.05) {
  vals <- unlist(lapply(comparisons, function(entry) {
    d <- clean_lpv_df(entry$df, entry$col_x, entry$col_y)
    d <- d[pmax(abs(d$lpv_x), abs(d$lpv_y)) >= 1.0, ]
    c(d$lpv_x, d$lpv_y)
  }))
  max_abs <- max(abs(vals), na.rm = TRUE)
  if (!is.finite(max_abs)) max_abs <- min_span / 2
  lim <- max(max_abs * (1 + pad_frac), min_span / 2)
  c(-lim, lim)
}

make_panel <- function(df, title, col_x, col_y, x_lab, y_lab, global_lim) {
  dat <- clean_lpv_df(df, col_x, col_y)
  dat <- subset(dat, pmax(abs(dat$lpv_x), abs(dat$lpv_y)) >= 1.0)
  
  same <- (dat$lpv_x >= 0 & dat$lpv_y >= 0) | (dat$lpv_x <= 0 & dat$lpv_y <= 0)
  dat$rel_class <- factor(ifelse(same, "Direct", "Inverse"), levels = REL_LEVELS)
  dat$sig_cat   <- sig_category3(dat$lpv_x, dat$lpv_y, levels_vec = SIG_LEVELS)
  
  x_limits <- global_lim
  y_limits <- global_lim
  
  span       <- diff(global_lim)
  jitter_amt <- span * 0.006
  set.seed(42)
  dat$x_jit <- pmin(pmax(jitter(dat$lpv_x, amount = jitter_amt), global_lim[1]), global_lim[2])
  dat$y_jit <- pmin(pmax(jitter(dat$lpv_y, amount = jitter_amt), global_lim[1]), global_lim[2])
  
  st <- safe_stats(dat$lpv_x, dat$lpv_y)
  lab_text <- if (is.na(st$r)) {
    sprintf("n = %d", st$n)
  } else {
    sprintf("r = %.2f, p = %s, n = %d", st$r, fmt_num(st$p), st$n)
  }
  
  ref_levels <- c("Direct 1:1", "Inverse 1:1")
  
  ggplot(dat, aes(x = lpv_x, y = lpv_y)) +
    geom_hline(yintercept = 0, color = "grey85") +
    geom_vline(xintercept = 0, color = "grey85") +
    
    geom_point(
      aes(x = x_jit, y = y_jit, color = rel_class, fill = rel_class,
          alpha = sig_cat, stroke = sig_cat, size = sig_cat),
      shape = 21
    ) +
    
    geom_point(
      data = subset(dat, sig_cat == SIG_LEVELS[3]),
      aes(x = x_jit, y = y_jit, fill = rel_class, alpha = sig_cat, size = sig_cat),
      color = "grey15", shape = 21, stroke = 0.40
    ) +
    
    geom_point(
      data = data.frame(sig_cat = factor(SIG_LEVELS, levels = SIG_LEVELS)),
      mapping = aes(x = 0, y = 0, alpha = sig_cat),
      inherit.aes = FALSE, shape = 21, size = 0, stroke = 0, fill = NA, color = NA,
      show.legend = TRUE, na.rm = TRUE
    ) +
    
    geom_abline(intercept = 0, slope =  1, color = "firebrick", linewidth = 1,
                linetype = "dashed", alpha = 0.5, show.legend = FALSE) +
    geom_abline(intercept = 0, slope = -1, color = "firebrick", linewidth = 1,
                linetype = "dotted", alpha = 0.5, show.legend = FALSE) +
    
    geom_segment(
      data = data.frame(x = 0, y = 0, xend = 0, yend = 0,
                        ref = factor(ref_levels, levels = ref_levels)),
      mapping = aes(x = x, y = y, xend = xend, yend = yend, linetype = ref),
      inherit.aes = FALSE, color = "firebrick", linewidth = 1, alpha = 0.5,
      show.legend = TRUE
    ) +
    
    annotate("text", x = x_limits[1], y = y_limits[2], label = lab_text,
             hjust = 0, vjust = 1.1, size = 4.2) +
    
    coord_fixed(xlim = x_limits, ylim = y_limits, expand = FALSE, clip = "on") +
    
    scale_color_manual(values = rel_cols, breaks = REL_LEVELS, limits = REL_LEVELS,
                       drop = FALSE, guide = "none") +
    scale_discrete_manual("size", values = sig_size, breaks = SIG_LEVELS, limits = SIG_LEVELS,
                          guide = "none") +
    scale_fill_manual(values = rel_cols, breaks = REL_LEVELS, limits = REL_LEVELS,
                      drop = FALSE, name = REL_NAME) +
    scale_alpha_manual(values = sig_alpha, breaks = SIG_LEVELS, limits = SIG_LEVELS,
                       drop = FALSE, name = SIG_NAME) +
    scale_discrete_manual("stroke", values = sig_stroke,
                          breaks = SIG_LEVELS, limits = SIG_LEVELS, guide = "none") +
    scale_linetype_manual(
      name   = "Reference",
      values = c("Direct 1:1" = "dashed", "Inverse 1:1" = "dotted")
    ) +
    guides(
      fill  = guide_legend(order = 1,
                           override.aes = list(shape = 21, size = 1.6, linetype = 0,
                                               stroke = c(0.45, 0.45))),
      alpha = guide_legend(order = 2,
                           override.aes = list(shape = 21, fill = "grey50",
                                               color = c("grey30", "grey30", "grey15"),
                                               size = 1.6, linetype = 0,
                                               stroke = c(0.00, 0.20, 0.40))),
      linetype = guide_legend(order = 3,
                              override.aes = list(color = "firebrick", linewidth = 1, alpha = 0.5))
    ) +
    labs(
      x = str_wrap(paste0(x_lab, " (signed ", MINUS, "log10 p)"), width = 28),
      y = str_wrap(paste0(y_lab, " (signed ", MINUS, "log10 p)"), width = 28),
      title = str_wrap(title, width = 32)
    ) +
    theme_classic(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text  = element_text(color = "black"),
      legend.position = "right",
      axis.line = element_blank(),
      plot.margin = margin(8, 20, 8, 8)
    )
}

classify_for_filters <- function(df, col_x, col_y) {
  d <- clean_lpv_df(df, col_x, col_y)
  same <- (d$lpv_x >= 0 & d$lpv_y >= 0) | (d$lpv_x <= 0 & d$lpv_y <= 0)
  d$rel_class    <- ifelse(same, "Direct", "Inverse")
  d$abs_x        <- abs(d$lpv_x)
  d$abs_y        <- abs(d$lpv_y)
  d$delta_log10p <- abs(d$abs_x - d$abs_y)
  d$p_x          <- 10^(-d$abs_x)
  d$p_y          <- 10^(-d$abs_y)
  d$p_ratio      <- pmax(d$p_x, d$p_y) / pmin(d$p_x, d$p_y)
  d
}

rbind_align <- function(...) {
  dfs  <- list(...)
  cols <- unique(unlist(lapply(dfs, names)))
  dfs2 <- lapply(dfs, function(x) {
    if (is.null(x) || !nrow(x)) {
      return(as.data.frame(setNames(replicate(length(cols), logical(0), simplify = FALSE), cols)))
    }
    miss <- setdiff(cols, names(x))
    for (m in miss) x[[m]] <- NA
    x[, cols, drop = FALSE]
  })
  do.call(rbind, dfs2)
}

build_panels <- function(comparisons, ncol = 3) {
  global_lim <- compute_global_limits(comparisons)
  panels <- mapply(
    FUN      = function(title, entry) make_panel(entry$df, title, entry$col_x, entry$col_y,
                                                 entry$x_lab, entry$y_lab, global_lim),
    title    = names(comparisons),
    entry    = comparisons,
    SIMPLIFY = FALSE
  )
  nrow_grid <- ceiling(length(panels) / ncol)
  for (i in seq_along(panels)) {
    row <- ceiling(i / ncol)
    col <- i - (row - 1) * ncol
    if (col != 1) panels[[i]] <- panels[[i]] + theme(axis.title.y = element_blank())
    if (row != nrow_grid) panels[[i]] <- panels[[i]] + theme(axis.title.x = element_blank())
  }
  one_legend  <- cowplot::get_legend(panels[[1]] + theme(legend.position = "right"))
  legend_plot <- cowplot::ggdraw(one_legend)
  panels_noleg <- lapply(panels, function(p) p + theme(legend.position = "none"))
  grid  <- wrap_plots(panels_noleg, ncol = ncol)
  panel <- grid | legend_plot
  panel + plot_layout(widths = c(1, 0.30))
}

filter_discordant <- function(comparisons, group_col) {
  filter_one <- function(entry, group_name) {
    d         <- classify_for_filters(entry$df, entry$col_x, entry$col_y)
    meets_min <- pmax(d$abs_x, d$abs_y) >= min_sig
    
    inverse <- d[d$rel_class == "Inverse" & meets_min, , drop = FALSE]
    if (nrow(inverse)) {
      inverse$filter_type           <- "Inverse"
      inverse$relationship_category <- "Inverse"
      inverse$discord_direction     <- NA_character_
      inverse$delta_cut_used        <- NA_real_
      inverse$high_cut_used         <- NA_real_
      inverse$low_cut_used          <- NA_real_
      inverse$ratio_cut_used        <- NA_real_
      inverse[[group_col]]          <- group_name
    }
    
    strong_x_weak_y <- d$abs_x >= high_cut_dd & d$abs_y <= low_cut_dd
    strong_y_weak_x <- d$abs_y >= high_cut_dd & d$abs_x <= low_cut_dd
    cross_tier <- strong_x_weak_y | strong_y_weak_x
    dd_idx <- d$rel_class == "Direct" &
      d$delta_log10p >= delta_cut_dd &
      d$p_ratio      >= ratio_cut_dd &
      cross_tier & meets_min
    
    direct_discordant <- d[dd_idx, , drop = FALSE]
    if (nrow(direct_discordant)) {
      so <- direct_discordant$abs_x >= high_cut_dd & direct_discordant$abs_y <= low_cut_dd
      direct_discordant$filter_type           <- "DirectDiscordant"
      direct_discordant$relationship_category <- "Direct (discordant significance)"
      direct_discordant$discord_direction     <- ifelse(so,
                                                        paste0(entry$x_lab, " >> ", entry$y_lab),
                                                        paste0(entry$y_lab, " >> ", entry$x_lab))
      direct_discordant$delta_cut_used        <- delta_cut_dd
      direct_discordant$high_cut_used         <- high_cut_dd
      direct_discordant$low_cut_used          <- low_cut_dd
      direct_discordant$ratio_cut_used        <- ratio_cut_dd
      direct_discordant[[group_col]]          <- group_name
    }
    
    rbind_align(inverse, direct_discordant)
  }
  
  do.call(rbind_align, lapply(names(comparisons), function(nm) filter_one(comparisons[[nm]], nm)))
}
```


# CORRELATION PLOTS TREATMENT

```{r correlation-plots-treatment}
treatment_comparisons <- list(
  "Bleaching vs OA" = list(
    df    = bleaching_oa,
    col_x = "lpv.LC_CC", col_y = "lpv.CH_CC",
    x_lab = "OA",         y_lab = "Bleaching"),
  "OA + Bleaching vs OA" = list(
    df    = oableaching_oa,
    col_x = "lpv.LC_CC", col_y = "lpv.LH_CC",
    x_lab = "OA",         y_lab = "OA + Bleaching"),
  "OA + Bleaching vs Bleaching" = list(
    df    = bleaching_oableaching,
    col_x = "lpv.CH_CC", col_y = "lpv.LH_CC",
    x_lab = "Bleaching",  y_lab = "OA + Bleaching")
)

panel_treatment <- build_panels(treatment_comparisons, ncol = 3)
print(panel_treatment)
ggsave("../../outputs/transcriptomics/ofav/limma/commongenes correlation treatment.pdf",
       panel_treatment, width = 14, height = 5.5)
```


# CORRELATION DATAFRAME TREATMENT

```{r correlation-dataframe-treatment}
treatment_filtered_genes <- filter_discordant(treatment_comparisons, group_col = "treatment")

if (nrow(treatment_filtered_genes) > 0) {
  cat("Rows total:", nrow(treatment_filtered_genes), "\n")
  print(table(treatment_filtered_genes$relationship_category, useNA = "ifany"))
  print(table(treatment_filtered_genes$treatment, useNA = "ifany"))
} else {
  warning("treatment_filtered_genes is empty with current strict thresholds.")
}
dplyr::count(treatment_filtered_genes, treatment, relationship_category)
write.csv(treatment_filtered_genes,
          file = "../../outputs/transcriptomics/ofav/limma/commongenes correlation discordant treatment.csv")
```






# HEATMAPS

```{r heatmaps}
# first removing unannotated genes
commongenes_treatment %>%
  filter(!is.na(annot)) ->  commongenes_treatment_heatmap

urban_reef_degs %>%
  filter(!is.na(annot)) ->  urban_reef_heatmap

load("../../outputs/transcriptomics/ofav/limma/vsd.RData")
vsd_treatment <- subset(vsd, rownames(vsd) %in% commongenes_treatment_heatmap$gene)
# keep the significance metric in the same gene order as the vsd rows
commongenes_treatment_heatmap <- commongenes_treatment_heatmap[match(rownames(vsd_treatment), commongenes_treatment_heatmap$gene), ]
vsd_site <- subset(vsd, rownames(vsd) %in% urban_reef_heatmap$gene)
# keep the significance metric in the same gene order as the vsd rows
urban_reef_heatmap <- urban_reef_heatmap[match(rownames(vsd_site), urban_reef_heatmap$gene), ]

# Make sure the 'uniHeatmap.R' script is in your working directory
source("uniHeatmap.R")

# heatmap drawing with square cells; the page is sized to fit the cells.
# Up to 100 genes: 9-pt cells with gene names; more genes: smaller square cells (page ~12.5 in tall), names hidden.
draw_heatmap <- function(file, vsd, gene_names, metric, cutoff, sort, gaps_col = NULL) {
  n <- sum(metric <= cutoff, na.rm = TRUE)
  message(basename(file), ": ", n, " annotated genes")
  if (n < 2) return(invisible(NULL)) # skipped: too few genes pass this cutoff
  show_names <- n <= 100
  cell <- if (show_names) 9 else max(4, min(9, 900 / n))
  pdf_w <- (ncol(vsd) * cell + (if (show_names) 330 else 60) + 110) / 72
  pdf_h <- (n * cell + 140) / 72
  save_plot(file = file, width = pdf_w, height = pdf_h, {
    uniHeatmap(vsd = vsd, gene.names = gene_names, metric = metric, cutoff = cutoff, sort = sort, pdf = F,
               cellwidth = cell, cellheight = cell, fontsize_row = 7, fontsize_col = min(6, cell * 0.8),
               show_rownames = show_names, gaps_col = gaps_col)
  })
}
# shows the heatmaps drawn in a loop in the knitted HTML
show_heatmaps <- function(imgs) {
  imgs <- unlist(Filter(Negate(is.null), imgs))
  if (length(imgs) && isTRUE(getOption("knitr.in.progress"))) knitr::include_graphics(imgs, rel_path = FALSE, dpi = 150) else invisible(imgs)
}

# Treatment
# creating a lookup table of gene ID to gene annotations
gene_names <- as.data.frame(cbind(commongenes_treatment_heatmap$gene, commongenes_treatment_heatmap$annot))
# long annotations are shortened for the row labels (full annotations are in the DEG tables)
gene_names$V2 <- ifelse(nchar(gene_names$V2) > 60, paste0(substr(gene_names$V2, 1, 57), "..."), gene_names$V2)
# heatmaps of LH vs CC genes, columns in design order (grouped by treatment, gaps between treatments)
treat_gaps <- cumsum(rle(as.character(design$treat))$lengths)
treat_gaps <- treat_gaps[-length(treat_gaps)]
treat_files <- c("-1" = "0.1", "-3" = "0.001", "-6" = "1e-6")
hm_treatment <- lapply(c(-1, -3, -6), function(cut) # p < 0.1, 0.001, 1e-6; cutoffs with < 2 annotated genes are skipped
  draw_heatmap(file = paste0("../../outputs/transcriptomics/ofav/limma/heatmap_treatment_p", treat_files[as.character(cut)], ".pdf"),
               vsd = vsd_treatment, gene_names = gene_names,
               metric = -(abs(commongenes_treatment_heatmap$lpv.LH_CC)), # metric of gene significance
               cutoff = cut, sort = c(1:ncol(vsd_treatment)), gaps_col = treat_gaps))
show_heatmaps(hm_treatment)

# Urban vs reef
# creating a lookup table of gene ID to gene annotations
gene_names_site <- as.data.frame(cbind(urban_reef_heatmap$gene, urban_reef_heatmap$annot))
# long annotations are shortened for the row labels (full annotations are in urban_reef_DEGs.csv)
gene_names_site$V2 <- ifelse(nchar(gene_names_site$V2) > 60, paste0(substr(gene_names_site$V2, 1, 57), "..."), gene_names_site$V2)
# columns ordered by habitat, then site
# (uniHeatmap orders columns by order(sort), so 'sort' is a sorting key)
site_key <- 10 * as.numeric(factor(ifelse(design$site %in% c("Star","MacN"), "urban", "reef"), levels=c("reef","urban"))) +
            as.numeric(factor(as.character(design$site), levels=c("Emerald","Rainbow","Star","MacN")))
n_reef <- sum(site_key < 20) # reef columns come first; a gap separates reef from urban

hm_site <- lapply(c(-1, -2, -3, -6), function(cut) # p < 0.1, 0.01, 0.001, 1e-6; cutoffs with < 2 annotated genes are skipped
  draw_heatmap(file = paste0("../../outputs/transcriptomics/ofav/limma/heatmap_urban_reef_p1e", cut, ".pdf"),
               vsd = vsd_site, gene_names = gene_names_site,
               metric = -(abs(urban_reef_heatmap$lpv.urban_reef)), # metric of gene significance
               cutoff = cut, sort = site_key, gaps_col = n_reef)) # columns ordered by habitat and site
show_heatmaps(hm_site)
```


# KOG MATCHING

```{r kog-matching}
# Treatment
# LC vs CC 
commongenes_treatment %>%
  mutate(KOG = replace(KOG, KOG == "", NA)) %>%
  filter(lpv.LC_CC >= 1) %>%
  count(KOG) %>%
  rename("KOG" = KOG, "LC_CC_up" = n) -> KOG_LC_CC_up

commongenes_treatment %>%
  mutate(KOG = replace(KOG, KOG == "", NA)) %>%
  filter(lpv.LC_CC <= -1) %>%
  count(KOG) %>%
  rename("KOG" = KOG, "LC_CC_down" = n) -> KOG_LC_CC_down

# CH vs CC 
commongenes_treatment %>%
  mutate(KOG = replace(KOG, KOG == "", NA)) %>%
  filter(lpv.CH_CC >= 1) %>%
  count(KOG) %>%
  rename("KOG" = KOG, "CH_CC_up" = n) -> KOG_CH_CC_up

commongenes_treatment %>%
  mutate(KOG = replace(KOG, KOG == "", NA)) %>%
  filter(lpv.CH_CC <= -1) %>%
  count(KOG) %>%
  rename("KOG" = KOG, "CH_CC_down" = n) -> KOG_CH_CC_down

# LH vs CC 
commongenes_treatment %>%
  mutate(KOG = replace(KOG, KOG == "", NA)) %>%
  filter(lpv.LH_CC >= 1) %>%
  count(KOG) %>%
  rename("KOG" = KOG, "LH_CC_up" = n) -> KOG_LH_CC_up

commongenes_treatment %>%
  mutate(KOG = replace(KOG, KOG == "", NA)) %>%
  filter(lpv.LH_CC <= -1) %>%
  count(KOG) %>%
  rename("KOG" = KOG, "LH_CC_down" = n) -> KOG_LH_CC_down

# joining all KOG class sums in a single dataframe
KOG_LC_CC_up %>%
  full_join(KOG_CH_CC_up, by = "KOG") %>%
  full_join(KOG_LH_CC_up, by = "KOG") %>%
  full_join(KOG_LC_CC_down, by = "KOG") %>%
  full_join(KOG_CH_CC_down, by = "KOG") %>%
  full_join(KOG_LH_CC_down, by = "KOG") -> KOG_match
KOG_match <- KOG_match %>% mutate(across(-KOG, ~ tidyr::replace_na(.x, 0L)))

# melting dataframe for plotting
KOG_match %>%
  melt(id = "KOG") %>%
  rename(comparison = variable, sum = value) -> KOG_melt

# creating a custom color palette
colorCount = length(unique(KOG_match$KOG))
getPalette = colorRampPalette(brewer.pal(8, "Accent"))

# relative abundance plot
KOG_sum <- ggplot(KOG_melt, aes(fill = KOG, y = sum, x = comparison)) +
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(8, "Accent"))(colorCount)) +
  labs(x = "Comparison",
       y = "Number of DEGs") +
  theme_classic()
KOG_sum
ggsave("../../outputs/transcriptomics/ofav/limma/KOG_treatment.pdf", plot= KOG_sum, width=10, height=6, units="in", dpi=300)

# Urban vs reef
urban_reef_degs %>%
  mutate(KOG = replace(KOG, KOG == "", NA),
         direction = ifelse(lpv.urban_reef > 0, "higher in urban", "lower in urban")) %>%
  count(direction, KOG) -> KOG_urban_reef

# (skipped when there are no urban vs reef DEGs)
if (nrow(KOG_urban_reef) > 0) {
# creating a custom color palette
colorCount = length(unique(KOG_urban_reef$KOG))

# relative abundance plot
KOG_sum <- ggplot(KOG_urban_reef, aes(fill = KOG, y = n, x = direction)) +
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(8, "Accent"))(colorCount)) +
  labs(x = "Urban vs reef",
       y = "Number of DEGs") +
  guides(fill = guide_legend(ncol = 1)) +
  theme_classic()
print(KOG_sum)
ggsave("../../outputs/transcriptomics/ofav/limma/KOG_site.pdf", plot= KOG_sum, width=10, height=8, units="in", dpi=300)
} else cat("no urban vs reef DEGs: KOG_site.pdf not written\n")
```
