---
title: "Cross-Species Ortholog Expression — OrthoFinder Analysis"
author: "Michael Studivan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# -----------------------------------------------------------------------------
# Global chunk options
# echo = TRUE:     show code in rendered output
# warning = FALSE: suppress warnings in rendered document
#                  (warnings still print to console during knitting)
# message = FALSE: suppress package loading messages
# -----------------------------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 300
)
```


# PACKAGES

```{r packages}
library(tidyverse)
library(VennDiagram)
invisible(futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")) # no VennDiagram .log file per plot
library(pheatmap)
library(reshape2)
library(RColorBrewer)
library(ggplot2)
library(ggpubr)
library(patchwork)   
library(cowplot)
library(stringr)
```


# ORTHOFINDER

```{r orthofinder}
# Install orthofinder on your local machine using the tutorials (https://davidemms.github.io/menu/tutorials.html)

# Copy your translated protein fasta files (_out_PRO.fas) that you want to compare into a directory called 'orthofinder'
# If you have not already filtered by the longest contig per isogroup (by using fasta2SBH.pl during transcriptome annotation), follow step 7 of tutorial 2 above

# Run the following command in Terminal: 'orthofinder -f orthofinder/'
# Check the number of genes assigned to orthogroups (e.g., 'OrthoFinder assigned 39416 genes (89.6% of total) to 13781 orthogroups')
# Ideally, it should be >80%
```


# ORTHOLOGS

```{r orthologs}
# reading the OrthoFinder pairwise orthologs table
# NOTE: Ofaveolata_out_PRO__v__Ssiderea_out_PRO.tsv must be present. It is expected in data/transcriptomics/orthofinder/,
# but in the repo it currently lives in outputs/transcriptomics/orthofinder/, so that location is used as a fallback.
orthologs_file <- "../../data/transcriptomics/orthofinder/Ofaveolata_out_PRO__v__Ssiderea_out_PRO.tsv"
if (!file.exists(orthologs_file)) {
  orthologs_file <- "../../outputs/transcriptomics/orthofinder/Ofaveolata_out_PRO__v__Ssiderea_out_PRO.tsv"
  message("Ortholog table not found in data/; using ", orthologs_file)
}
if (!file.exists(orthologs_file)) stop("OrthoFinder ortholog table (Ofaveolata_out_PRO__v__Ssiderea_out_PRO.tsv) not found in data/ or outputs/")
orthologs <- read.table(file = orthologs_file, sep = "\t", header = TRUE, quote="", fill=FALSE)

# or, if loading in previously run data
# load("../../outputs/transcriptomics/orthofinder/orthofinder_DEGs.RData")
```


# DESEQ IMPORT TREATMENT

```{r deseq-import-treatment}
ofav_LC_CC_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/LC_CC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ofav_CH_CC_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/CH_CC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ofav_LH_CC_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/LH_CC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ofav_CH_LC_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/CH_LC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ofav_LH_CH_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/LH_CH_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ofav_LH_LC_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/LH_LC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ssid_LC_CC_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/LC_CC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)

ssid_CH_CC_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/CH_CC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)

ssid_LH_CC_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/LH_CC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)

ssid_CH_LC_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/CH_LC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)

ssid_LH_CH_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/LH_CH_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)

ssid_LH_LC_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/LH_LC_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)
```


# DESEQ IMPORT URBAN vs REEF

```{r deseq-import-site}
# urban vs reef (colony-level limma test from the DESeq2 scripts)
ofav_urban_reef_lpv <- read.csv(file = "../../outputs/transcriptomics/ofav/deseq2/urban_reef_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ofav" = lpv)

ssid_urban_reef_lpv <- read.csv(file = "../../outputs/transcriptomics/ssid/deseq2/urban_reef_lpv.csv") %>%
  select(gene, lpv) %>%
  rename("lpv_ssid" = lpv)
```


# DEG MATCHING TREATMENT

```{r deg-matching-treatment}
# This section of code does several things: 1) rename common orthologs, 2) join with signed -log10(padj), 3) filter by padj 0.1 cutoff (-log10(0.1)=1), 4) adds ofav and ssid gene annotations, and 5) then pulls on corresponding KOG classes

# LC vs CC for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_LC_CC_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_LC_CC_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="LC_CC", .before="Orthogroup") -> LC_CC
LC_CC$Orthogroup <- make.unique(LC_CC$Orthogroup, sep = "_") 

# CH vs CC for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_CH_CC_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_CH_CC_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="CH_CC", .before="Orthogroup") -> CH_CC
CH_CC$Orthogroup <- make.unique(CH_CC$Orthogroup, sep = "_") 

# LH vs CC for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_LH_CC_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_LH_CC_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="LH_CC", .before="Orthogroup") -> LH_CC
LH_CC$Orthogroup <- make.unique(LH_CC$Orthogroup, sep = "_") 

# CH vs LC for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_CH_LC_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_CH_LC_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="CH_LC", .before="Orthogroup") -> CH_LC
CH_LC$Orthogroup <- make.unique(CH_LC$Orthogroup, sep = "_") 

# LH vs CH for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_LH_CH_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_LH_CH_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="LH_CH", .before="Orthogroup") -> LH_CH
LH_CH$Orthogroup <- make.unique(LH_CH$Orthogroup, sep = "_") 

# LH vs LC for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_LH_LC_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_LH_LC_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="LH_LC", .before="Orthogroup") -> LH_LC
LH_LC$Orthogroup <- make.unique(LH_LC$Orthogroup, sep = "_") 

# joining all matching DEGs into a single dataframe
orthofinder_treatment <- bind_rows(LC_CC,CH_CC,LH_CC,CH_LC,LH_CH,LH_LC)
write.csv(orthofinder_treatment, file="../../outputs/transcriptomics/orthofinder/orthofinder_treatment.csv")
```


# DEG MATCHING URBAN vs REEF

```{r deg-matching-site}
# This section of code does several things: 1) rename common orthologs, 2) join with signed -log10(padj), 3) filter by padj 0.1 cutoff (-log10(0.1)=1), 4) adds ofav and ssid gene annotations, and 5) then pulls on corresponding KOG classes

# urban vs reef for both species
orthologs %>%
  rename("Protein_ofav" = 
           Ofaveolata_out_PRO, "Protein_ssid" = 	
           Ssiderea_out_PRO) %>%
  separate_rows(., Protein_ofav, sep = ",\\s*") %>%
  separate_rows(., Protein_ssid, sep = ",\\s*") %>%
  unique() %>%
  inner_join(ssid_urban_reef_lpv, by = c("Protein_ssid" = "gene")) %>%
  inner_join(ofav_urban_reef_lpv, by = c("Protein_ofav" = "gene")) %>%
  filter(abs(lpv_ssid) >= 1 & abs(lpv_ofav) >= 1) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ofav = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2kogClass.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     KOG_ssid = V2) %>%
              dplyr::select(-V1, -V2), by = c("Protein_ssid" = "gene")) %>%
  mutate(comparison="urban_reef", .before="Orthogroup") -> urban_reef
urban_reef$Orthogroup <- make.unique(urban_reef$Orthogroup, sep = "_") 

# all matching DEGs
orthofinder_site <- urban_reef
write.csv(orthofinder_site, file="../../outputs/transcriptomics/orthofinder/orthofinder_site.csv")
```


# SAVING DATAFRAMES

```{r saving-dataframes}
# saving dataframes
save(orthologs, LC_CC, CH_CC, LH_CC, CH_LC, LH_CH, LH_LC, urban_reef, file = "../../outputs/transcriptomics/orthofinder/orthofinder_DEGs.RData")

# if loading in previously run data
load("../../outputs/transcriptomics/orthofinder/orthofinder_DEGs.RData") 
```


# CORRELATION PLOT HELPERS

```{r correlation-plot-helpers}
# --- Unicode-safe symbols (avoid warnings on some devices) ---
use_unicode <- l10n_info()[["UTF-8"]] && capabilities("cairo")
LTE   <- if (use_unicode) "\u2264" else "<="
MINUS <- if (use_unicode) "\u2212" else "-"

# --- Parameters ---
inf_cap  <- 20
REL_NAME <- "Relationship"
SIG_NAME <- "padj"

# Discordant filter knobs (used by both dataframe sections)
delta_cut_dd <- 2.0
ratio_cut_dd <- 100
high_cut_dd  <- 2.0   # ~ padj <= 0.01
low_cut_dd   <- 1.3   # ~ padj <= 0.05
min_sig      <- 1

# --- Relationship colors (colorblind-friendly) ---
REL_LEVELS <- c("Direct", "Inverse")
rel_cols   <- setNames(c("#009E73", "#D55E00"), REL_LEVELS)

# --- Significance tiers (shared by all plots) ---
SIG_LEVELS <- c("0.05 < padj \u2264 0.1", "padj \u2264 0.05", "padj \u2264 1e\u22126")
sig_alpha  <- setNames(c(0.15, 0.90, 0.90), SIG_LEVELS)
sig_size   <- setNames(c(1.0,  1.2,  1.2),  SIG_LEVELS)
sig_stroke <- setNames(c(0.00, 0.20, 0.00), SIG_LEVELS)

# --- Shared helper functions ---
clean_lpv_df <- function(df, cap = inf_cap) {
  d <- df
  d$lpv_ofav <- as.numeric(d$lpv_ofav)
  d$lpv_ssid <- as.numeric(d$lpv_ssid)
  d$lpv_ofav[ is.infinite(d$lpv_ofav) & d$lpv_ofav > 0 ] <- cap
  d$lpv_ofav[ is.infinite(d$lpv_ofav) & d$lpv_ofav < 0 ] <- -cap
  d$lpv_ssid[ is.infinite(d$lpv_ssid) & d$lpv_ssid > 0 ] <- cap
  d$lpv_ssid[ is.infinite(d$lpv_ssid) & d$lpv_ssid < 0 ] <- -cap
  d <- d[is.finite(d$lpv_ofav) & is.finite(d$lpv_ssid), , drop = FALSE]
  d
}

safe_stats <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]; y <- y[ok]
  n <- sum(complete.cases(x, y))
  if (n >= 3 && sd(x) > 0 && sd(y) > 0) {
    ct <- suppressWarnings(cor.test(x, y, method = "pearson"))
    r  <- unname(ct$estimate); p <- ct$p.value
    list(r = r, p = p, n = n)
  } else list(r = NA_real_, p = NA_real_, n = n)
}

fmt_num <- function(x) {
  if (is.na(x)) "NA" else if (abs(x) < 1e-3) format(x, digits = 2, scientific = TRUE) else sprintf("%.3f", x)
}

sig_category3 <- function(lpv_ofav, lpv_ssid, levels_vec = SIG_LEVELS) {
  lv <- pmin(abs(lpv_ofav), abs(lpv_ssid))
  p  <- 10^(-lv)
  out <- dplyr::case_when(
    p <= 1e-6 ~ levels_vec[3],
    p <= 0.05 ~ levels_vec[2],
    TRUE      ~ levels_vec[1]
  )
  factor(out, levels = levels_vec)
}

compute_global_limits <- function(comparisons, min_span = 0.6, pad_frac = 0.05) {
  vals <- unlist(lapply(comparisons, function(df) {
    d <- clean_lpv_df(df)
    d <- d[pmax(abs(d$lpv_ofav), abs(d$lpv_ssid)) >= 1.0, ]
    c(d$lpv_ofav, d$lpv_ssid)
  }))
  max_abs <- max(abs(vals), na.rm = TRUE)
  if (!is.finite(max_abs)) max_abs <- min_span / 2
  lim <- max(max_abs * (1 + pad_frac), min_span / 2)
  c(-lim, lim)
}

make_panel <- function(df, title, global_lim) {
  dat <- clean_lpv_df(df)
  dat <- subset(dat, pmax(abs(dat$lpv_ofav), abs(dat$lpv_ssid)) >= 1.0)
  
  same <- (dat$lpv_ofav >= 0 & dat$lpv_ssid >= 0) | (dat$lpv_ofav <= 0 & dat$lpv_ssid <= 0)
  dat$rel_class <- factor(ifelse(same, "Direct", "Inverse"), levels = REL_LEVELS)
  dat$sig_cat   <- sig_category3(dat$lpv_ofav, dat$lpv_ssid, levels_vec = SIG_LEVELS)
  
  x_limits <- global_lim
  y_limits <- global_lim
  
  span       <- diff(global_lim)
  jitter_amt <- span * 0.006
  set.seed(42)
  dat$x_jit <- pmin(pmax(jitter(dat$lpv_ofav, amount = jitter_amt), global_lim[1]), global_lim[2])
  dat$y_jit <- pmin(pmax(jitter(dat$lpv_ssid, amount = jitter_amt), global_lim[1]), global_lim[2])
  
  st <- safe_stats(dat$lpv_ofav, dat$lpv_ssid)
  lab_text <- if (is.na(st$r)) {
    sprintf("n = %d", st$n)
  } else {
    sprintf("r = %.2f, p = %s, n = %d", st$r, fmt_num(st$p), st$n)
  }
  
  ref_levels <- c("Direct 1:1", "Inverse 1:1")
  
  ggplot(dat, aes(x = lpv_ofav, y = lpv_ssid)) +
    geom_hline(yintercept = 0, color = "grey85") +
    geom_vline(xintercept = 0, color = "grey85") +
    
    geom_point(
      aes(x = x_jit, y = y_jit, color = rel_class, fill = rel_class,
          alpha = sig_cat, stroke = sig_cat, size = sig_cat),
      shape = 21
    ) +
    
    geom_point(
      data = subset(dat, sig_cat == SIG_LEVELS[3]),
      aes(x = x_jit, y = y_jit, fill = rel_class, alpha = sig_cat, size = sig_cat),
      color = "grey15", shape = 21, stroke = 0.40
    ) +
    
    geom_point(
      data = data.frame(sig_cat = factor(SIG_LEVELS, levels = SIG_LEVELS)),
      mapping = aes(x = 0, y = 0, alpha = sig_cat),
      inherit.aes = FALSE, shape = 21, size = 0, stroke = 0, fill = NA, color = NA,
      show.legend = TRUE
    ) +
    
    geom_abline(intercept = 0, slope =  1, color = "firebrick", linewidth = 1,
                linetype = "dashed", alpha = 0.5, show.legend = FALSE) +
    geom_abline(intercept = 0, slope = -1, color = "firebrick", linewidth = 1,
                linetype = "dotted", alpha = 0.5, show.legend = FALSE) +
    
    geom_segment(
      data = data.frame(x = 0, y = 0, xend = 0, yend = 0,
                        ref = factor(ref_levels, levels = ref_levels)),
      mapping = aes(x = x, y = y, xend = xend, yend = yend, linetype = ref),
      inherit.aes = FALSE, color = "firebrick", linewidth = 1, alpha = 0.5,
      show.legend = TRUE
    ) +
    
    annotate("text", x = x_limits[1], y = y_limits[2], label = lab_text,
             hjust = 0, vjust = 1.1, size = 4.2) +
    
    coord_fixed(xlim = x_limits, ylim = y_limits, expand = FALSE, clip = "on") +
    
    scale_color_manual(values = rel_cols, breaks = REL_LEVELS, limits = REL_LEVELS,
                       drop = FALSE, guide = "none") +
    scale_discrete_manual("size", values = sig_size, breaks = SIG_LEVELS, limits = SIG_LEVELS,
                          guide = "none") +
    scale_fill_manual(values = rel_cols, breaks = REL_LEVELS, limits = REL_LEVELS,
                      drop = FALSE, name = REL_NAME) +
    scale_alpha_manual(values = sig_alpha, breaks = SIG_LEVELS, limits = SIG_LEVELS,
                       drop = FALSE, name = SIG_NAME) +
    scale_discrete_manual("stroke", values = sig_stroke,
                          breaks = SIG_LEVELS, limits = SIG_LEVELS, guide = "none") +
    scale_linetype_manual(
      name   = "Reference",
      values = c("Direct 1:1" = "dashed", "Inverse 1:1" = "dotted")
    ) +
    guides(
      fill  = guide_legend(order = 1,
                           override.aes = list(shape = 21, size = 1.6, linetype = 0,
                                               stroke = c(0.45, 0.45))),
      alpha = guide_legend(order = 2,
                           override.aes = list(shape = 21, fill = "grey50",
                                               color = c("grey30", "grey30", "grey15"),
                                               size = 1.6, linetype = 0,
                                               stroke = c(0.00, 0.20, 0.40))),
      linetype = guide_legend(order = 3,
                              override.aes = list(color = "firebrick", linewidth = 1, alpha = 0.5))
    ) +
    labs(
      x = paste0("O. faveolata (signed ", MINUS, "log10 padj)"),
      y = paste0("S. siderea (signed ", MINUS, "log10 padj)"),
      title = str_wrap(title, width = 32)
    ) +
    theme_classic(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text  = element_text(color = "black"),
      legend.position = "right",
      axis.line = element_blank(),
      plot.margin = margin(8, 20, 8, 8)
    )
}

classify_for_filters <- function(df) {
  d <- clean_lpv_df(df)
  same <- (d$lpv_ofav >= 0 & d$lpv_ssid >= 0) | (d$lpv_ofav <= 0 & d$lpv_ssid <= 0)
  d$rel_class    <- ifelse(same, "Direct", "Inverse")
  d$abs_ofav     <- abs(d$lpv_ofav)
  d$abs_ssid     <- abs(d$lpv_ssid)
  d$delta_log10p <- abs(d$abs_ofav - d$abs_ssid)
  d$p_ofav       <- 10^(-d$abs_ofav)
  d$p_ssid       <- 10^(-d$abs_ssid)
  d$p_ratio      <- pmax(d$p_ofav, d$p_ssid) / pmin(d$p_ofav, d$p_ssid)
  d
}

rbind_align <- function(...) {
  dfs  <- list(...)
  cols <- unique(unlist(lapply(dfs, names)))
  dfs2 <- lapply(dfs, function(x) {
    if (is.null(x) || !nrow(x)) {
      return(as.data.frame(setNames(replicate(length(cols), logical(0), simplify = FALSE), cols)))
    }
    miss <- setdiff(cols, names(x))
    for (m in miss) x[[m]] <- NA
    x[, cols, drop = FALSE]
  })
  do.call(rbind, dfs2)
}

build_panels <- function(comparisons, ncol = 3) {
  global_lim <- compute_global_limits(comparisons)
  panels <- mapply(
    FUN      = function(title, df) make_panel(df, title, global_lim),
    title    = names(comparisons),
    df       = comparisons,
    SIMPLIFY = FALSE
  )
  nrow_grid <- ceiling(length(panels) / ncol)
  for (i in seq_along(panels)) {
    row <- ceiling(i / ncol)
    col <- i - (row - 1) * ncol
    if (col != 1) panels[[i]] <- panels[[i]] + theme(axis.title.y = element_blank())
    if (row != nrow_grid) panels[[i]] <- panels[[i]] + theme(axis.title.x = element_blank())
  }
  one_legend  <- cowplot::get_legend(panels[[1]] + theme(legend.position = "right"))
  legend_plot <- cowplot::ggdraw(one_legend)
  panels_noleg <- lapply(panels, function(p) p + theme(legend.position = "none"))
  grid  <- wrap_plots(panels_noleg, ncol = ncol)
  panel <- grid | legend_plot
  panel + plot_layout(widths = c(1, 0.30))
}

filter_discordant <- function(comparisons, group_col) {
  filter_one <- function(df, group_name) {
    d         <- classify_for_filters(df)
    meets_min <- pmax(d$abs_ofav, d$abs_ssid) >= min_sig
    
    inverse <- d[d$rel_class == "Inverse" & meets_min, , drop = FALSE]
    if (nrow(inverse)) {
      inverse$filter_type           <- "Inverse"
      inverse$relationship_category <- "Inverse"
      inverse$discord_direction     <- NA_character_
      inverse$delta_cut_used        <- NA_real_
      inverse$high_cut_used         <- NA_real_
      inverse$low_cut_used          <- NA_real_
      inverse$ratio_cut_used        <- NA_real_
      inverse[[group_col]]          <- group_name
    }
    
    strong_ofav_weak_ssid <- d$abs_ofav >= high_cut_dd & d$abs_ssid <= low_cut_dd
    strong_ssid_weak_ofav <- d$abs_ssid >= high_cut_dd & d$abs_ofav <= low_cut_dd
    cross_tier <- strong_ofav_weak_ssid | strong_ssid_weak_ofav
    dd_idx <- d$rel_class == "Direct" &
      d$delta_log10p >= delta_cut_dd &
      d$p_ratio      >= ratio_cut_dd &
      cross_tier & meets_min
    
    direct_discordant <- d[dd_idx, , drop = FALSE]
    if (nrow(direct_discordant)) {
      direct_discordant$filter_type           <- "DirectDiscordant"
      direct_discordant$relationship_category <- "Direct (discordant significance)"
      so <- direct_discordant$abs_ofav >= high_cut_dd & direct_discordant$abs_ssid <= low_cut_dd
      direct_discordant$discord_direction <- ifelse(so, "Ofav >> Ssid", "Ssid >> Ofav")
      direct_discordant$delta_cut_used    <- delta_cut_dd
      direct_discordant$high_cut_used     <- high_cut_dd
      direct_discordant$low_cut_used      <- low_cut_dd
      direct_discordant$ratio_cut_used    <- ratio_cut_dd
      direct_discordant[[group_col]]      <- group_name
    }
    
    rbind_align(inverse, direct_discordant)
  }
  
  do.call(rbind_align, lapply(names(comparisons), function(nm) filter_one(comparisons[[nm]], nm)))
}
```


# CORRELATION PLOTS TREATMENT

```{r correlation-plots-treatment}
treatment_comparisons <- list(
  "OA"                          = LC_CC,
  "Bleaching"                   = CH_CC,
  "OA + Bleaching"              = LH_CC,
  "Bleaching vs OA"             = CH_LC,
  "OA + Bleaching vs OA"        = LH_LC,
  "OA + Bleaching vs Bleaching" = LH_CH
)

panel_treatment <- build_panels(treatment_comparisons)
print(panel_treatment)
ggsave("../../outputs/transcriptomics/orthofinder/orthogroup correlation treatment.pdf",
       panel_treatment, width = 14, height = 9)
```


# CORRELATION DATAFRAME TREATMENT

```{r correlation-dataframe-treatment}
treatment_filtered_genes <- filter_discordant(treatment_comparisons, group_col = "treatment")

if (nrow(treatment_filtered_genes) > 0) {
  cat("Rows total:", nrow(treatment_filtered_genes), "\n")
  print(table(treatment_filtered_genes$relationship_category, useNA = "ifany"))
  print(table(treatment_filtered_genes$treatment, useNA = "ifany"))
} else {
  warning("treatment_filtered_genes is empty with current strict thresholds.")
}
dplyr::count(treatment_filtered_genes, treatment, relationship_category)
write.csv(treatment_filtered_genes,
          file = "../../outputs/transcriptomics/orthofinder/orthogroup correlation discordant treatment.csv")
```


# CORRELATION PLOTS URBAN vs REEF

```{r correlation-plots-site}
site_comparisons <- list(
  "Urban vs Reef" = urban_reef
)

panel_site <- build_panels(site_comparisons, ncol = 1)
print(panel_site)
ggsave("../../outputs/transcriptomics/orthofinder/orthogroup correlation site.pdf",
       panel_site, width = 6.5, height = 4.5)
```


# CORRELATION DATAFRAME URBAN vs REEF

```{r correlation-dataframe-site}
site_filtered_genes <- filter_discordant(site_comparisons, group_col = "site")

if (nrow(site_filtered_genes) > 0) {
  cat("Rows total:", nrow(site_filtered_genes), "\n")
  print(table(site_filtered_genes$relationship_category, useNA = "ifany"))
  print(table(site_filtered_genes$site, useNA = "ifany"))
} else {
  warning("site_filtered_genes is empty with current strict thresholds.")
}
dplyr::count(site_filtered_genes, site, relationship_category)
write.csv(site_filtered_genes,
          file = "../../outputs/transcriptomics/orthofinder/orthogroup correlation discordant site.csv")
```


# URBAN vs REEF ACROSS ALL ORTHOLOGS

```{r urban-vs-reef-all-orthologs}
# The DEG-matching sections above only keep orthologs that are significant in BOTH species, which leaves very
# few urban-vs-reef orthologs. Here the full response is compared across species instead: every one-to-one
# ortholog pair is used, whether or not it is significant, and the test statistics are correlated.
# One-to-one orthologs (one gene per species) avoid counting the same gene several times.
# Treatment contrasts are included as a benchmark for how similar the two species' responses are.
library(DESeq2)
library(tidyverse)

one2one <- orthologs %>%
  filter(!grepl(",", Ofaveolata_out_PRO), !grepl(",", Ssiderea_out_PRO)) %>%
  transmute(Orthogroup, Protein_ofav = trimws(Ofaveolata_out_PRO), Protein_ssid = trimws(Ssiderea_out_PRO))
nrow(one2one) # number of one-to-one ortholog pairs

# test statistics for each species: limma t for urban vs reef, DESeq2 Wald statistic for treatments
get_stats <- function(sp) {
  e <- new.env()
  load(sprintf("../../outputs/transcriptomics/%s/deseq2/urban_reef.RData", sp), envir = e)
  load(sprintf("../../outputs/transcriptomics/%s/deseq2/pvals.RData", sp), envir = e)
  out <- list(urban_reef = data.frame(gene = rownames(e$urban_reef), stat = e$urban_reef$t,
                                      padj = e$urban_reef$adj.P.Val))
  for (n in c("LC_CC", "CH_CC", "LH_CC", "CH_LC", "LH_LC", "LH_CH")) {
    r <- e[[n]]; out[[n]] <- data.frame(gene = rownames(r), stat = r$stat, padj = r$padj)
  }
  out
}
stats_ofav <- get_stats("ofav")
stats_ssid <- get_stats("ssid")

# input check: do the one-to-one ortholog protein IDs match the gene IDs used in each species' DE results?
# A low match rate usually means an ID format mismatch (e.g. isogroup vs isotig/protein names) or the wrong ortholog table.
id_match <- c(
  Ofaveolata = mean(one2one$Protein_ofav %in% unique(unlist(lapply(stats_ofav, `[[`, "gene")))),
  Ssiderea   = mean(one2one$Protein_ssid %in% unique(unlist(lapply(stats_ssid, `[[`, "gene"))))
)
cat("Fraction of one-to-one ortholog IDs found in DE gene IDs:\n"); print(round(id_match, 3))
if (any(id_match < 0.9))
  warning(sprintf("Only %s of one-to-one ortholog IDs found in DE results (%s); check that ortholog and DE gene IDs match.",
                  paste(sprintf("%.1f%%", 100 * id_match), collapse = " / "), paste(names(id_match), collapse = " / ")))

comparison_labels <- c(urban_reef = "Urban vs Reef", LC_CC = "OA", CH_CC = "Bleaching", LH_CC = "OA + Bleaching",
                       CH_LC = "Bleaching vs OA", LH_LC = "OA + Bleaching vs OA", LH_CH = "OA + Bleaching vs Bleaching")

ortholog_stats <- lapply(names(comparison_labels), function(n) {
  one2one %>%
    inner_join(stats_ofav[[n]], by = c("Protein_ofav" = "gene")) %>%
    inner_join(stats_ssid[[n]], by = c("Protein_ssid" = "gene"), suffix = c("_ofav", "_ssid")) %>%
    filter(is.finite(stat_ofav), is.finite(stat_ssid)) %>%
    mutate(comparison = n,
           sig = case_when(!is.na(padj_ofav) & padj_ofav < 0.1 & !is.na(padj_ssid) & padj_ssid < 0.1 ~ "both species",
                           !is.na(padj_ofav) & padj_ofav < 0.1 ~ "O. faveolata only",
                           !is.na(padj_ssid) & padj_ssid < 0.1 ~ "S. siderea only",
                           TRUE ~ "neither"))
})
names(ortholog_stats) <- names(comparison_labels)

# Spearman correlation across all one-to-one orthologs, plus direction agreement among orthologs
# significant (padj < 0.1) in at least one species. Genes are not independent, so treat p-values as approximate.
ortholog_summary <- do.call(rbind, lapply(names(ortholog_stats), function(n) {
  d <- ortholog_stats[[n]]
  ct <- suppressWarnings(cor.test(d$stat_ofav, d$stat_ssid, method = "spearman"))
  s <- d[d$sig != "neither", ]
  agree <- sum(sign(s$stat_ofav) == sign(s$stat_ssid))
  data.frame(comparison = comparison_labels[[n]], orthologs = nrow(d),
             spearman_rho = round(unname(ct$estimate), 3), p = signif(ct$p.value, 2),
             sig_either = nrow(s), sig_both = sum(d$sig == "both species"),
             same_direction = agree,
             pct_same_direction = round(100 * agree / max(nrow(s), 1), 1),
             p_direction = if (nrow(s)) signif(binom.test(agree, nrow(s))$p.value, 2) else NA)
}))
ortholog_summary
write.csv(ortholog_summary, file = "../../outputs/transcriptomics/orthofinder/ortholog correlation all 1to1 summary.csv", row.names = FALSE)

# urban vs reef ortholog table with annotations
ortholog_stats$urban_reef %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t", quote = "", fill = FALSE) %>%
              dplyr::select(gene = V1, annot_ofav = V2), by = c("Protein_ofav" = "gene")) %>%
  left_join(read.table(file = "../../../Siderastrea-siderea-annotated-transcriptome/Ssiderea_iso2geneName.tab",
                       sep = "\t", quote = "", fill = FALSE) %>%
              dplyr::select(gene = V1, annot_ssid = V2), by = c("Protein_ssid" = "gene")) %>%
  arrange(sig, -abs(stat_ofav) - abs(stat_ssid)) -> urban_reef_orthologs
write.csv(urban_reef_orthologs, file = "../../outputs/transcriptomics/orthofinder/ortholog correlation urban_reef 1to1.csv", row.names = FALSE)

# scatterplot of urban vs reef statistics across all one-to-one orthologs
ur <- ortholog_stats$urban_reef %>%
  mutate(sig = factor(sig, levels = c("neither", "O. faveolata only", "S. siderea only", "both species"))) %>%
  arrange(sig)
ur_rho <- ortholog_summary$spearman_rho[ortholog_summary$comparison == "Urban vs Reef"]
p_ur <- ggplot(ur, aes(x = stat_ofav, y = stat_ssid)) +
  geom_hline(yintercept = 0, color = "grey85") +
  geom_vline(xintercept = 0, color = "grey85") +
  geom_point(aes(color = sig, alpha = sig), size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  scale_color_manual(values = c("neither" = "grey70", "O. faveolata only" = "#E69F00",
                                "S. siderea only" = "#56B4E9", "both species" = "#D55E00"), name = "padj < 0.1") +
  scale_alpha_manual(values = c("neither" = 0.25, "O. faveolata only" = 0.8, "S. siderea only" = 0.8, "both species" = 1),
                     guide = "none") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = sprintf("Spearman rho = %.2f, n = %d orthologs", ur_rho, nrow(ur))) +
  labs(x = "O. faveolata urban vs reef (t)", y = "S. siderea urban vs reef (t)",
       title = "Urban vs reef across one-to-one orthologs") +
  theme_classic(base_size = 12)

# species agreement for each comparison (benchmark for the urban vs reef correlation)
p_rho <- ortholog_summary %>%
  mutate(comparison = factor(comparison, levels = rev(comparison))) %>%
  ggplot(aes(x = spearman_rho, y = comparison)) +
  geom_vline(xintercept = 0, color = "grey70") +
  geom_segment(aes(x = 0, xend = spearman_rho, yend = comparison), color = "grey50") +
  geom_point(size = 3) +
  labs(x = "Spearman rho (O. faveolata vs S. siderea)", y = NULL,
       title = "Cross-species agreement") +
  theme_classic(base_size = 12)

p_orth_all <- p_ur + p_rho + plot_layout(widths = c(1.3, 1))
print(p_orth_all)
ggsave("../../outputs/transcriptomics/orthofinder/ortholog correlation all 1to1.pdf", p_orth_all, width = 13, height = 5.5)
```


# GENE BOXPLOTS

```{r gene-boxplots}
# ---- (tiny preamble) ----
while (dev.cur() > 1) try(dev.off(), silent = TRUE)
options(device = "pdf")

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggpubr)
  library(DESeq2)
  library(SummarizedExperiment)
  library(readr)
  library(cowplot)
  library(grid)
  library(scales)
  library(rstatix)
  library(stringr)
  library(rlang)
})
`%||%` <- rlang::`%||%`

# --- Ensure dplyr functions override Bioconductor ones ---
for (fn in c("filter","select","mutate","arrange","summarize","slice"))
  assign(fn, get(fn, envir = asNamespace("dplyr")), envir = globalenv())

set.seed(42)
options(ggplot2.useDingbats = FALSE)

# ---- Helpers ----
.safe <- function(x) gsub("[^[:alnum:]_\\-\\.]+", "_", x)

# ---- Top-height calculator for brackets ----
.bracket_required_top <- function(df, tukey_tbl, lvls,
                                  pad_dex   = 0.15,
                                  gap_dex   = 0.15,
                                  lbl_dex   = 0.005,
                                  min_sep_dex = 0.08,
                                  micro_dex = 0.020) {
  if (is.null(tukey_tbl) || !nrow(tukey_tbl)) return(NA_real_)
  tuk <- tukey_tbl
  tuk$group1 <- as.character(tuk$group1)
  tuk$group2 <- as.character(tuk$group2)
  tuk <- tuk[tuk$group1 %in% lvls & tuk$group2 %in% lvls, , drop=FALSE]
  if (!nrow(tuk)) return(NA_real_)
  if ("p.adj.signif" %in% names(tuk)) tuk <- dplyr::filter(tuk, !is.na(p.adj.signif) & tolower(p.adj.signif) != "ns")
  else if ("p.adj" %in% names(tuk))   tuk <- dplyr::filter(tuk, !is.na(p.adj) & p.adj < 0.05)
  else if ("p" %in% names(tuk))       tuk <- dplyr::filter(tuk, !is.na(p) & p < 0.05)
  else return(NA_real_)
  if (!nrow(tuk)) return(NA_real_)
  
  x1 <- match(tuk$group1, lvls); x2 <- match(tuk$group2, lvls)
  span <- abs(x2 - x1)
  tuk  <- tuk[order(span), , drop=FALSE]
  
  overlaps <- function(a1,a2,b1,b2) !(a2 < b1 || b2 < a1)
  lanes <- list(); lane_id <- integer(nrow(tuk))
  for (i in seq_len(nrow(tuk))) {
    a1 <- min(x1[i], x2[i]); a2 <- max(x1[i], x2[i]); placed <- FALSE
    for (L in seq_along(lanes)) {
      lane_ok <- TRUE
      for (j in seq_len(nrow(lanes[[L]]))) if (overlaps(a1,a2, lanes[[L]]$a1[j], lanes[[L]]$a2[j])) { lane_ok <- FALSE; break }
      if (lane_ok) { lanes[[L]] <- rbind(lanes[[L]], data.frame(a1=a1,a2=a2)); lane_id[i] <- L; placed <- TRUE; break }
    }
    if (!placed) { lanes[[length(lanes)+1]] <- data.frame(a1=a1,a2=a2); lane_id[i] <- length(lanes) }
  }
  
  df_pos <- df %>% dplyr::filter(is.finite(count), count > 0)
  group_max <- tapply(df_pos$count, factor(df_pos$grp, levels=lvls), max, na.rm=TRUE)
  group_max[is.infinite(group_max) | is.na(group_max)] <- 1
  base_h   <- pmax(group_max[tuk$group1], group_max[tuk$group2])
  base_log <- log10(pmax(base_h, 1))
  
  within_lane_idx <- ave(seq_along(lane_id), lane_id, FUN=function(ix) rank(ix, ties.method="first"))
  y_log   <- base_log + pad_dex + (lane_id - 1)*gap_dex + (within_lane_idx - 1)*pmax(micro_dex, min_sep_dex)
  lbl_log <- y_log + lbl_dex
  10^(max(lbl_log, na.rm=TRUE) + 0.010)
}

# ---- Draw significant Tukey brackets ----
.add_sig_brackets <- function(p, tukey_tbl, lvls, top_y, df,
                              pad_dex=0.2, gap_dex=0.15, tip_dex=0.025,
                              lbl_dex=0.005, cap_dex=0.010, min_clr=0.06, min_sep_dex=0.08){
  if (is.null(tukey_tbl) || !nrow(tukey_tbl)) return(p)
  tuk <- tukey_tbl
  tuk$group1 <- as.character(tuk$group1); tuk$group2 <- as.character(tuk$group2)
  tuk <- tuk[tuk$group1 %in% lvls & tuk$group2 %in% lvls, , drop = FALSE]
  if (!nrow(tuk)) return(p)
  if ("p.adj.signif" %in% names(tuk)) tuk <- dplyr::filter(tuk, !is.na(p.adj.signif) & tolower(p.adj.signif) != "ns")
  else if ("p.adj" %in% names(tuk))   tuk <- dplyr::filter(tuk, !is.na(p.adj) & p.adj < 0.05)
  else if ("p" %in% names(tuk))       tuk <- dplyr::filter(tuk, !is.na(p) & p < 0.05) else return(p)
  if (!nrow(tuk)) return(p)
  
  x1 <- match(tuk$group1, lvls); x2 <- match(tuk$group2, lvls)
  span <- abs(x2 - x1)
  p_val <- if ("p.adj" %in% names(tuk)) tuk$p.adj else if ("p" %in% names(tuk)) tuk$p else NA_real_
  ord  <- order(span, dplyr::coalesce(p_val, Inf))
  tuk  <- tuk[ord, , drop = FALSE]; x1 <- x1[ord]; x2 <- x2[ord]
  
  overlaps <- function(a1, a2, b1, b2) !(a2 < b1 || b2 < a1)
  lanes <- list(); lane_id <- integer(nrow(tuk))
  for (i in seq_len(nrow(tuk))) {
    a1 <- min(x1[i], x2[i]); a2 <- max(x1[i], x2[i]); placed <- FALSE
    for (L in seq_along(lanes)) {
      lane_ok <- TRUE
      for (j in seq_len(nrow(lanes[[L]]))) if (overlaps(a1, a2, lanes[[L]]$a1[j], lanes[[L]]$a2[j])) { lane_ok <- FALSE; break }
      if (lane_ok) { lanes[[L]] <- rbind(lanes[[L]], data.frame(a1=a1,a2=a2)); lane_id[i] <- L; placed <- TRUE; break }
    }
    if (!placed) { lanes[[length(lanes)+1]] <- data.frame(a1=a1,a2=a2); lane_id[i] <- length(lanes) }
  }
  
  df_pos <- df %>% dplyr::filter(is.finite(count), count > 0)
  group_max <- tapply(df_pos$count, factor(df_pos$grp, levels = lvls), max, na.rm = TRUE)
  group_max[is.infinite(group_max) | is.na(group_max)] <- 1
  base_h   <- pmax(group_max[tuk$group1], group_max[tuk$group2])
  base_log <- log10(pmax(base_h, 1)); top_log  <- log10(pmax(top_y, 1))
  y_log <- base_log + pmax(pad_dex, min_clr) + (lane_id - 1) * pmax(gap_dex, min_clr)
  cap <- top_log - cap_dex
  
  for (L in unique(lane_id)) {
    idx <- which(lane_id == L); if (length(idx) <= 1) next
    ordL <- idx[order(y_log[idx], na.last = NA)]
    for (k in seq_along(ordL)) {
      i <- ordL[k]
      min_i <- base_log[i] + min_clr + (L - 1) * pmax(gap_dex, min_clr)
      y_log[i] <- max(y_log[i], min_i)
      if (k > 1) y_log[i] <- max(y_log[i], y_log[ordL[k-1]] + min_sep_dex)
    }
    lane_lbl_top <- max(y_log[ordL] + lbl_dex, na.rm = TRUE)
    if (lane_lbl_top > cap) {
      shift <- lane_lbl_top - cap
      y_log[ordL] <- y_log[ordL] - shift
      for (k in seq_along(ordL)) {
        i <- ordL[k]
        min_i <- base_log[i] + min_clr + (L - 1) * pmax(gap_dex, min_clr)
        if (y_log[i] < min_i) y_log[i] <- min_i
        if (k > 1) y_log[i] <- max(y_log[i], y_log[ordL[k-1]] + min_sep_dex)
      }
    }
  }
  
  tip_log <- y_log - tip_dex; lbl_log <- y_log + lbl_dex
  y <- 10^y_log; y0 <- 10^tip_log; ylbl <- 10^lbl_log
  xm <- (pmin(x1, x2) + pmax(x1, x2)) / 2
  lab <- if ("p.adj.signif" %in% names(tuk)) as.character(tuk$p.adj.signif)
  else if ("p.adj" %in% names(tuk)) sprintf("p=%.3g", tuk$p.adj)
  else sprintf("p=%.3g", tuk$p)
  
  seg_df <- data.frame(x1=pmin(x1,x2), x2=pmax(x1,x2), y=y, y0=y0, xm=xm, lab=lab, ylbl=ylbl)
  p +
    geom_segment(data=seg_df, aes(x=x1, xend=x2, y=y, yend=y), linewidth=0.5, inherit.aes=FALSE) +
    geom_segment(data=seg_df, aes(x=x1, xend=x1, y=y0, yend=y), linewidth=0.5, inherit.aes=FALSE) +
    geom_segment(data=seg_df, aes(x=x2, xend=x2, y=y0, yend=y), linewidth=0.5, inherit.aes=FALSE) +
    annotate("text", x=seg_df$xm, y=seg_df$ylbl, label=seg_df$lab, size=3, vjust=0, hjust=0.5)
}

# ---- Crash-guard helpers ----
.as_grob_safe <- function(p) {
  tryCatch(cowplot::as_grob(p),
           error = function(e) cowplot::as_grob(ggplot() + theme_void()))
}
.is_panel_ok <- function(p) inherits(p, c("gg", "grob", "gtable"))

safe_ggsave_pdf <- function(plot, filename, width, height, limitsize = TRUE) {
  tmp <- tempfile(fileext = ".pdf")
  ok <- FALSE
  try({
    ggsave(tmp, plot = plot, width = width, height = height, units = "in",
           dpi = 300, device = grDevices::pdf, useDingbats = FALSE,
           limitsize = limitsize)
    ok <- TRUE
  }, silent = TRUE)
  if (ok) {
    file.copy(tmp, filename, overwrite = TRUE)
  } else {
    message("⚠️ ggsave failed for: ", filename)
  }
  unlink(tmp)
  invisible(ok)
}
```


## 0) Load DESeq2 objects

```{r 0-load-deseq2-objects}
load("../../outputs/transcriptomics/ofav/deseq2/realModels.RData"); dds_ofav <- dds
load("../../outputs/transcriptomics/ssid/deseq2/realModels.RData"); dds_ssid <- dds
message("ofav colData: ", paste(colnames(SummarizedExperiment::colData(dds_ofav)), collapse=", "))
message("ssid colData: ", paste(colnames(SummarizedExperiment::colData(dds_ssid)), collapse=", "))
```


## 1) Read lookup and normalize labels

```{r 1-read-lookup-and-normalize-labels}
# The boxplot input is rebuilt from the two discordance tables written above (CORRELATION DATAFRAME sections):
# inverse orthologs + direct orthologs with discordant significance, for treatment and urban vs reef.
# Stress family / short gene name come from a small curated lookup keyed by orthogroup.
orthofinder_out <- "../../outputs/transcriptomics/orthofinder"
lookup_file     <- "../../data/transcriptomics/orthofinder/gene boxplot lookup.csv"

read_discordant <- function(file, factor_label, group_col) {
  if (!file.exists(file)) {
    warning("Discordance table not found (run the CORRELATION DATAFRAME sections first): ", file)
    return(NULL)
  }
  d <- tryCatch(read.csv(file, header = TRUE, check.names = FALSE, stringsAsFactors = FALSE),
                error = function(e) data.frame())
  if (!nrow(d)) return(NULL)
  if (!group_col %in% names(d)) d[[group_col]] <- NA_character_
  data.frame(
    factor                = factor_label,
    comparison            = as.character(d[[group_col]]),
    orthogroup_ID         = as.character(d$Orthogroup),
    Ofaveolata_ID         = trimws(as.character(d$Protein_ofav)),
    Ssiderea_ID           = trimws(as.character(d$Protein_ssid)),
    Ofaveolata_lpv        = as.numeric(d$lpv_ofav),
    Ssiderea_lpv          = as.numeric(d$lpv_ssid),
    relationship          = as.character(d$rel_class),
    Ofaveolata_annotation = as.character(d$annot_ofav),
    Ssiderea_annotation   = as.character(d$annot_ssid),
    stringsAsFactors = FALSE
  )
}

df <- dplyr::bind_rows(
  read_discordant(file.path(orthofinder_out, "orthogroup correlation discordant treatment.csv"), "Treatment", "treatment"),
  read_discordant(file.path(orthofinder_out, "orthogroup correlation discordant site.csv"),      "Site",      "site")
)
if (!nrow(df)) stop("No discordant orthologs found; nothing to plot in the gene boxplot section.")

# Orthogroup IDs were made unique within each comparison (e.g. OG0001234_1 for a second protein pair);
# the lookup is keyed by the base orthogroup ID
df$og_base <- sub("_[0-9]+$", "", df$orthogroup_ID)

if (!file.exists(lookup_file)) {
  template <- df %>%
    dplyr::group_by(orthogroup_ID = og_base) %>%
    dplyr::summarise(stress_family = "", gene_name = "",
                     Ofaveolata_annotation = dplyr::first(Ofaveolata_annotation),
                     Ssiderea_annotation   = dplyr::first(Ssiderea_annotation), .groups = "drop") %>%
    dplyr::arrange(orthogroup_ID)
  dir.create(dirname(lookup_file), recursive = TRUE, showWarnings = FALSE)
  write.csv(template, lookup_file, row.names = FALSE)
  message("Wrote a gene boxplot lookup template with ", nrow(template), " orthogroups to '", lookup_file,
          "'. Fill in stress_family (one of: Apoptosis/cell death, DNA repair, Heat shock proteins, Immune response, ",
          "Oxidative stress/antioxidants, Ubiquitin-proteasome, Other stress-related) and gene_name, then re-run. ",
          "Only orthogroups with a stress_family are plotted.")
}
gene_lookup <- read.csv(lookup_file, header = TRUE, check.names = FALSE, colClasses = "character",
                        na.strings = c("", "NA")) %>%
  dplyr::transmute(og_base = sub("_[0-9]+$", "", trimws(orthogroup_ID)),
                   stress_family = trimws(stress_family), gene_name = trimws(gene_name)) %>%
  dplyr::distinct(og_base, .keep_all = TRUE)

df <- df %>%
  dplyr::left_join(gene_lookup, by = "og_base") %>%
  dplyr::mutate(
    # empty stress family -> 'unassigned' (dropped below); empty gene name -> O. faveolata annotation (shortened), else orthogroup
    stress_family = dplyr::coalesce(dplyr::na_if(stress_family, ""), "unassigned"),
    gene_name     = dplyr::coalesce(dplyr::na_if(gene_name, ""),
                                    stringr::str_trunc(dplyr::na_if(Ofaveolata_annotation, ""), 40),
                                    og_base)
  ) %>%
  dplyr::select(-og_base)
cat("Boxplot orthogroups without a curated stress family (not plotted):", sum(df$stress_family == "unassigned"), "of", nrow(df), "rows\n")
# only orthogroups curated into a stress family in the lookup are plotted (several hundred discordant orthologs otherwise)
df <- df %>% dplyr::filter(stress_family != "unassigned")
if (!nrow(df)) stop("No curated stress-family orthologs among the discordant orthologs; fill in the gene boxplot lookup.")

# One boxplot panel per gene pair. An orthogroup ID does not identify a gene pair: an orthogroup can hold several
# pairs (the ortholog table is not restricted to one-to-one orthologs), and the _1, _2 suffixes are assigned within
# each comparison, so the same ID can refer to different pairs in different comparisons. From here on `orthogroup_ID`
# is therefore a pair label: orthogroup, O. faveolata gene and S. siderea gene (shown in the panel titles and file names).
df$orthogroup_ID <- paste0(sub("_[0-9]+$", "", df$orthogroup_ID), ": ", df$Ofaveolata_ID, " / ", df$Ssiderea_ID)

# One panel per orthogroup: where an orthogroup has several gene pairs in a subset (factor x relationship), only the
# pair with the strongest combined signal is plotted (largest |lpv O. faveolata| + |lpv S. siderea| in any comparison).
# All pairs remain in the discordance tables.
df <- df %>%
  dplyr::mutate(og_group = sub(":.*$", "", orthogroup_ID), signal = abs(Ofaveolata_lpv) + abs(Ssiderea_lpv)) %>%
  dplyr::group_by(factor, relationship, og_group) %>%
  dplyr::mutate(best_pair = orthogroup_ID[which.max(signal)]) %>%
  dplyr::ungroup()
cat("Gene pairs not plotted because another pair of the same orthogroup is shown:",
    dplyr::n_distinct(paste(df$relationship, df$orthogroup_ID)[df$orthogroup_ID != df$best_pair]), "\n")
df <- df %>% dplyr::filter(orthogroup_ID == best_pair) %>% dplyr::select(-og_group, -signal, -best_pair)

needed <- c("factor","comparison","stress_family","orthogroup_ID",
            "Ofaveolata_ID","Ssiderea_ID","Ofaveolata_lpv","Ssiderea_lpv",
            "relationship","Ofaveolata_annotation","Ssiderea_annotation","gene_name")
stopifnot(all(needed %in% names(df)))

df_norm <- df %>%
  mutate(
    factor = case_when(tolower(factor)=="site" ~ "Site",
                       tolower(factor)=="treatment" ~ "Treatment",
                       TRUE ~ factor),
    relationship = case_when(tolower(relationship)=="direct" ~ "Direct",
                             tolower(relationship)=="inverse" ~ "Inverse",
                             TRUE ~ relationship)
  ) %>%
  mutate(
    Ofaveolata_ID_safe = .safe(Ofaveolata_ID),
    Ssiderea_ID_safe   = .safe(Ssiderea_ID)
  )
```


## 2) Subsets

```{r 2-subsets}
site_direct       <- df_norm %>% filter(factor=="Site",      relationship=="Direct")  %>% distinct(orthogroup_ID, .keep_all = TRUE)
site_inverse      <- df_norm %>% filter(factor=="Site",      relationship=="Inverse") %>% distinct(orthogroup_ID, .keep_all = TRUE)
treatment_direct  <- df_norm %>% filter(factor=="Treatment", relationship=="Direct")  %>% distinct(orthogroup_ID, .keep_all = TRUE)
treatment_inverse <- df_norm %>% filter(factor=="Treatment", relationship=="Inverse") %>% distinct(orthogroup_ID, .keep_all = TRUE)

cat("\nGene pairs per subset:\n",
    "Site-Direct:   ", nrow(site_direct), "\n",
    "Site-Inverse:  ", nrow(site_inverse), "\n",
    "Treat-Direct:  ", nrow(treatment_direct), "\n",
    "Treat-Inverse: ", nrow(treatment_inverse), "\n", sep = "")
```


## 3) Export counts CSVs

```{r 3-export-counts-csvs}
counts_root <- "../../outputs/transcriptomics/orthofinder/counts"
# clear previous exports so stale gene files from earlier runs are not read back in below
unlink(counts_root, recursive = TRUE)
dir.create(counts_root, recursive = TRUE, showWarnings = FALSE)
# subset/species folders are created when the first gene file is written, so empty subsets leave no empty folders

.export_counts_species <- function(dds, gene_id, intgroup, out_dir, species_tag){
  stopifnot(is.character(gene_id), length(gene_id)==1, nzchar(gene_id))
  cd <- SummarizedExperiment::colData(dds)
  if (!all(intgroup %in% colnames(cd)))
    stop(sprintf("intgroup '%s' not in colData(%s). Have: %s", intgroup, species_tag, paste(colnames(cd), collapse=", ")))
  if (!(gene_id %in% rownames(dds)))
    stop(sprintf("Gene '%s' not in rownames(%s).", gene_id, species_tag))
  dat <- DESeq2::plotCounts(dds, gene = gene_id, intgroup = intgroup, returnData = TRUE)
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  fn <- file.path(out_dir, paste0(.safe(gene_id), ".csv"))
  write.csv(dat, fn, row.names = TRUE)
  message(sprintf("[%s] Wrote: %s (n=%d, cols: %s)", species_tag, normalizePath(fn, mustWork = FALSE), nrow(dat), paste(names(dat), collapse=", ")))
  invisible(fn)
}

export_counts_for_orthogroup <- function(df_sub, og_id, intgroup, out_dir){
  row <- df_sub %>% filter(orthogroup_ID == og_id) %>% slice(1)
  if (nrow(row)==0) stop("OG not found in subset: ", og_id)
  ofav_id <- row$Ofaveolata_ID; ssid_id <- row$Ssiderea_ID
  .export_counts_species(dds_ofav, ofav_id, intgroup, file.path(out_dir, "Ofaveolata"), "Ofaveolata")
  .export_counts_species(dds_ssid, ssid_id, intgroup, file.path(out_dir, "Ssiderea"),  "Ssiderea")
  invisible(NULL)
}

export_counts_batch <- function(df_subset, intgroup, out_dir) {
  if (!nrow(df_subset)) {
    message("Subset is empty; nothing to export.")
    return(invisible(list(success = 0, failed = 0, total = 0)))
  }
  ogs <- unique(df_subset$orthogroup_ID)
  total <- length(ogs)
  message(sprintf("Exporting %d orthogroups to '%s' (intgroup=%s)...",
                  total, normalizePath(out_dir, mustWork = FALSE), paste(intgroup, collapse = "+")))
  ok <- 0; fail <- 0
  for (og in ogs) {
    tryCatch({
      export_counts_for_orthogroup(df_subset, og, intgroup = intgroup, out_dir = out_dir)
      ok <- ok + 1
    }, error = function(e) {
      message(sprintf("❌ OG %s failed: %s", og, e$message))
      fail <<- fail + 1
    })
  }
  message(sprintf("✅ Export complete for '%s': %d succeeded, %d failed (of %d total).",
                  out_dir, ok, fail, total))
  invisible(list(success = ok, failed = fail, total = total))
}

# Site exports habitat + parent colony (genotype) so counts can be averaged per colony (urban vs reef);
# Treatment uses 'treat' (fragment level)
site_intgroup <- c("habitat", "genotype")
res_counts <- list(
  Site_Direct       = export_counts_batch(site_direct,       intgroup = site_intgroup, out_dir = file.path(counts_root,"Site_Direct")),
  Site_Inverse      = export_counts_batch(site_inverse,      intgroup = site_intgroup, out_dir = file.path(counts_root,"Site_Inverse")),
  Treatment_Direct  = export_counts_batch(treatment_direct,  intgroup = "treat", out_dir = file.path(counts_root,"Treatment_Direct")),
  Treatment_Inverse = export_counts_batch(treatment_inverse, intgroup = "treat", out_dir = file.path(counts_root,"Treatment_Inverse"))
)
```


## 4) Read count CSVs back to long data

```{r 4-read-count-csvs-back-to-long-data}
csv_files <- list.files(counts_root, pattern="\\.csv$", full.names = TRUE, recursive = TRUE)

read_gene_file <- function(file_path){
  # read grouping columns as character so Site (habitat/genotype) and Treatment (treat) files stack cleanly
  dfc <- suppressMessages(readr::read_csv(file_path, show_col_types = FALSE,
                                          col_types = readr::cols(.default = readr::col_character(),
                                                                  count = readr::col_double())))
  species <- basename(dirname(file_path))
  subset  <- basename(dirname(dirname(file_path)))
  subset_bits <- str_split(subset, "_", simplify = TRUE)
  factor_name <- subset_bits[,1, drop=TRUE]
  relationship <- subset_bits[,2, drop=TRUE]
  gene_id <- tools::file_path_sans_ext(basename(file_path))
  id_col      <- if (species == "Ofaveolata") "Ofaveolata_ID" else "Ssiderea_ID"
  id_col_safe <- paste0(id_col, "_safe")
  
  # a gene can belong to more than one gene pair (e.g. one O. faveolata gene matched to two S. siderea genes),
  # so its counts are returned once for every pair it belongs to in this subset
  meta_rows <- df_norm %>%
    filter(factor == factor_name,
           relationship == .env$relationship,
           (!!rlang::sym(id_col) == gene_id) | (!!rlang::sym(id_col_safe) == gene_id)) %>%
    distinct(orthogroup_ID, .keep_all = TRUE)
  if (!nrow(meta_rows)) {
    meta_rows <- df_norm %>%
      filter((!!rlang::sym(id_col) == gene_id) | (!!rlang::sym(id_col_safe) == gene_id)) %>% slice(1)
  }
  if (!nrow(meta_rows)) # gene not in the boxplot table: keep one row with missing labels
    meta_rows <- tibble(orthogroup_ID = NA_character_, gene_name = NA_character_, stress_family = NA_character_)
  
  tibble(
    species       = species,
    gene_id       = gene_id,
    orthogroup_ID = meta_rows$orthogroup_ID,
    gene_name     = meta_rows$gene_name,
    stress_family = meta_rows$stress_family,
    relationship  = relationship,
    factor        = factor_name,
    subset        = paste(factor_name, relationship, sep = "_"),
    count_data    = rep(list(dfc), nrow(meta_rows))
  )
}

counts_list <- purrr::map_dfr(csv_files, read_gene_file)
counts_long <- counts_list %>% tidyr::unnest(count_data)
```


## 5) Plotting + stats

```{r 5-plotting-stats}
species_titles <- list(
  Ofaveolata = expression(italic("O. faveolata")),
  Ssiderea   = expression(italic("S. siderea"))
)
# Site panels compare habitat (reef vs urban) using parent-colony means (colony = genotype),
# because fragments from the same colony are not independent and habitat is constant within a colony
habitat_levels <- c("reef","urban")
habitat_colors <- c("#018571","#a6611a")
treat_levels  <- c("CC","LC","CH","LH")
treat_colors  <- c("#92c5de","#0571b0","#f4a582","#ca0020")
stress_family_palette <- c(
  "Apoptosis/cell death"          = "#EF4444",
  "DNA repair"                    = "#10B981",
  "Heat shock proteins"           = "#8B5CF6",
  "Immune response"               = "#F59E0B",
  "Oxidative stress/antioxidants" = "#3B82F6",
  "Ubiquitin-proteasome"          = "#6366F1",
  "Other stress-related"          = "#6B7280"
)
# orthogroups not yet curated in the gene boxplot lookup
if (any(df_norm$stress_family == "unassigned", na.rm = TRUE))
  stress_family_palette <- c(stress_family_palette, "unassigned" = "#BDBDBD")
use_fixed_y_limits <- TRUE
# y limits computed from the plotted counts (all orthogroups; colony means fall within this range).
# Note: the panels below use per-orthogroup limits shared by both species; this global range is for reference.
.pos_counts <- counts_long$count[is.finite(counts_long$count) & counts_long$count > 0]
fixed_y_limits <- if (length(.pos_counts)) c(min(.pos_counts) / 1.3, max(.pos_counts) * 1.3) else c(0.1, 1)

habitat_labels <- c(
  "reef"  = "Reef\n(colony means)",
  "urban" = "Urban\n(colony means)"
)
treat_labels <- c(
  "CC" = "Contemporary\npH + Ambient\nTemperature",
  "LC" = "Acidified\n+ Ambient\nTemperature",
  "CH" = "Contemporary\npH +\nBleaching",
  "LH" = "Acidified\n+\nBleaching"
)

panel_with_colored_title <- function(left_plot, right_plot, title_text, stress_family){
  fill_col <- if (!is.null(stress_family) && stress_family %in% names(stress_family_palette)) {
    stress_family_palette[stress_family]
  } else "#6B7280"
  
  gL <- ggplotGrob(left_plot); gR <- ggplotGrob(right_plot)
  maxw <- grid::unit.pmax(gL$widths[2:5], gR$widths[2:5])
  gL$widths[2:5] <- maxw; gR$widths[2:5] <- maxw
  
  core <- cowplot::plot_grid(cowplot::as_grob(gL), cowplot::as_grob(gR), ncol = 2, rel_widths = c(1, 0.92))
  
  title_strip <- ggplot() +
    geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-Inf,ymax=Inf), fill=fill_col, color=NA) +
    annotate("text", x=0, y=0, label=title_text, color="white", fontface="bold",
             size=if (nchar(title_text) > 65) 4.2 * 65 / nchar(title_text) else 4.2, # long gene-pair titles are shrunk to fit the bar
             hjust=0.5, vjust=0.5) +
    coord_cartesian(xlim=c(-1,1), ylim=c(-1,1), expand=FALSE) + theme_void()
  
  cowplot::plot_grid(title_strip, core, ncol=1, rel_heights=c(0.12, 1))
}

compute_stats <- function(df_grp){
  df2 <- df_grp %>% dplyr::filter(!is.na(grp), is.finite(count))
  out <- list(aov_text = NULL, aov_table = NULL, tukey = NULL, aov_label = NULL, aov_p = NA_real_)
  if (!nrow(df2)) return(out)
  
  lvl_counts <- table(droplevels(df2$grp))
  if (sum(lvl_counts > 0) < 2L) return(out)
  
  aov_res <- tryCatch(aov(count ~ grp, data = df2), error = function(e) NULL)
  got_aov <- !is.null(aov_res)
  pv <- NA_real_; Fv <- NA_real_; df1 <- NA_integer_; df2res <- NA_integer_
  
  if (got_aov) {
    sm <- summary(aov_res)[[1]]
    out$aov_table <- as.data.frame(sm) %>% tibble::rownames_to_column("Term") %>%
      dplyr::mutate(across(where(is.numeric), ~ round(.x, 4)))
    term_names <- attr(terms(aov_res), "term.labels")
    term <- if (length(term_names)) term_names[1] else "grp"
    row_idx <- which(rownames(sm) == term)
    if (!length(row_idx)) row_idx <- which(grepl(paste0("^", term, "$"), rownames(sm)))
    if (length(row_idx)) {
      df1     <- sm[row_idx, "Df"]
      df2res  <- sm["Residuals", "Df"]
      Fv      <- sm[row_idx, "F value"]
      pv      <- sm[row_idx, "Pr(>F)"]
    }
  }
  
  if (!is.finite(pv)) {
    welch <- tryCatch(oneway.test(count ~ grp, data = df2, var.equal = FALSE), error = function(e) NULL)
    if (!is.null(welch)) {
      pv <- as.numeric(welch$p.value)
      df1 <- unname(signif(welch$parameter[1], 3))
      df2res <- unname(signif(welch$parameter[2], 3))
      Fv <- unname(signif(welch$statistic[[1]], 3))
    }
  }
  if (!is.finite(pv)) {
    kw <- tryCatch(kruskal.test(count ~ grp, data = df2), error = function(e) NULL)
    if (!is.null(kw)) {
      pv <- as.numeric(kw$p.value)
      Fv <- NA_real_; df1 <- NA; df2res <- NA
    }
  }
  
  out$aov_p <- pv
  
  # --- Plain text label (CSV/logs) with star when p<0.05 ---
  pv_str_plain <- if (is.finite(pv)) formatC(pv, format = "g", digits = 3) else "NA"
  star_plain   <- if (is.finite(pv) && pv < 0.05) "*" else ""
  if (is.finite(Fv) && is.finite(pv) && is.finite(as.numeric(df1)) && is.finite(as.numeric(df2res))) {
    out$aov_text <- sprintf("ANOVA: F(%s,%s)=%.2f, p=%s%s", df1, df2res, as.numeric(Fv), pv_str_plain, star_plain)
  } else if (is.finite(pv)) {
    out$aov_text <- sprintf("ANOVA: p=%s%s", pv_str_plain, star_plain)
  } else {
    out$aov_text <- "ANOVA: p=NA"
  }
  
  # --- Plotmath label with star when p<0.05 ---
  if (is.finite(pv)) {
    pretty_p <- if (pv < 1e-4) "'< 1e-4'"
    else if (pv < 0.001) "'< 0.001'"
    else formatC(pv, format = "g", digits = 3)
    star_plot <- if (pv < 0.05) ", '*'" else ""  # appended as another paste() arg
    
    Fv_str <- if (is.finite(Fv)) formatC(as.numeric(Fv), format = "f", digits = 2) else "NA"
    if (is.finite(Fv) && is.finite(as.numeric(df1)) && is.finite(as.numeric(df2res))) {
      out$aov_label <- sprintf(
        "paste(italic(F)[%s*','*%s], ' = ', %s, ', ', italic(p), ' = ', %s%s)",
        df1, df2res, Fv_str, pretty_p, star_plot
      )
    } else {
      out$aov_label <- sprintf("paste(italic(p), ' = ', %s%s)", pretty_p, star_plot)
    }
  } else {
    out$aov_label <- NULL
  }
  
  # Tukey only makes sense for classic AOV; compute but we'll gate display elsewhere
  out$tukey <- if (got_aov) tryCatch(rstatix::tukey_hsd(aov_res), error = function(e) NULL) else NULL
  out
}

# Site (urban vs reef): two-group test on parent-colony means (one value per colony), replacing the
# fragment-level ANOVA + Tukey across the four sites. Welch t-test on log10 colony means (plotCounts adds a
# 0.5 pseudocount, so counts are > 0); no post hoc test is needed for two groups.
compute_stats_habitat <- function(df_grp){
  df2 <- df_grp %>% dplyr::filter(!is.na(grp), is.finite(count), count > 0)
  out <- list(aov_text = NULL, aov_table = NULL, tukey = NULL, aov_label = NULL, aov_p = NA_real_)
  if (!nrow(df2)) return(out)
  n_grp <- table(droplevels(df2$grp))
  if (length(n_grp) < 2L || any(n_grp < 2L)) {
    out$aov_text <- "Welch t-test (colony means): too few colonies"
    return(out)
  }
  tt <- tryCatch(t.test(log10(count) ~ grp, data = df2, var.equal = FALSE), error = function(e) NULL)
  if (is.null(tt)) return(out)
  pv <- tt$p.value; tv <- unname(tt$statistic); dfv <- unname(tt$parameter)
  out$aov_p <- pv
  out$aov_table <- data.frame(Test = "Welch t-test, log10(colony mean normalized count)",
                              group1 = levels(droplevels(df2$grp))[1], group2 = levels(droplevels(df2$grp))[2],
                              n1 = as.integer(n_grp[1]), n2 = as.integer(n_grp[2]),
                              t = round(tv, 4), df = round(dfv, 2), p = signif(pv, 4))
  star <- if (is.finite(pv) && pv < 0.05) "*" else ""
  out$aov_text <- sprintf("Welch t(%.1f)=%.2f, p=%s%s", dfv, tv, formatC(pv, format = "g", digits = 3), star)
  pretty_p <- if (pv < 1e-4) "'< 1e-4'" else if (pv < 0.001) "'< 0.001'" else sprintf("'%s'", formatC(pv, format = "g", digits = 3))
  out$aov_label <- sprintf("paste(italic(t)['%.1f'], ' = ', '%.2f', ', ', italic(p), ' = ', %s%s)",
                           dfv, tv, pretty_p, if (nzchar(star)) ", '*'" else "")
  out
}

# Treatment panels: fragment-level one-way ANOVA + Tukey across the four treatments. Fragments from the same
# colony are not independent, so these tests are descriptive; inference on treatment comes from the DESeq2 model
# (~ genotype + treat). Site panels use colony means (compute_stats_habitat).
compute_stats_for <- function(df_grp, fac) {
  if (fac == "Site") compute_stats_habitat(df_grp) else compute_stats(df_grp)
}

pair_order <- function(levels_vec){
  pairs <- list()
  for (i in seq_along(levels_vec)) for (j in i + seq_len(length(levels_vec) - i)) pairs[[length(pairs)+1]] <- c(levels_vec[i], levels_vec[j])
  do.call(rbind, pairs) %>% as.data.frame() %>% setNames(c("group1","group2"))
}
order_tukey_by_levels <- function(tukey_tbl, levels_vec){
  if (is.null(tukey_tbl) || !nrow(tukey_tbl)) return(NULL)
  po <- pair_order(levels_vec)
  tukey_tbl %>%
    mutate(order_key = match(paste(group1, group2, sep="__"), paste(po$group1, po$group2, sep="__"))) %>%
    arrange(order_key, p.adj) %>% select(-order_key)
}

make_boxplot <- function(df, factor_name, species_label,
                         species_name,
                         tukey_tbl = NULL, aov_label = NULL, aov_text = NULL,
                         y_limits = NULL,
                         suppress_y_title = FALSE,   # NEW
                         suppress_x_text  = FALSE) { # NEW
  # determine levels, colors, and labels
  if (factor_name == "Site") {
    lvls <- habitat_levels
    cols <- habitat_colors
    label_map <- habitat_labels
  } else {
    lvls <- treat_levels
    cols <- treat_colors
    label_map <- treat_labels
  }
  df$grp <- factor(df$grp, levels = lvls)
  
  # base plot
  p <- ggpubr::ggboxplot(df, x = "grp", y = "count", color = "grey30", fill = "grp",
                         add = "jitter", add.params = list(size = 1, jitter = 0.25),
                         width = 0.7, size = 0.5) +
    labs(
      title = species_label,
      x = NULL,
      y = if (species_name == "Ssiderea") NULL else "Normalized counts (log scale)",
      fill = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "none",
      axis.text.x = element_text(
        size = if (factor_name == "Treatment") 8 else 10,
        lineheight = 0.9,
        angle = 0, hjust = 0.5, vjust = 1
      ),
      axis.text.y  = if (species_name == "Ssiderea") element_blank() else element_text(),
      axis.ticks.y = if (species_name == "Ssiderea") element_blank() else element_line(),
      plot.margin = margin(t = 6, r = 6, b = 18, l = 6),
      panel.border = element_rect(color = "black", fill = NA, size = 0.8),
      panel.grid = element_blank()
    ) +
    scale_fill_manual(values = setNames(cols, lvls)) +
    scale_x_discrete(labels = label_map)
  
  # --- unified y scale ---
  if (!is.null(y_limits)) {
    ylims_used <- y_limits
  } else {
    pos <- df$count[is.finite(df$count) & df$count > 0]
    if (length(pos)) {
      min_pos <- min(pos, na.rm = TRUE)
      max_pos <- max(pos, na.rm = TRUE)
      lower_pad <- 1/1.3
      upper_pad <- 1.3
      ymin <- max(min_pos * lower_pad, min_pos * 0.5, .Machine$double.eps)
      ymax <- max_pos * upper_pad
    } else {
      ymin <- 0.1; ymax <- 1
    }
    ylims_used <- c(ymin, ymax)
  }
  
  # ensure room for Tukey if present
  if (!is.null(tukey_tbl) && nrow(tukey_tbl)) {
    need_top <- .bracket_required_top(
      df, tukey_tbl, lvls,
      pad_dex   = 0.15,
      gap_dex   = 0.15,
      lbl_dex   = 0.005,
      min_sep_dex = 0.08,
      micro_dex = 0.020
    )
    if (is.finite(need_top)) ylims_used[2] <- max(ylims_used[2], need_top)
  }
  
  p <- p +
    scale_y_log10(
      breaks = scales::trans_breaks("log10", function(x) 10^x),
      labels = scales::trans_format("log10", scales::math_format(10^.x)),
      expand = expansion(mult = c(0.02, 0.12))
    ) +
    coord_cartesian(ylim = ylims_used, clip = "off")
  
  # --- Tukey brackets ---
  top_y <- ylims_used[2] * 0.98
  if (!is.null(tukey_tbl) && nrow(tukey_tbl)) {
    p <- .add_sig_brackets(
      p, tukey_tbl, lvls, top_y, df,
      pad_dex = 0.2,
      gap_dex = 0.15,
      tip_dex = 0.025,
      lbl_dex = 0.005
    )
  }
  
  # --- ANOVA label (bottom-center) ---
  lab_text  <- if (!is.null(aov_label) && nzchar(aov_label)) aov_label else aov_text
  use_parse <- !is.null(aov_label) && nzchar(aov_label)
  center_x  <- mean(seq_along(lvls))
  bottom_y  <- 10^(log10(ylims_used[1]) - 0.03)
  if (!is.null(lab_text) && nzchar(lab_text)) {
    p <- p + annotate("text", x = center_x, y = bottom_y, label = lab_text,
                      size = 3.3, vjust = 0, hjust = 0.5, parse = use_parse)
  }
  
  # --- NEW: per-position axis suppression for one-pagers ---
  if (isTRUE(suppress_y_title)) p <- p + theme(axis.title.y = element_blank())
  if (isTRUE(suppress_x_text))  p <- p + theme(axis.text.x  = element_blank())
  
  list(plot = p)
}
```


## 5b) One-page composite helpers (8.5x11 auto layout)

```{r 5b-one-page-composite-helpers-8-5x11-auto-layout}
# (keep .best_grid_dims and .page_title_strip you already have, or paste from below if missing)

.best_grid_dims <- function(n, content_w, content_h, target_ar, max_rc = 20L) {
  cand <- expand.grid(r = 1:max_rc, c = 1:max_rc)
  cand <- cand[cand$r * cand$c >= n, , drop = FALSE]
  cand$cell_w <- content_w / cand$c
  cand$cell_h <- content_h / cand$r
  cand$ar     <- cand$cell_w / cand$cell_h
  cand$score  <- abs(cand$ar - target_ar)
  cand$empty  <- (cand$r * cand$c) - n
  cand$area   <- cand$cell_w * cand$cell_h
  cand <- cand[order(cand$score, cand$empty, -cand$area), ]
  head(cand, 1)
}

.page_title_strip <- function(title_text) {
  ggplot() +
    geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf), fill="grey10", color=NA) +
    annotate("text", x=0, y=0, label=title_text, color="white", fontface="bold", size=12, hjust=0.5, vjust=0.5) +
    coord_cartesian(xlim=c(-1,1), ylim=c(-1,1), expand=FALSE) +
    theme_void()
}

.build_stress_family_legend <- function(palette, ncol = 3, title = "Stress family") {
  df_leg <- data.frame(fam = factor(names(palette), levels = names(palette)))
  p_leg <- ggplot(df_leg, aes(x = fam, y = 1, fill = fam)) +
    geom_col() +
    scale_fill_manual(values = palette, name = title) +
    guides(fill = guide_legend(ncol = ncol, byrow = TRUE)) +
    theme_void() +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text  = element_text(size = 8),
      legend.key.height = unit(0.4, "lines"),
      legend.key.width  = unit(0.8, "lines")
    )
  cowplot::get_legend(p_leg)
}

# ===== Fixed-grid, no-shrink, landscape, legend-in-empty-cell exporter =====

# ===== Fixed-grid, no-shrink, landscape, legend-in-empty-cell (NO page title) =====
export_subset_fixed_grid_no_shrink <- function(dat_sub, fac, rel, ogs, out_dir,
                                               rows, cols,
                                               panel_w = 8, panel_h = 4.5,
                                               margin_l = 0.35, margin_r = 0.35,
                                               margin_t = 0.30, margin_b = 0.30,
                                               legend_ncol = 3, tag = NULL) { # tag: added to the file name (stress family)
  if (!length(ogs)) return(invisible(NULL))
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  
  # Bigger legend
  .build_stress_family_legend_big <- function(palette, ncol = 3, title = "Stress family") {
    df_leg <- data.frame(fam = factor(names(palette), levels = names(palette)))
    p_leg <- ggplot(df_leg, aes(x = fam, y = 1, fill = fam)) +
      geom_col() +
      scale_fill_manual(values = palette, name = title) +
      guides(fill = guide_legend(ncol = ncol, byrow = TRUE)) +
      theme_void() +
      theme(
        legend.position = "bottom",
        legend.title = element_text(size = 12, face = "bold"),
        legend.text  = element_text(size = 11),
        legend.key.height = unit(0.6, "lines"),
        legend.key.width  = unit(1.2, "lines"),
        legend.box.margin = margin(2,2,2,2)
      )
    cowplot::get_legend(p_leg)
  }
  
  final_panels <- list()
  lvls <- if (fac == "Site") habitat_levels else treat_levels
  
  max_slots <- rows * cols
  n_place <- min(length(ogs), max_slots)
  
  # First pass: collect per-OG data
  tmp <- vector("list", n_place)
  for (i in seq_len(n_place)) {
    og <- ogs[i]
    row_meta <- dat_sub %>% dplyr::filter(orthogroup_ID==og) %>% dplyr::slice(1)
    if (!nrow(row_meta)) next
    of_df <- dat_sub %>% dplyr::filter(orthogroup_ID==og, species=="Ofaveolata")
    ss_df <- dat_sub %>% dplyr::filter(orthogroup_ID==og, species=="Ssiderea")
    if (!nrow(of_df) || !nrow(ss_df)) next
    pos <- c(of_df$count, ss_df$count); pos <- pos[is.finite(pos) & pos > 0]
    if (length(pos)) { ymin <- max(min(pos)/1.3, min(pos)*0.5, .Machine$double.eps); ymax <- max(pos)*1.3 } else { ymin <- 0.1; ymax <- 1 }
    ylims <- c(ymin, ymax)
    of_stats <- compute_stats_for(of_df, fac)
    ss_stats <- compute_stats_for(ss_df, fac)
    tmp[[i]] <- list(of_df=of_df, ss_df=ss_df, ylims=ylims, of_stats=of_stats, ss_stats=ss_stats,
                     gname = row_meta$gene_name %>% .[1], sfam = row_meta$stress_family %>% .[1])
  }
  
  # Second pass: build grobs with correct axis suppression
  for (i in seq_len(n_place)) {
    it <- tmp[[i]]; if (is.null(it)) next
    # Grid position
    r <- ceiling(i / cols)
    c <- ((i - 1) %% cols) + 1
    # X labels on any plot that does NOT have another real plot under it
    has_below <- (i + cols) <= n_place
    suppress_x_text <- has_below
    # Y title only on first column
    suppress_y_title_left <- !(c == 1)
    
    of_tukey_tbl <- if (!is.na(it$of_stats$aov_p) && is.finite(it$of_stats$aov_p) && it$of_stats$aov_p < 0.05)
      order_tukey_by_levels(it$of_stats$tukey, lvls) else NULL
    ss_tukey_tbl <- if (!is.na(it$ss_stats$aov_p) && is.finite(it$ss_stats$aov_p) && it$ss_stats$aov_p < 0.05)
      order_tukey_by_levels(it$ss_stats$tukey, lvls) else NULL
    
    of_plot <- make_boxplot(it$of_df, fac, species_titles$Ofaveolata,
                            species_name = "Ofaveolata",
                            tukey_tbl = of_tukey_tbl,
                            aov_label = it$of_stats$aov_label,
                            aov_text  = it$of_stats$aov_text,
                            y_limits  = it$ylims,
                            suppress_y_title = suppress_y_title_left,
                            suppress_x_text  = suppress_x_text)$plot
    
    ss_plot <- make_boxplot(it$ss_df, fac, species_titles$Ssiderea,
                            species_name = "Ssiderea",
                            tukey_tbl = ss_tukey_tbl,
                            aov_label = it$ss_stats$aov_label,
                            aov_text  = it$ss_stats$aov_text,
                            y_limits  = it$ylims,
                            suppress_y_title = TRUE,
                            suppress_x_text  = suppress_x_text)$plot
    
    # Keep your colored stress-family bar inside each mini-panel:
    final_panels[[length(final_panels)+1L]] <- panel_with_colored_title(of_plot, ss_plot,
                                                                        paste0(it$gname,"  (",ogs[i],")"),
                                                                        it$sfam)
  }
  
  # Fill remaining slots; put BIG legend in the first empty slot
  total_now <- length(final_panels)
  empty <- max_slots - total_now
  if (empty > 0) {
    legend_panel <- {
      leg <- .build_stress_family_legend_big(stress_family_palette, ncol = legend_ncol)
      cowplot::ggdraw() + cowplot::draw_grob(leg, 0.5, 0.5, 0.98, 0.98)
    }
    final_panels[[length(final_panels)+1L]] <- legend_panel
    if (empty > 1) {
      blanks <- replicate(empty - 1L, ggplot() + theme_void(), simplify = FALSE)
      final_panels <- c(final_panels, blanks)
    }
  }
  
  # Grid only (NO page title) — sanitize panels first
  final_panels <- Filter(.is_panel_ok, final_panels)
  final_panels <- lapply(final_panels, .as_grob_safe)
  
  grid_core <- cowplot::plot_grid(
    plotlist = final_panels,
    nrow = rows, ncol = cols,
    align = "hv", axis = "tblr"
  )
  
  # Page size (landscape), no title height
  page_w <- margin_l + cols * panel_w + margin_r
  page_h <- margin_t + rows * panel_h + margin_b
  
  outfile <- file.path(out_dir, sprintf("ONEPAGE_%s_%s%s_fixedgrid.pdf", fac, rel, if (is.null(tag)) "" else paste0("_", .safe(tag))))
  invisible(safe_ggsave_pdf(grid_core, outfile, width = page_w, height = page_h, limitsize = FALSE))
  # PNG copy of the one-page composite, shown in the knitted HTML after the main loop
  png_file <- file.path(tempdir(), sub("\\.pdf$", ".png", basename(outfile)))
  try(ggsave(png_file, plot = grid_core, width = page_w, height = page_h, units = "in", dpi = 100,
             limitsize = FALSE), silent = TRUE)
  
  # Aggressive cleanup to avoid OOM before next subset
  suppressWarnings(rm(list = intersect(c("final_panels", "grid_core"), ls()))); gc()
  invisible(if (file.exists(png_file)) png_file else NULL) # PNG path, shown in the knitted HTML
}
```


## Output root

```{r output-root}
panels_root <- "../../outputs/transcriptomics/orthofinder/plots_panels"
# clear previous panels/stats so outputs from earlier runs (e.g. old site-level panels) do not linger
unlink(panels_root, recursive = TRUE)
dir.create(panels_root, recursive = TRUE, showWarnings = FALSE)

targets <- tibble::tribble(
  ~factor_name, ~relationship, ~out_subdir,
  "Site", "Direct", "Site_Direct",
  "Site", "Inverse","Site_Inverse",
  "Treatment","Direct","Treatment_Direct",
  "Treatment","Inverse","Treatment_Inverse"
)
```


## Main loop (with fixed-grid, no-shrink, legend-in-empty-cell)

```{r main-loop-with-fixed-grid-no-shrink-legend-in-empty-cell}
onepage_imgs <- character(0) # one-page composites, shown in the knitted HTML after the loop
for (k in seq_len(nrow(targets))) {
  fac <- targets$factor_name[k]
  rel <- targets$relationship[k]
  out_dir <- file.path(panels_root, targets$out_subdir[k])
  # Site: Welch t-test on colony means (no Tukey); Treatment: fragment-level ANOVA + Tukey (descriptive)
  stats_subdir <- if (fac == "Site") "ttest" else "anova"
  stats_prefix <- if (fac == "Site") "Ttest" else "ANOVA"
  
  dat_sub <- counts_long %>% dplyr::filter(factor == fac, relationship == rel)
  if (!nrow(dat_sub)) {
    message(sprintf("No rows for %s | %s", fac, rel))
    next
  }
  # output folders are created only for subsets that have panels
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  if (fac != "Site") dir.create(file.path(out_dir, "tukey"), showWarnings = FALSE)
  dir.create(file.path(out_dir, stats_subdir), showWarnings = FALSE)
  
  if (fac == "Site") {
    # urban vs reef: average each gene's normalized counts per parent colony (genotype) within habitat,
    # so each colony contributes one point and the test compares colonies, not fragments
    stopifnot(all(c("habitat", "genotype") %in% names(dat_sub)))
    dat_sub <- dat_sub %>%
      dplyr::mutate(habitat = tolower(habitat)) %>%
      dplyr::filter(is.finite(count), !is.na(habitat), !is.na(genotype)) %>%
      dplyr::group_by(species, gene_id, orthogroup_ID, gene_name, stress_family, relationship,
                      factor, subset, habitat, genotype) %>%
      dplyr::summarise(count = mean(count), n_fragments = dplyr::n(), .groups = "drop") %>%
      dplyr::mutate(grp = factor(habitat, levels = habitat_levels))
    lvls <- habitat_levels
  } else {
    # Treatment panels stay fragment-level (one point per fragment); tests are descriptive
    stopifnot("treat" %in% names(dat_sub))
    dat_sub <- dat_sub %>% dplyr::mutate(grp = factor(treat, levels = treat_levels))
    lvls <- treat_levels
  }
  
  # Order by stress_family then orthogroup_ID for panel sequence
  ogs <- df_norm %>%
    dplyr::filter(factor==fac, relationship==rel) %>%
    dplyr::distinct(orthogroup_ID, .keep_all = TRUE) %>%
    dplyr::arrange(stress_family, orthogroup_ID) %>%
    dplyr::pull(orthogroup_ID)
  ogs <- intersect(ogs, unique(dat_sub$orthogroup_ID))
  
  panel_list <- list()
  p_i <- 0L
  
  for (og in ogs) {
    row_meta <- dat_sub %>% dplyr::filter(orthogroup_ID == og) %>% dplyr::slice(1)
    gname <- row_meta$gene_name %>% .[1]
    sfam  <- row_meta$stress_family %>% .[1]
    
    of_df <- dat_sub %>% dplyr::filter(orthogroup_ID == og, species == "Ofaveolata")
    ss_df <- dat_sub %>% dplyr::filter(orthogroup_ID == og, species == "Ssiderea")
    
    if (!nrow(of_df) || !nrow(ss_df)) {
      missing_species <- paste(
        c("Ofaveolata", "Ssiderea")[c(!nrow(of_df), !nrow(ss_df))],
        collapse = ", "
      )
      message(sprintf("Skipping OG=%s (missing species: %s)", og, missing_species))
      next
    }
    
    # shared y-limits across both species for this OG (log-safe)
    counts_both <- c(of_df$count, ss_df$count)
    pos <- counts_both[is.finite(counts_both) & counts_both > 0]
    
    if (length(pos)) {
      min_pos <- min(pos, na.rm = TRUE)
      max_pos <- max(pos, na.rm = TRUE)
      lower_pad <- 1 / 1.3
      upper_pad <- 1.3
      ymin <- max(min_pos * lower_pad, min_pos * 0.5, .Machine$double.eps)
      ymax <- max_pos * upper_pad
    } else {
      ymin <- 0.1
      ymax <- 1
    }
    ylims <- c(ymin, ymax)
    
    of_stats <- compute_stats_for(of_df, fac)
    ss_stats <- compute_stats_for(ss_df, fac)
    
    # Save stats CSVs
    if (!is.null(of_stats$tukey) && nrow(of_stats$tukey))
      readr::write_csv(order_tukey_by_levels(of_stats$tukey, lvls),
                       file.path(out_dir, "tukey", paste0("Tukey_Ofaveolata_", .safe(og), ".csv")))
    if (!is.null(ss_stats$tukey) && nrow(ss_stats$tukey))
      readr::write_csv(order_tukey_by_levels(ss_stats$tukey, lvls),
                       file.path(out_dir, "tukey", paste0("Tukey_Ssiderea_", .safe(og), ".csv")))
    if (!is.null(of_stats$aov_table))
      readr::write_csv(of_stats$aov_table,
                       file.path(out_dir, stats_subdir, paste0(stats_prefix, "_Ofaveolata_", .safe(og), ".csv")))
    if (!is.null(ss_stats$aov_table))
      readr::write_csv(ss_stats$aov_table,
                       file.path(out_dir, stats_subdir, paste0(stats_prefix, "_Ssiderea_", .safe(og), ".csv")))
    
    # Gate Tukey: only pass it through if omnibus ANOVA is significant (Site has no Tukey table)
    of_tukey_tbl <- if (!is.na(of_stats$aov_p) && is.finite(of_stats$aov_p) && of_stats$aov_p < 0.05)
      order_tukey_by_levels(of_stats$tukey, lvls) else NULL
    ss_tukey_tbl <- if (!is.na(ss_stats$aov_p) && is.finite(ss_stats$aov_p) && ss_stats$aov_p < 0.05)
      order_tukey_by_levels(ss_stats$tukey, lvls) else NULL
    
    of_plot <- make_boxplot(of_df, fac, species_titles$Ofaveolata,
                            species_name = "Ofaveolata",
                            tukey_tbl = of_tukey_tbl,
                            aov_label = of_stats$aov_label,
                            aov_text  = of_stats$aov_text,
                            y_limits  = ylims)$plot
    
    ss_plot <- make_boxplot(ss_df, fac, species_titles$Ssiderea,
                            species_name = "Ssiderea",
                            tukey_tbl = ss_tukey_tbl,
                            aov_label = ss_stats$aov_label,
                            aov_text  = ss_stats$aov_text,
                            y_limits  = ylims)$plot
    
    panel <- panel_with_colored_title(of_plot, ss_plot, paste0(gname, "  (", og, ")"), sfam)
    p_i <- p_i + 1L
    outfile <- file.path(out_dir,
                         paste0("PANEL_", sprintf("%03d", p_i), "_", .safe(og), "_", .safe(gname), ".pdf"))
    # old:
    # ggsave(outfile, panel, width = 8, height = 4.5, dpi = 300, device = grDevices::pdf, useDingbats = FALSE)
    # new:
    safe_ggsave_pdf(panel, outfile, width = 8, height = 4.5, limitsize = TRUE)
    message("Wrote panel: ", normalizePath(outfile, mustWork = FALSE))
    panel_list[[p_i]] <- panel
  }
  
  if (length(panel_list)) {
    # (Optional) keep your ALL_panels export
    ggexport(plotlist = panel_list,
             filename = file.path(out_dir, paste0("ALL_panels_", fac, "_", rel, ".pdf")),
             nrow = 1, ncol = 1, width = 8, height = 4.5, dpi = 300)
    
    # Fixed-grid, no-shrink, legend-in-empty-cell (LANDSCAPE), NO page title
    # grid size from the number of panels (+1 cell for the stress-family legend)
    # one composite per stress family, so the panels stay readable
    for (sf in names(stress_family_palette)) {
      ogs_sf <- ogs[ogs %in% unique(dat_sub$orthogroup_ID[dat_sub$stress_family == sf])]
      if (!length(ogs_sf)) next
      n_cells   <- length(ogs_sf) + 1L
      grid_cols <- ceiling(sqrt(n_cells))
      grid_rows <- ceiling(n_cells / grid_cols)
      onepage_imgs <- c(onepage_imgs, export_subset_fixed_grid_no_shrink(dat_sub[dat_sub$stress_family == sf, ], fac, rel, ogs_sf, out_dir,
                                                                         rows = grid_rows, cols = grid_cols, tag = sf))
    }
  }
  
  # ✅ per-subset panel summary
  message(sprintf("📊 %d panels exported for %s | %s → %s",
                  length(panel_list), fac, rel, normalizePath(out_dir, mustWork = FALSE)))
  
  # Free memory before moving to the next subset
  suppressWarnings(rm(list = intersect(c("dat_sub", "panel_list", "of_plot", "ss_plot", "of_df", "ss_df",
                                         "of_stats", "ss_stats", "ylims", "counts_both", "pos"), ls())))
  gc()
  
}
if (length(onepage_imgs) && isTRUE(getOption("knitr.in.progress"))) knitr::include_graphics(onepage_imgs, rel_path = FALSE, dpi = 100)
```


## Summaries

```{r summaries}
message("\nSummary by subset (panels):")
for (k in seq_len(nrow(targets))) {
  subdir <- file.path(panels_root, targets$out_subdir[k])
  npan <- length(list.files(subdir, pattern="^PANEL_.*\\.pdf$", full.names=TRUE))
  message(sprintf("  %s | %s : %d panels", targets$factor_name[k], targets$relationship[k], npan))
}

message("\nCounts export summary (from earlier):"); print(res_counts)
message("\n✅ Done. Counts in 'counts/', panels + Tukey/ANOVA (Treatment) and t-test (Site) CSVs in 'plots_panels/'.")

message("\nSummary by subset (Tukey & ANOVA / t-test CSVs):")
for (k in seq_len(nrow(targets))) {
  subdir  <- file.path(panels_root, targets$out_subdir[k])
  tuk_dir <- file.path(subdir, "tukey")
  aov_dir <- file.path(subdir, if (targets$factor_name[k] == "Site") "ttest" else "anova")
  ntukey <- if (dir.exists(tuk_dir)) length(list.files(tuk_dir, pattern="\\.csv$", full.names=TRUE)) else 0L
  nanova <- if (dir.exists(aov_dir)) length(list.files(aov_dir, pattern="\\.csv$", full.names=TRUE)) else 0L
  message(sprintf("  %s | %s : Tukey=%d, %s=%d", targets$factor_name[k], targets$relationship[k], ntukey,
                  if (targets$factor_name[k] == "Site") "t-test" else "ANOVA", nanova))
}

# 🔎 List the single-page outputs
onepagers <- list.files(panels_root, pattern = "^ONEPAGE_.*\\.pdf$", recursive = TRUE, full.names = TRUE)
message("\nFound ", length(onepagers), " one-page PDFs.")
if (length(onepagers)) cat(paste0(" - ", normalizePath(onepagers, mustWork = FALSE), "\n"), sep = "")
```
