---
title: "O. faveolata Coexpression Networks — WGCNA"
author: "Michael Studivan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# -----------------------------------------------------------------------------
# Global chunk options
# echo = TRUE:     show code in rendered output
# warning = FALSE: suppress warnings in rendered document
#                  (warnings still print to console during knitting)
# message = FALSE: suppress package loading messages
# -----------------------------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 300
)

# save_plot(): writes a figure to PDF (outputs folder) and shows the same figure in the knitted HTML.
# The plotting code is run once for the PDF and once for a PNG copy at the same size, which is embedded in
# the HTML; when run interactively (not knitting) the plot is also drawn in the plot pane.
save_plot <- function(file, width, height, expr) {
  expr <- substitute(expr); env <- parent.frame()
  # draw on a newly opened device and close exactly that device (never whichever device happens to be current).
  # Some plotting functions move drawing away from that device: pheatmap() opens and closes a temporary device
  # (pdf(NULL) ... dev.off()) before drawing, and R then makes some other open device current, which left PDFs blank or
  # missing from the HTML. RStudio/knitr graphics hooks can do the same. While the file is drawn, those hooks are
  # replaced by one that switches back to the file's device each time a new plot page starts (plot.new / grid.newpage).
  hooks <- c("before.plot.new", "plot.new", "before.grid.newpage", "grid.newpage")
  draw_on <- function(open_device) {
    saved <- lapply(setNames(hooks, hooks), getHook)
    open_device(); d <- dev.cur()
    back_to_d <- function() if (d %in% dev.list() && dev.cur() != d) dev.set(d)
    for (h in hooks) setHook(h, NULL, "replace")
    setHook("before.plot.new", back_to_d); setHook("before.grid.newpage", back_to_d)
    on.exit({ if (d %in% dev.list()) dev.off(d)
              for (h in hooks) setHook(h, saved[[h]], "replace") })
    eval(expr, env)
  }
  draw_on(function() pdf(file = file, width = width, height = height))
  if (!isTRUE(getOption("knitr.in.progress"))) { # interactive: also draw in the plot pane (skipped if the pane is too small)
    ok <- try(eval(expr, env), silent = TRUE)
    if (inherits(ok, "try-error") && dev.cur() > 1)
      try(dev.off(dev.cur()), silent = TRUE) # reset the screen device left in a broken state (the PDF is already closed)
    return(invisible(file)) }
  png_file <- file.path(tempdir(), sub("\\.pdf$", ".png", basename(file)))
  draw_on(function() png(png_file, width = width, height = height, units = "in", res = 150))
  if (!file.exists(png_file)) { warning("nothing was drawn for ", basename(file)); return(invisible(file)) }
  knitr::include_graphics(png_file, rel_path = FALSE, dpi = 150) # displayed at its true size
}
```


# PACKAGES

```{r packages}
# installing WGCNA:
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.17")
# BiocManager::install(c("AnnotationDbi", "impute", "GO.db", "preprocessCore"))
# install.packages("flashClust")
# install.packages("WGCNA",dependencies=TRUE)
# repos="http://cran.us.r-project.org"
# run these above commands once, then comment out

# always run these before running any of the following script chunks
library(WGCNA)
library(flashClust)
library(ape)
options(stringsAsFactors=FALSE)
allowWGCNAThreads() # ?WGCNA
```


# DATA IMPORT and TRAITS

```{r data-import-and-traits}
# importing data generated from DESeq2 script
lnames=load("../../outputs/transcriptomics/ofav/deseq2/data4wgcna.RData")
lnames # "vsd.wg"  "design" # log-transformed variance-stabilized gene expression, and table or experimental conditions
datt=t(vsd.wg)
ncol(datt)
nrow(datt)

head(design)
str(design)

# assembling table of traits
# physiological traits are rebuilt here from the raw data files (rather than taken from design columns)
# so each value is traceable to its source analysis; one row per sample, in design order
library(dplyr)
stopifnot(identical(sub("\\..*$", "", rownames(datt)), as.character(design$id))) # expression rows and design rows aligned

# fragments removed from the survivorship analysis (non-empty `exclude`); their samples are removed upstream in
# the DESeq2 script, and any that remain get NA for all measured traits
surv_raw <- read.csv("../../data/survivorship/urban survivorship.csv", head=TRUE)
surv_excluded <- unique(surv_raw$sample[!is.na(surv_raw$exclude) & surv_raw$exclude != ""])
excl_samples <- design$id[design$sampleid %in% surv_excluded]
cat("samples of survivorship-excluded fragments still in design:", length(excl_samples), "\n")
print(excl_samples)

# growth rate (g cm-2 yr-1): buoyant weights 26 Feb and 3 Apr 2021 (36 days), annualized, minus the mean growth of
# the dead skeleton controls for each pH treatment (as in the growth analysis)
bw <- read.csv("../../data/growth/urban buoyant weight.csv", head=TRUE) %>%
  mutate(growth = (mass2 - mass1) / surfacearea / 36 * 365)
bw_controls <- bw %>%
  filter(species == "control") %>%
  group_by(pH) %>%
  summarise(control_growth = mean(growth), .groups = "drop")
bw <- bw %>%
  filter(species != "control") %>%
  left_join(bw_controls, by = "pH") %>%
  mutate(growthadj = growth - control_growth) %>%
  group_by(species) %>%
  # growth outliers (1.5 x IQR within species, as in the growth analysis) set to NA
  mutate(growth_outlier = growthadj %in% boxplot.stats(growthadj)$out) %>%
  ungroup() %>%
  mutate(growthrate = ifelse(growth_outlier | sampleID %in% surv_excluded, NA, growthadj))
bw %>% filter(growth_outlier) %>% dplyr::select(species, sampleID, growthadj) # growth outliers set to NA
stopifnot(!anyDuplicated(bw$sampleID))

# Fv/Fm and Fv/Fm ratio at each fragment's cohort day 17 (the data file's second-cohort timepoints are already
# on the logger timeline, so day 17 is time 17 for every group)
ipam <- read.csv("../../data/photosynthesis/urban ipam.csv", head=TRUE, colClasses = c(time = "character")) %>%
  filter(time == "17") %>%
  mutate(temp = recode(temp, "control temp" = "controltemp", "high temp" = "hightemp")) %>%
  dplyr::select(sampleid = sampleID, temp, fvfm = FvFm, fvfm_ratio = ratio)
stopifnot(!anyDuplicated(ipam[c("sampleid", "temp")]))

# survival (days at 33 C set point until death or end of experiment) from each fragment's heat-treatment row;
# (second heat cohort dates in the data file follow the tank loggers: start 5/10/21).
# The value applies to every sample of the fragment: for control temperature samples it is the fragment's
# later survival under heat stress (second cohort), not survival during the control temperature period
surv <- surv_raw %>%
  filter(is.na(exclude) | exclude == "", temp == "high") %>%
  mutate(survival = as.numeric(as.Date(end, "%m/%d/%y") - as.Date(start, "%m/%d/%y"))) %>%
  dplyr::select(sampleid = sample, survival)
stopifnot(!anyDuplicated(surv$sampleid))

# coding experimental factors as binary (0/1, yes/no)
# habitat: urban (Star Island, MacArthur North) vs reef (Emerald, Rainbow)
phys <- data.frame(id = as.character(design$id), sampleid = as.character(design$sampleid),
                   temp = as.character(design$temp)) %>%
  left_join(dplyr::select(bw, sampleid = sampleID, growthrate), by = "sampleid") %>%
  left_join(ipam, by = c("sampleid", "temp")) %>%
  left_join(surv, by = "sampleid")
stopifnot(identical(phys$id, as.character(design$id))) # joins kept design order, one row per sample

traits <- data.frame(
  urban       = as.numeric(design$site %in% c("Star","MacN")),
  lowpH       = as.numeric(design$ph == "lowpH"),
  hightemp    = as.numeric(design$temp == "hightemp"),
  growthrate  = phys$growthrate,
  fvfm        = phys$fvfm,
  fvfm_ratio  = phys$fvfm_ratio,
  survival    = phys$survival,
  propB       = design$propB,
  propD       = design$propD,
  temp_actual = design$temp_actual,
  ph_actual   = design$ph_actual,
  row.names   = design$id
)
# measured traits NA for any remaining samples of survivorship-excluded fragments
traits[rownames(traits) %in% excl_samples, c("growthrate","fvfm","fvfm_ratio","survival","propB","propD","temp_actual","ph_actual")] <- NA
summary(traits)
colSums(is.na(traits)) # missing values per trait
```


# OUTLIER DETECTION

```{r outlier-detection}
# identifies outlier genes
gsg = goodSamplesGenes(datt, verbose = 3);
gsg$allOK #if TRUE, no outlier genes or samples
# drop any flagged genes/samples, keeping expression, traits and design aligned
if (!gsg$allOK) {
  cat("removing", sum(!gsg$goodGenes), "genes and", sum(!gsg$goodSamples), "samples flagged by goodSamplesGenes\n")
  datt = datt[gsg$goodSamples, gsg$goodGenes, drop = FALSE]
  traits = traits[gsg$goodSamples, ]
  design = design[gsg$goodSamples, ]
}
stopifnot(identical(sub("\\..*$", "", rownames(datt)), rownames(traits)),
          identical(rownames(traits), as.character(design$id)))

# calculates mean expression per array, then the number of missing values per array
meanExpressionByArray=apply( datt,1,mean, na.rm=T)
NumberMissingByArray=apply( is.na(data.frame(datt)),1, sum)
NumberMissingByArray
```

Keep samples with missing values under 500  

```{r outlier-detection-cont}
# plots mean expression across all samples
save_plot("../../outputs/transcriptomics/ofav/wgcna/sample mean expression.pdf",height=4, width=8, {
barplot(meanExpressionByArray,
        xlab = "Sample", ylab = "Mean expression",
        main ="Mean expression across samples",
        names.arg = 1:nrow(datt), cex.names = 0.7)
}) # look for any obvious deviations in expression across samples

# sample dendrogram and trait heat map showing outliers
A=adjacency(t(datt),type="signed")                 #SELECT SIGNED OR UNSIGNED HERE
# this calculates the whole network connectivity
k=as.numeric(apply(A,2,sum))-1
# standardized connectivity
Z.k=scale(k)
thresholdZ.k=-2.5 # often -2.5
outlierColor=ifelse(Z.k<thresholdZ.k,"red","black")
sampleTree = flashClust(as.dist(1-A), method = "average")
# Convert traits to a color representation where red indicates high values
traitColors=data.frame(numbers2colors(traits,signed=FALSE))
dimnames(traitColors)[[2]]=paste(names(traits))
datColors=data.frame(outlierC=outlierColor,traitColors)
# Plot the sample dendrogram and the colors underneath.
# quartz()  # Mac-only; plots now render inline / in RStudio
# page size scales with the number of samples (width) and trait rows (height)
save_plot("../../outputs/transcriptomics/ofav/wgcna/outlier detection.pdf", width = max(10, 3 + 0.15 * nrow(datt)), height = 5 + 0.3 * ncol(datColors), {
plotDendroAndColors(sampleTree,groupLabels=names(datColors), colors=datColors,main="Sample dendrogram and trait heatmap")
})
```

The resulting plot shows a sample dendrogram and the spread of your traits across the cluster  
Outlier samples will show as red in the outlierC row

```{r outlier-detection-cont2}
# Remove outlying samples from expression and trait data
remove.samples= Z.k<thresholdZ.k | is.na(Z.k)
# report which samples (if any) are dropped from the network; state these in the Methods
cat("samples removed as connectivity outliers (Z.k <", thresholdZ.k, "):",
    if (any(remove.samples)) paste(rownames(datt)[remove.samples], collapse = ", ") else "none", "\n")
datt=datt[!remove.samples,]
traits=traits[!remove.samples,]
# parent colony of each fragment (used for colony-level tests of habitat)
colony=factor(design$genotype[!remove.samples])
names(colony)=as.character(design$id[!remove.samples])
# expression, traits and colony still refer to the same samples in the same order
stopifnot(identical(sub("\\..*$", "", rownames(datt)), rownames(traits)),
          identical(rownames(traits), names(colony)))

# trait table with the sample id as an explicit column
write.csv(data.frame(id = rownames(traits), traits), file="../../outputs/transcriptomics/ofav/wgcna/traits.csv", row.names=FALSE)

save(datt,traits,colony,file="../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData")
```


# SOFT THRESHOLDS

```{r soft-thresholds}
library(WGCNA)
library(flashClust)
library(ape)
options(stringsAsFactors=FALSE)
allowWGCNAThreads()
load("../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData")

# Try different betas ("soft threshold") - power factor for calling connections between genes
powers = c(seq(from = 2, to=26, by=1))
# Call the network topology analysis function
sft = pickSoftThreshold(datt, powerVector = powers, verbose = 5,networkType="signed")

# Plot the results:
save_plot("../../outputs/transcriptomics/ofav/wgcna/soft threshold signed.pdf",height=4, width=8, {

par(mfrow = c(1,2));
cex1 = 0.9;

# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
labels=powers,cex=cex1,col="red");

# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
})
```


# MAKING MODULES

```{r making-modules}
# take a look at the threshold plots produced above, and the output table from the pickSoftThreshold command
# pick the power that corresponds with a SFT.R.sq value above 0.90

# run from the line below to the save command
s.th=8 # re-specify according to previous section
# memory: each gene x gene matrix is ~4 GB for ~23,000 genes, so holding adjacency, TOM and dissTOM together exceeded
# R's 16 GB limit. TOM is computed directly from expression (the same signed adjacency and signed TOM as
# adjacency() + TOMsimilarity()) and then converted to dissimilarity in place; only dissTOM is kept.
dissTOM = TOMsimilarityFromExpr(datt, power = s.th, networkType = "signed", TOMType = "signed")
dissTOM = 1 - dissTOM # TOM -> TOM-based dissimilarity
invisible(gc())
# Call the hierarchical clustering function
geneTree = flashClust(as.dist(dissTOM), method = "average")
save_plot("../../outputs/transcriptomics/ofav/wgcna/gene dendrogram.pdf", width = 12, height = 6, {
plot(geneTree, xlab="", sub="", main="Gene Clustering on TOM-based dissimilarity", labels= FALSE,hang=0.04)
})

# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30; 
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
deepSplit = 2, pamRespectsDendro = FALSE,
minClusterSize = minModuleSize);
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)

# Calculate eigengenes
MEList = moduleEigengenes(datt, colors = dynamicColors)
MEs = MEList$eigengenes
# the grey module contains unassigned genes and is not considered a real module
# if you have error messages trying to generate the eigengene correlations, run this below
# check MEs, if grey shows NaN for all samples, then make sure to eliminate it using removeGreyME
# MEs = removeGreyME(MEs, greyMEName = paste(moduleColor.getMEprefix(), "grey", sep=""))
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
METree = flashClust(as.dist(MEDiss), method = "average");

save(dynamicMods,dynamicColors,MEs,METree,geneTree,s.th,file="../../outputs/transcriptomics/ofav/wgcna/1stPassModules.RData")
```


# MERGING MODULES

```{r merging-modules}
mm=load('../../outputs/transcriptomics/ofav/wgcna/1stPassModules.RData')
mm
lnames=load('../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData')
# traits
# head(datt)

# quartz()  # Mac-only; plots now render inline / in RStudio

MEDissThres = 0.25 # fixed a priori (WGCNA tutorial default): merge modules whose eigengenes correlate at r > 0.75; see MERGE THRESHOLD SENSITIVITY for 0.15-0.35 (set to 0 only to view unmerged modules)
# page width scales with the number of modules
save_plot("../../outputs/transcriptomics/ofav/wgcna/dendrogram raw.pdf", # unmerged module eigengenes, with the merge cut line
          width = max(6, 2 + 0.35 * length(METree$order)), height = 6, {
plot(METree, main = "Clustering of module eigengenes",
xlab = "", sub = "")
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")  # on 2nd pass: does this cut height meet your merging goals? If not, reset MEDissThres and replot
})

# Call an automatic merging function
merge = mergeCloseModules(datt, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors;
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs

# plotting the fabulous ridiculogram
# quartz()  # Mac-only; plots now render inline / in RStudio
save_plot("../../outputs/transcriptomics/ofav/wgcna/ridiculogram.pdf", # unmerged (dynamic tree cut) and merged module colours
          width = 12, height = 6, {
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
c("Dynamic Tree Cut", "Merged dynamic"),
dendroLabels = FALSE, hang = 0.03,
addGuide = FALSE, guideHang = 0.05,lwd=0.3)
})

# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;

# Calculate dissimilarity of module eigengenes
# quartz()  # Mac-only; plots now render inline / in RStudio
# the grey module contains unassigned genes and is not considered a real module
# if you have error messages trying to generate the eigengene correlations, run this below
# check MEs, if grey shows NaN for all samples, then make sure to eliminate it using removeGreyME
# MEs = removeGreyME(MEs, greyMEName = paste(moduleColor.getMEprefix(), "grey", sep=""))
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = flashClust(as.dist(MEDiss), method = "average");
# Plot the result
save_plot("../../outputs/transcriptomics/ofav/wgcna/dendrogram merged.pdf",
          width = max(6, 2 + 0.35 * length(METree$order)), height = 6, {
plot(METree, main = "Clustering of module eigengenes",
xlab = "", sub = "")
})

# how many genes in each module?
table(moduleColors)
# Save module colors and labels (and the soft-threshold power) for use in subsequent parts
save(MEs, geneTree, moduleLabels, moduleColors, s.th, MEDissThres, file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
```


# MODULE CORRELATIONS

```{r module-correlations}
# plotting correlations with traits:
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData");

# Define numbers of genes and samples
nGenes = ncol(datt);
nSamples = nrow(datt);
# Recalculate MEs with color labels
MEs0 = moduleEigengenes(datt, moduleColors)$eigengenes
MEs = orderMEs(MEs0)

# correlations of genes with eigengenes
moduleGeneCor=cor(MEs,datt)
moduleGenePvalue = corPvalueStudent(moduleGeneCor, nSamples);

# module-trait heatmaps (all correlations; significant only) and module size barplot for one module assignment.
# Drawn for the unmerged modules ("raw", dynamic tree cut) and the merged modules used in all later analyses ("merged").
library(RColorBrewer)
module_trait_plots = function(MEs, moduleColors, label) {
  # p-values from each module-trait pair's own number of non-missing samples (traits can have NAs)
  P = corAndPvalue(MEs, traits, use = "p")
  cor_m = P$cor; p_m = P$p
  modLabels = sub("ME", "", names(MEs))
  # page size scales with the number of traits (width) and modules (height): ~0.55 x 0.3 in per cell
  hm_w = 2.5 + 0.55 * ncol(traits); hm_h = 1.5 + 0.3 * ncol(MEs)
  text_all = paste(signif(cor_m, 2), "\n(", signif(p_m, 1), ")", sep = "")
  dim(text_all) = dim(cor_m)
  text_sig = signif(cor_m, 2)
  text_sig[signif(p_m, 1) > 0.05] = "-"
  dim(text_sig) = dim(cor_m)
  mct = table(moduleColors)
  labelShift = 0.08 * max(mct) # label offset scales with the largest module
  write.csv(data.frame(module = modLabels, genes = as.vector(mct[modLabels]), cor_m, check.names = FALSE),
            paste0("../../outputs/transcriptomics/ofav/wgcna/module trait correlations ", label, ".csv"), row.names = FALSE)
  write.csv(data.frame(module = modLabels, p_m, check.names = FALSE),
            paste0("../../outputs/transcriptomics/ofav/wgcna/module trait pvalues ", label, ".csv"), row.names = FALSE)
  imgs = list(
    save_plot(paste0("../../outputs/transcriptomics/ofav/wgcna/module trait correlations ", label, " all.pdf"), width = hm_w, height = hm_h, {
      par(mar = c(6, 8.5, 3, 3))
      labeledHeatmap(Matrix = cor_m, xLabels = names(traits), yLabels = names(MEs), ySymbols = names(MEs),
                     colorLabels = FALSE, colors = blueWhiteRed(50), textMatrix = text_all,
                     setStdMargins = FALSE, cex.text = 0.5, zlim = c(-1, 1),
                     main = paste0("Module-trait relationships (", label, ")"))
    }),
    save_plot(paste0("../../outputs/transcriptomics/ofav/wgcna/module trait correlations ", label, " sig.pdf"), width = hm_w, height = hm_h, {
      par(mar = c(6, 8.5, 3, 3))
      labeledHeatmap(Matrix = cor_m, xLabels = names(traits), ySymbols = modLabels, yLabels = modLabels,
                     colorLabels = FALSE, colors = colorRampPalette(c("blue","lightblue","white","coral","red"))(50),
                     textMatrix = text_sig, setStdMargins = FALSE, cex.text = 0.7, zlim = c(-0.7,0.7),
                     main = paste0("O. faveolata Module-Trait correlations (", label, ")"))
    }),
    save_plot(paste0("../../outputs/transcriptomics/ofav/wgcna/module size barplot ", label, ".pdf"), width = 7, height = 2 + 0.3 * length(modLabels), {
      par(mar = c(6, 8.5, 3, 3))
      x = barplot(mct[rev(modLabels)], horiz = T, las = 1, xlim = c(0, 1.2 * max(mct)), col = rev(modLabels))
      text(mct[rev(modLabels)] + labelShift, y = x, mct[rev(modLabels)], cex = 0.9)
    }))
  list(cor = cor_m, p = p_m, imgs = unlist(imgs))
}

# unmerged modules (dynamic tree cut, before merging), for comparison
raw_env = new.env()
load(file = "../../outputs/transcriptomics/ofav/wgcna/1stPassModules.RData", envir = raw_env) # dynamicColors
raw_colors = raw_env$dynamicColors
raw_MEs = orderMEs(moduleEigengenes(datt, raw_colors)$eigengenes)
raw = module_trait_plots(raw_MEs, raw_colors, "raw")
table(raw_colors) # genes in each unmerged module

# merged modules (used for everything below)
merged = module_trait_plots(MEs, moduleColors, "merged")
table(moduleColors) # genes in each merged module
moduleTraitCor = merged$cor
moduleTraitPvalue = merged$p
modLabels = sub("ME", "", names(MEs))
if (isTRUE(getOption("knitr.in.progress"))) knitr::include_graphics(c(raw$imgs, merged$imgs), rel_path = FALSE, dpi = 150)

# modules significantly correlated (p < 0.05) with each trait, and with any trait; grey (unassigned genes) is not a module
sig_by_trait = lapply(setNames(colnames(moduleTraitPvalue), colnames(moduleTraitPvalue)), function(tr)
  setdiff(sub("^ME", "", rownames(moduleTraitPvalue)[which(moduleTraitPvalue[, tr] < 0.05)]), "grey"))
sig_by_trait
traits_of_interest = colnames(traits) # restrict here to limit which traits define sig_modules
sig_modules = unique(unlist(sig_by_trait[traits_of_interest]))
urban_modules = sig_by_trait[["urban"]] # modules correlated with habitat (fragment level; see colony-level test below)
sig_modules
urban_modules
# modules used for scatterplots, sanity checks, GO/KOG export and heatmaps below
modules_to_plot = sig_modules # or set manually, e.g. modules_to_plot = c("<module>", "<module>")
save(moduleTraitCor, moduleTraitPvalue, sig_by_trait, sig_modules, urban_modules, modules_to_plot,
     file = "../../outputs/transcriptomics/ofav/wgcna/moduleTraitSig.RData")

# the merge threshold is fixed a priori (MEDissThres = 0.25), not chosen from these trait correlations;
# MERGE THRESHOLD SENSITIVITY below checks that the trait associations hold across thresholds
```


# URBAN MODULES AT COLONY LEVEL

```{r urban-modules-colony-level}
# module-trait correlations above treat fragments as independent; urban vs reef differs only between parent colonies,
# so habitat associations are re-tested with colony as the replicate (treatment effects removed, fragments averaged per colony)
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData")
MEs = orderMEs(moduleEigengenes(datt, moduleColors)$eigengenes)
treat_group = factor(paste(traits$lowpH, traits$hightemp))
# treatment effects estimated within colonies (lm with colony), then only the treatment contribution is subtracted,
# so colony (and therefore habitat) differences stay in the adjusted eigengenes
MEs_adj = apply(MEs, 2, function(y) {
  fit = lm(y ~ treat_group + colony)
  b = coef(fit)[grepl("^treat_group", names(coef(fit)))]
  b[is.na(b)] = 0
  y - drop(model.matrix(fit)[, names(b), drop = FALSE] %*% b)
})
colony_MEs = apply(MEs_adj, 2, function(y) tapply(y, colony, mean))          # average fragments per colony
colony_urban = tapply(traits$urban, colony, `[`, 1)
table(colony_urban) # colonies per habitat (0 = reef, 1 = urban)
urban_colony = t(apply(colony_MEs, 2, function(y) {
  ct = cor.test(y, colony_urban); c(r = unname(ct$estimate), p = ct$p.value) }))
# Benjamini-Hochberg adjustment across modules (grey, unassigned genes, not included)
is_module = rownames(urban_colony) != "MEgrey"
p_BH = rep(NA, nrow(urban_colony))
p_BH[is_module] = p.adjust(urban_colony[is_module, "p"], method = "BH")
urban_colony = data.frame(r = round(urban_colony[, "r"], 2), p = signif(urban_colony[, "p"], 2), p_BH = signif(p_BH, 2),
                          row.names = rownames(urban_colony))
urban_colony
write.csv(urban_colony, file = "../../outputs/transcriptomics/ofav/wgcna/urban modules colony level.csv")
```


# MERGE THRESHOLD SENSITIVITY

```{r merge-threshold-sensitivity}
# The merge threshold is fixed a priori (MEDissThres = 0.25). Here the unmerged modules are re-merged at a range of
# thresholds to check that the trait associations do not depend on that choice. For each threshold: the number of
# modules (grey, unassigned genes, excluded) and, per trait, the number of modules significantly correlated with it
# (p < 0.05, fragment level as in the module-trait heatmap); urban_colony_BH = modules differing between urban and
# reef colonies (colony-level test as in URBAN MODULES AT COLONY LEVEL, BH-adjusted p < 0.05).
library(WGCNA)
load(file = "../../outputs/transcriptomics/ofav/wgcna/1stPassModules.RData") # dynamicColors (unmerged modules)
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData")      # datt, traits, colony
urban_colony_p = function(MEs) {
  treat_group = factor(paste(traits$lowpH, traits$hightemp))
  adj = apply(MEs, 2, function(y) {
    fit = lm(y ~ treat_group + colony)
    b = coef(fit)[grepl("^treat_group", names(coef(fit)))]; b[is.na(b)] = 0
    y - drop(model.matrix(fit)[, names(b), drop = FALSE] %*% b)
  })
  colony_means = apply(adj, 2, function(y) tapply(y, colony, mean))
  colony_urban = tapply(traits$urban, colony, `[`, 1)
  p = apply(colony_means, 2, function(y) cor.test(y, colony_urban)$p.value)
  p.adjust(p, method = "BH")
}
thresholds = c(0, 0.15, 0.20, 0.25, 0.30, 0.35)
merge_sens = do.call(rbind, lapply(thresholds, function(thr) {
  MEs_t = if (thr == 0) moduleEigengenes(datt, dynamicColors)$eigengenes else
    mergeCloseModules(datt, dynamicColors, cutHeight = thr, verbose = 0)$newMEs
  MEs_t = orderMEs(MEs_t)
  MEs_t = MEs_t[, colnames(MEs_t) != "MEgrey", drop = FALSE]
  P = corAndPvalue(MEs_t, traits, use = "p")$p
  sig = colSums(P < 0.05, na.rm = TRUE)
  data.frame(threshold = thr, modules = ncol(MEs_t), t(sig), urban_colony_BH = sum(urban_colony_p(MEs_t) < 0.05),
             check.names = FALSE)
}))
merge_sens
write.csv(merge_sens, file = "../../outputs/transcriptomics/ofav/wgcna/merge threshold sensitivity.csv", row.names = FALSE)
```


# MODULE MEMBERSHIP SCATTERPLOTS

```{r module-membership-scatterplots}
# scatterplots of gene significance (correlation-based) vs kME
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData");
load(file = "../../outputs/transcriptomics/ofav/wgcna/moduleTraitSig.RData") # sig_by_trait, modules_to_plot
traits
table(moduleColors)

# one figure per trait that has significantly correlated modules (p < 0.05; sig_by_trait from MODULE CORRELATIONS);
# to plot a single trait instead, set e.g. scatter_traits = "hightemp"
scatter_traits = names(sig_by_trait)[lengths(sig_by_trait) > 0]
scatter_traits

nGenes = ncol(datt);
nSamples = nrow(datt);
# names (colors) of the modules
modNames = substring(names(MEs), 3)
geneModuleMembership = as.data.frame(signedKME(datt, MEs));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));
names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

mm_plots = list()
for (whichTrait in scatter_traits) {
  selTrait = as.data.frame(traits[,whichTrait]);
  names(selTrait) = whichTrait
  # gene significance; p-values use the number of samples with a non-missing trait value
  GS = corAndPvalue(datt, selTrait, use = "p")
  geneTraitSignificance = as.data.frame(GS$cor);
  names(geneTraitSignificance) = paste("GS.", names(selTrait), sep="");
  # modules significantly correlated (p < 0.05) with this trait; one panel per module
  moduleCols = sig_by_trait[[whichTrait]]
  mm_plots[[whichTrait]] = save_plot(paste0("../../outputs/transcriptomics/ofav/wgcna/module membership ", whichTrait, ".pdf"),
                                     width = 4 * length(moduleCols), height = 4.5, {
    par(mfrow=c(1, length(moduleCols)))
    for (module in moduleCols) {
      column = match(module, modNames);
      moduleGenes = moduleColors==module;
      verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                         abs(geneTraitSignificance[moduleGenes, 1]),
                         xlab = paste(module,"module membership"),
                         ylab = paste("GS for", whichTrait),
                         col = "grey50",mgp=c(2.3,1,0))
    }
  })
}
if (length(mm_plots) && isTRUE(getOption("knitr.in.progress"))) knitr::include_graphics(unlist(mm_plots), rel_path = FALSE, dpi = 150)
```


# EIGENGENE SANITY CHECK

```{r eigengene-sanity-check}
# eigengene-heatmap plot (sanity check - is the whole module driven by just one crazy sample?)
# note: this part does not make much sense for unsigned modules
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData");
load(file = "../../outputs/transcriptomics/ofav/wgcna/moduleTraitSig.RData") # modules_to_plot

datME=MEs
datExpr=datt
# one plot per module in modules_to_plot (modules significantly correlated with any trait, unless set manually)
sanity_plots = list()
for (which.module in modules_to_plot) {
  ME=datME[, paste("ME",which.module, sep="")]
  sanity_plots[[which.module]] = save_plot(paste0("../../outputs/transcriptomics/ofav/wgcna/module eigengene ", which.module, ".pdf"), width = 10, height = 7, {
    par(mfrow=c(2,1), mar=c(0.3, 5.5, 3, 2))
    plotMat(t(scale(datExpr[,moduleColors==which.module ]) ),
            nrgcols=30,rlabels=F,rcols=which.module,
            main=which.module, cex.main=2)
    par(mar=c(5, 4.2, 0, 0.7))
    barplot(ME, col=which.module, main="", cex.main=2,
            ylab="eigengene expression",xlab="sample")
  })
  cat(which.module, ":", sum(moduleColors==which.module), "genes\n") # number of genes in module
}
if (length(sanity_plots) && isTRUE(getOption("knitr.in.progress"))) knitr::include_graphics(unlist(sanity_plots), rel_path = FALSE, dpi = 150)

# If individual samples appear to be driving expression of significant modules, they are likely outliers
# count the array numbers and go back to the outlier detection section to remove outliers and rerun all code,
# or drop that module from modules_to_plot in the module correlations section
```


# GO/KOG EXPORT

```{r go-kog-export}
# saving selected modules for GO and KOG analysis (two-parts: Fisher test, MWU test within-module)
library(WGCNA)
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData") # moduleColors, MEs
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData") # datt, traits, colony
load(file = "../../outputs/transcriptomics/ofav/deseq2/data4wgcna.RData") # vsd table
load(file = "../../outputs/transcriptomics/ofav/wgcna/moduleTraitSig.RData") # modules_to_plot

# calculating modul memberships for all genes for all modules
allkME =as.data.frame(signedKME(datt, MEs)) 
names(allkME)=gsub("kME","",names(allkME))

# one file per module in modules_to_plot
for (whichModule in modules_to_plot) {
  print(table(moduleColors==whichModule)) # how many genes are in it?
  # Saving data for Fisher-MWU combo test (GO_MWU)
  inModuleBinary=as.numeric(moduleColors==whichModule)
  combo=data.frame("gene"=colnames(datt),"Fish_kME"=allkME[,whichModule]*inModuleBinary) # genes in datt order (matches allkME)
  write.csv(combo,file=paste0("../../outputs/transcriptomics/ofav/wgcna/",whichModule,".csv"),row.names=F,quote=F)
}
```


# HEATMAPS

```{r heatmaps}
# plotting heatmap for named top-kME genes
library(WGCNA)
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
load(file = "../../outputs/transcriptomics/ofav/deseq2/data4wgcna.RData") 
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData");
load(file = "../../outputs/transcriptomics/ofav/wgcna/moduleTraitSig.RData") # modules_to_plot
allkME =as.data.frame(signedKME(datt, MEs))
vsd.wg = vsd.wg[colnames(datt), ] # same genes, in the same order, as allkME
gg=read.delim(file="../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",sep="\t",header=FALSE,quote="")
library(pheatmap)

top=30 # number of named top-kME genes to plot

contrasting = colorRampPalette(rev(c("chocolate1","#FEE090","grey10", "cyan3","cyan")))(100)
contrasting2 = colorRampPalette(rev(c("chocolate1","chocolate1","#FEE090","grey10", "cyan3","cyan")))(100)
contrasting3 = colorRampPalette(rev(c("chocolate1","#FEE090","grey10", "cyan3","cyan","cyan")))(100)

datME=MEs
datExpr=datt
# one heatmap per module in modules_to_plot
hub_heatmaps <- list()
for (whichModule in modules_to_plot) {
  modcol=paste("kME",whichModule,sep="")
  sorted=vsd.wg[order(allkME[,modcol],decreasing=T),]
  # selection top N names genes, attaching gene names
  gnames=c();counts=0;hubs=c()
  for(i in 1:length(sorted[,1])) {
  	if (row.names(sorted)[i] %in% gg[,1]) { 
  		counts=counts+1
  		gn=gg[gg[,1]==row.names(sorted)[i],2]
  		if (nchar(gn) > 45) gn=paste0(substr(gn,1,42),"...") # shortened for the row labels
  		gn=paste(gn,row.names(sorted)[i],sep=".")
  		if (gn %in% gnames) {
  			gn=paste(gn,counts,sep=".")
  		}
  		gnames=append(gnames,gn) 
  		hubs=data.frame(rbind(hubs,sorted[i,]))
  		if (counts==top) {break}
  	}
  } 
  row.names(hubs)=gnames
  # gaps between treatment groups, as in the DESeq2 heatmaps (columns are in design order, grouped by treatment;
  # the treatment is the last part of each sample name, e.g. Ssid_001.Emerald.CC)
  col_treat <- sub(".*\\.", "", colnames(hubs))
  stopifnot(!anyDuplicated(rle(col_treat)$values)) # each treatment is one contiguous block
  treat_gaps <- head(cumsum(rle(col_treat)$lengths), -1)

  # square 9-pt cells, page sized to fit (as in the DESeq2 heatmaps)
  hub_heatmaps[[whichModule]] <- save_plot(file=paste0("../../outputs/transcriptomics/ofav/wgcna/heatmap top", top, " ", whichModule, ".pdf"),
                                           width=(ncol(hubs)*9 + 400)/72, height=(nrow(hubs)*9 + 140)/72, {
    pheatmap(hubs,scale="row",col=contrasting2,border_color=NA,treeheight_col=0,cluster_rows=F,cluster_cols=F,
             cellwidth=9,cellheight=9,fontsize_row=7,fontsize_col=6,main=whichModule,gaps_col=treat_gaps)
  })
}
# show the module heatmaps in the knitted HTML
if (length(hub_heatmaps) && isTRUE(getOption("knitr.in.progress"))) knitr::include_graphics(unlist(hub_heatmaps), rel_path = FALSE, dpi = 150)
```


# HUB GENES

```{r hub-genes}
library(WGCNA)
library(tidyverse)
load(file = "../../outputs/transcriptomics/ofav/wgcna/networkdata_signed.RData")
load(file = "../../outputs/transcriptomics/ofav/deseq2/data4wgcna.RData") 
load(file = "../../outputs/transcriptomics/ofav/wgcna/wgcnaData.RData");
allkME =as.data.frame(signedKME(datt, MEs))

# set the power based on the soft threshold used in making modules
hubgenes <- chooseTopHubInEachModule(datt, moduleColors, omitColors = "grey", 
                                     power = s.th, # soft-threshold power saved with the network
                                     type = "signed")
hubgenes <-data.frame(hubgenes)
hubgenes <- tibble::rownames_to_column(hubgenes, "module")
hubgenes

hubgenes %>%
  rename("gene" = 
           hubgenes) %>%
  left_join(read.table(file = "../../../Orbicella-faveolata-annotated-transcriptome/Ofaveolata_iso2geneName.tab",
                       sep = "\t",
                       quote="", fill=FALSE) %>%
              mutate(gene = V1,
                     annot = V2) %>%
              dplyr::select(-V1, -V2), by = "gene") -> hubgenes
hubgenes

write.csv(hubgenes, file="../../outputs/transcriptomics/ofav/wgcna/hubgenes.csv")
```
