---
title: "Coral Growth (Buoyant Weight) — Linear Mixed-Effects Analysis"
author: "Michael Studivan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# -----------------------------------------------------------------------------
# Global chunk options
# echo = TRUE:     show code in rendered output
# warning = FALSE: suppress warnings in rendered document
#                  (warnings still print to console during knitting)
# message = FALSE: suppress package loading messages
# -----------------------------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 300
)
```


# Packages

```{r packages}
library(ggplot2)
library(ggpubr)
library(stringr)
library(tidyr)
library(dplyr)
library(lme4)
library(lmerTest)
library(emmeans)
library(multcomp)
library(MuMIn)
```


# Data import

```{r data-import}
bw <- read.csv(file="../../data/growth/urban buoyant weight.csv", head=T)

# growth rates are calculated here from dry skeletal mass (mass1, mass2; g) and surface area (cm2), annualized from
# the actual interval: buoyant weights were measured on 26 Feb (initial) and 3 Apr 2021 (post-OA), 36 days apart.
# (The master spreadsheet annualized by x12, assuming a one-month interval; those columns were removed from the data file.)
bw_interval_days <- as.numeric(as.Date("2021-04-03") - as.Date("2021-02-26"))
bw %>%
  mutate(growth = (mass2 - mass1) / surfacearea / bw_interval_days * 365) -> bw # g cm-2 yr-1

# dead skeleton controls: mean control growth for each pH treatment (both species) is subtracted from every fragment
bw %>%
  filter(species == "control") %>%
  group_by(pH) %>%
  summarise(control_growth = mean(growth), n_control = n(), .groups = "drop") -> bw_controls
bw_controls

bw %>%
  left_join(dplyr::select(bw_controls, pH, control_growth), by = "pH") %>%
  mutate(growthadj = growth - control_growth) %>%
  dplyr::select(-control_growth) -> bw

bw$tank <- as.factor(bw$tank)
bw$pH <- as.factor(bw$pH)
bw$site=factor(bw$site, levels=c("Emerald", "Rainbow", "Star", "MacN", "control")) 
bw$genotype <- as.factor(bw$genotype)

str(bw)
head(bw)
```


# Data normality and assumption testing

```{r data-normality-and-assumption-testing}
bw_subset <- subset(bw, site!="control")

bw_ofav <- bw_subset[bw_subset$species == "Ofav",] # filtering dataset by species
bw_ssid <- bw_subset[bw_subset$species == "Ssid",]
str(bw_ofav)
str(bw_ssid)
```


## Outlier removal

```{r outliers}
# identifying outliers with IQR > 1.5
outliers_ofav <- boxplot(bw_ofav$growthadj, plot=FALSE)$out # 1 outlier
bw_ofav <- bw_ofav[!bw_ofav$growthadj %in% outliers_ofav,] # safe when there are no outliers

outliers_ssid <- boxplot(bw_ssid$growthadj, plot=FALSE)$out # no outliers
```


## Raw data normality

```{r raw-normality}
# Shapiro test, p-values below 0.05 indicate violations of normality assumptions
shapiro.test(bw_ofav$growthadj) # meets normality assumptions

shapiro.test(bw_ssid$growthadj) # violates normality assumptions

pdf("../../outputs/growth/urban bw normality.pdf")
par(mfrow=c(2,2))
hist(bw_ofav$growthadj)
hist(bw_ssid$growthadj)
qqnorm(bw_ofav$growthadj)
qqline(bw_ofav$growthadj)
qqnorm(bw_ssid$growthadj)
qqline(bw_ssid$growthadj)
dev.off() # despite violation of SW assumptions in ssid, both species data appear normally distributed
```


# Linear mixed-effects models


## *O. faveolata*


### Model fitting and interaction test

```{r models-ofav}
# full model, tank nested within pH, genotype nested within site
lm_ofav <- lmer(growthadj ~ pH * site + 
                  (1 | pH:tank) + 
                  (1 | site:genotype),
                data = bw_ofav, REML = TRUE)

# reduced model, no interaction of pH and site
lm_ofav_noint <- lmer(growthadj ~ pH + site + 
                        (1 | pH:tank) + 
                        (1 | site:genotype),
                      data = bw_ofav, REML = TRUE)

# LRT: interaction effect
lrt_int_ofav <- anova(update(lm_ofav, REML = FALSE), update(lm_ofav_noint, REML = FALSE))
lrt_int_ofav # nonsignificant p value, but AIC lower for noint model, so proceed without interaction terms
```


### Random effect tests

```{r random-effects-ofav}
# reduced model, dropped genotype as factor
lm_ofav_noG <- lmer(growthadj ~ pH + site + 
                      (1 | pH:tank),
                    data = bw_ofav, REML = TRUE)

# reduced model, dropped tank as factor
lm_ofav_noT <- lmer(growthadj ~ pH + site + 
                      (1 | site:genotype),
                    data = bw_ofav, REML = TRUE)

# LRT: genotype effect
lrt_geno_ofav <- anova(update(lm_ofav_noint, REML = FALSE), update(lm_ofav_noG, REML = FALSE))
lrt_geno_ofav # p value significant, and AIC much lower for noint model, so proceed with genotype term

# LRT: tank effect
lrt_tank_ofav <- anova(update(lm_ofav_noint, REML = FALSE), update(lm_ofav_noT, REML = FALSE))
lrt_tank_ofav # p value significant, and AIC much lower for noint model, so proceed with tank
```


### Tank-level check of pH effect

```{r tank-check-ofav}
# sanity check: is there a pH effect when looking at growth just across tanks?
tank_means_ofav <- aggregate(growthadj ~ pH + tank, bw_ofav, mean)
summary(lm(growthadj ~ pH, data = tank_means_ofav)) # nonsignificant; matches mixed-model pH test, confirming tank is the unit of replication for pH
```


### Variance components and R²

```{r variance-ofav}
# Random effects variance components
# NOTE: pct_nonfixed_var = % of random + residual variance only (fixed-effect variance excluded)
vc_ofav <- as.data.frame(VarCorr(lm_ofav_noint))
total_var_ofav <- sum(vc_ofav$vcov)
vc_ofav$pct_nonfixed_var <- round(100 * vc_ofav$vcov / total_var_ofav, 2)
colnames(vc_ofav)[colnames(vc_ofav) == "grp"] <- "component"

# Fixed effects R²
r2_ofav <- r.squaredGLMM(lm_ofav_noint)

# Individual fixed effect contributions
lm_ofav_ml      <- update(lm_ofav_noint, REML = FALSE)
lm_ofav_no_pH   <- update(lm_ofav_ml, . ~ . - pH)
lm_ofav_no_site <- update(lm_ofav_ml, . ~ . - site)

r2_ofav_full    <- r.squaredGLMM(lm_ofav_ml)[, "R2m"]
delta_pH_ofav   <- round(100 * (r2_ofav_full - r.squaredGLMM(lm_ofav_no_pH)[, "R2m"]), 2)
delta_site_ofav <- round(100 * (r2_ofav_full - r.squaredGLMM(lm_ofav_no_site)[, "R2m"]), 2)
```


### Model summaries

```{r summary-ofav}
# --- Console output ---
summary(lm_ofav_noint)
anova(lm_ofav_noint)
print(vc_ofav[, c("component", "vcov", "pct_nonfixed_var")])
cat("Marginal R² (fixed effects):", round(100 * r2_ofav[, "R2m"], 2), "%\n")
cat("Conditional R² (fixed + random):", round(100 * r2_ofav[, "R2c"], 2), "%\n")
cat("Delta R² pH:", delta_pH_ofav, "%\n")
cat("Delta R² site:", delta_site_ofav, "%\n")
```


### Export model outputs

```{r export-ofav}
# --- txt output ---
capture.output({
  cat("=== LRT: pH x Site Interaction (Fixed) ===\n")
  print(lrt_int_ofav)
  
  cat("\n=== LRT: Tank Effect (pH:tank) ===\n")
  print(lrt_tank_ofav)
  
  cat("\n=== LRT: Genotype Effect (site:genotype) ===\n")
  print(lrt_geno_ofav)
  
  cat("\n=== ANOVA (Fixed Effects) ===\n")
  print(anova(lm_ofav_noint))
  
  cat("\n=== Variance Components (Random Effects) ===\n")
  cat("pct_nonfixed_var = % of random + residual variance (excludes fixed-effect variance)\n")
  print(vc_ofav[, c("component", "vcov", "pct_nonfixed_var")])
  
  cat("\n=== R² (Fixed Effects) ===\n")
  cat("Marginal R² (fixed effects only):", round(100 * r2_ofav[, "R2m"], 2), "%\n")
  cat("Conditional R² (fixed + random):", round(100 * r2_ofav[, "R2c"], 2), "%\n")
  cat("Delta R² pH:", delta_pH_ofav, "%\n")
  cat("Delta R² site:", delta_site_ofav, "%\n")
  
}, file = "../../outputs/growth/urban bw ofav lme.txt")
```


### Pairwise site comparisons

```{r pairwise-ofav}
# pairwise site tests
emm_ofav <- emmeans(lm_ofav_noint, ~ site) # no pH x site interaction in model, so site contrasts are identical across pH levels
pairs(emm_ofav, adjust = "tukey")
capture.output(pairs(emm_ofav, adjust = "tukey"), file = "../../outputs/growth/urban bw ofav pairwise.txt")

# Create letters indicating significant differences for plot
cld_ofav <- cld(emm_ofav, adjust = "tukey", Letters = letters, alpha = 0.05)
cld_ofav
```


## *S. siderea*


### Model fitting and interaction test

```{r models-ssid}
# full model, tank nested within pH, genotype nested within site
lm_ssid <- lmer(growthadj ~ pH * site + 
                  (1 | pH:tank) + 
                  (1 | site:genotype),
                data = bw_ssid, REML = TRUE)

# reduced model, no interaction of pH and site
lm_ssid_noint <- lmer(growthadj ~ pH + site + 
                        (1 | pH:tank) + 
                        (1 | site:genotype),
                      data = bw_ssid, REML = TRUE)

# LRT: interaction effect
lrt_int_ssid <- anova(update(lm_ssid, REML = FALSE), update(lm_ssid_noint, REML = FALSE))
lrt_int_ssid # nonsignificant p value, and marginal reduction in AIC with interaction, so proceed without interaction terms for consistency with ofav
```


### Random effect tests

```{r random-effects-ssid}
# reduced model, dropped genotype as factor
lm_ssid_noG <- lmer(growthadj ~ pH + site + 
                      (1 | pH:tank),
                    data = bw_ssid, REML = TRUE)

# reduced model, dropped tank as factor
lm_ssid_noT <- lmer(growthadj ~ pH + site + 
                      (1 | site:genotype),
                    data = bw_ssid, REML = TRUE)

# LRT: genotype effect
lrt_geno_ssid <- anova(update(lm_ssid_noint, REML = FALSE), update(lm_ssid_noG, REML = FALSE))
lrt_geno_ssid # marginally significant p value, and AIC lower for noint model, so proceed with genotype term

# LRT: tank effect
lrt_tank_ssid <- anova(update(lm_ssid_noint, REML = FALSE), update(lm_ssid_noT, REML = FALSE))
lrt_tank_ssid # p value significant, and AIC much lower for noint model, so proceed with tank
```


### Tank-level check of pH effect

```{r tank-check-ssid}
# sanity check: is there a pH effect when looking at growth just across tanks?
tank_means_ssid <- aggregate(growthadj ~ pH + tank, bw_ssid, mean)
summary(lm(growthadj ~ pH, data = tank_means_ssid)) # nonsignificant; matches mixed-model pH test, confirming tank is the unit of replication for pH
```


### Variance components and R²

```{r variance-ssid}
# Random effects variance components
# NOTE: pct_nonfixed_var = % of random + residual variance only (fixed-effect variance excluded)
vc_ssid <- as.data.frame(VarCorr(lm_ssid_noint))
total_var_ssid <- sum(vc_ssid$vcov)
vc_ssid$pct_nonfixed_var <- round(100 * vc_ssid$vcov / total_var_ssid, 2)
colnames(vc_ssid)[colnames(vc_ssid) == "grp"] <- "component"

# Fixed effects R²
r2_ssid <- r.squaredGLMM(lm_ssid_noint)

# Individual fixed effect contributions
lm_ssid_ml      <- update(lm_ssid_noint, REML = FALSE)
lm_ssid_no_pH   <- update(lm_ssid_ml, . ~ . - pH)
lm_ssid_no_site <- update(lm_ssid_ml, . ~ . - site)

r2_ssid_full    <- r.squaredGLMM(lm_ssid_ml)[, "R2m"]
delta_pH_ssid   <- round(100 * (r2_ssid_full - r.squaredGLMM(lm_ssid_no_pH)[, "R2m"]), 2)
delta_site_ssid <- round(100 * (r2_ssid_full - r.squaredGLMM(lm_ssid_no_site)[, "R2m"]), 2)
```


### Model summaries

```{r summary-ssid}
# --- Console output ---
summary(lm_ssid_noint)
anova(lm_ssid_noint)
print(vc_ssid[, c("component", "vcov", "pct_nonfixed_var")])
cat("Marginal R² (fixed effects):", round(100 * r2_ssid[, "R2m"], 2), "%\n")
cat("Conditional R² (fixed + random):", round(100 * r2_ssid[, "R2c"], 2), "%\n")
cat("Delta R² pH:", delta_pH_ssid, "%\n")
cat("Delta R² site:", delta_site_ssid, "%\n")
```


### Export model outputs

```{r export-ssid}
# --- txt output ---
capture.output({
  cat("=== LRT: pH x Site Interaction (Fixed) ===\n")
  print(lrt_int_ssid)
  
  cat("\n=== LRT: Tank Effect (pH:tank) ===\n")
  print(lrt_tank_ssid)
  
  cat("\n=== LRT: Genotype Effect (site:genotype) ===\n")
  print(lrt_geno_ssid)
  
  cat("\n=== ANOVA (Fixed Effects) ===\n")
  print(anova(lm_ssid_noint))
  
  cat("\n=== Variance Components (Random Effects) ===\n")
  cat("pct_nonfixed_var = % of random + residual variance (excludes fixed-effect variance)\n")
  print(vc_ssid[, c("component", "vcov", "pct_nonfixed_var")])
  
  cat("\n=== R² (Fixed Effects) ===\n")
  cat("Marginal R² (fixed effects only):", round(100 * r2_ssid[, "R2m"], 2), "%\n")
  cat("Conditional R² (fixed + random):", round(100 * r2_ssid[, "R2c"], 2), "%\n")
  cat("Delta R² pH:", delta_pH_ssid, "%\n")
  cat("Delta R² site:", delta_site_ssid, "%\n")
  
}, file = "../../outputs/growth/urban bw ssid lme.txt")
```


### Pairwise site comparisons

```{r pairwise-ssid}
# pairwise site tests
# emm_ssid <- emmeans(lm_ssid_noint, ~ site)
# pairs(emm_ssid, adjust = "tukey")
# capture.output(pairs(emm_ssid, adjust = "tukey"), file = "../../outputs/growth/urban bw ssid pairwise.txt")

# Create letters indicating significant differences for plot
# cld_ssid <- cld(emm_ssid, adjust = "tukey", Letters = letters, alpha = 0.05)
# cld_ssid
```


# Model residual diagnostics

```{r model-residual-diagnostics}
# normality of the final models' residuals and random effects (the assumptions lmer actually makes),
# rather than of the raw response
resid_ofav <- resid(lm_ofav_noint)
resid_ssid <- resid(lm_ssid_noint)

# Shapiro test on model residuals, p-values below 0.05 indicate violations of normality assumptions
shapiro.test(resid_ofav)
shapiro.test(resid_ssid)

# random effect estimates (BLUPs), few levels so interpret visually
re_ofav <- ranef(lm_ofav_noint)
re_ssid <- ranef(lm_ssid_noint)

pdf("../../outputs/growth/urban bw model residuals.pdf", width = 10, height = 8)
par(mfrow = c(2, 4))
for (sp in c("ofav", "ssid")) {
  m <- get(paste0("lm_", sp, "_noint"))
  r <- resid(m)
  re <- ranef(m)
  # residual normality
  qqnorm(r, main = paste(sp, "residuals Q-Q")); qqline(r)
  # homogeneity of variance
  plot(fitted(m), r, xlab = "Fitted values", ylab = "Residuals", main = paste(sp, "residuals vs fitted"))
  abline(h = 0, lty = 2)
  # random effect normality
  qqnorm(re$`site:genotype`[, 1], main = paste(sp, "genotype effects Q-Q")); qqline(re$`site:genotype`[, 1])
  qqnorm(re$`pH:tank`[, 1], main = paste(sp, "tank effects Q-Q")); qqline(re$`pH:tank`[, 1])
}
dev.off()

capture.output({
  cat("=== Shapiro-Wilk test on model residuals ===\n")
  cat("\nO. faveolata (lm_ofav_noint):\n")
  print(shapiro.test(resid_ofav))
  cat("S. siderea (lm_ssid_noint):\n")
  print(shapiro.test(resid_ssid))
}, file = "../../outputs/growth/urban bw model residuals.txt")
```


# Percent change in growth

```{r percent-change-in-growth}
# % change in mean growth under low pH relative to control pH: (low - control) / control * 100
# raw means from outlier-filtered data (bw_ofav, bw_ssid)
bw_models <- rbind(bw_ofav, bw_ssid)

growth_means <- bw_models %>%
  group_by(species, site, pH) %>%
  summarise(mean = mean(growthadj, na.rm = TRUE),
            sd = sd(growthadj, na.rm = TRUE),
            n = sum(!is.na(growthadj)),
            .groups = "drop")

# by species and site
pct_change_site <- growth_means %>%
  pivot_wider(names_from = pH, values_from = c(mean, sd, n)) %>%
  mutate(pct_change = round(100 * (`mean_low pH` - `mean_control pH`) / `mean_control pH`, 2))

# by species, pooled across sites
pct_change_species <- bw_models %>%
  group_by(species, pH) %>%
  summarise(mean = mean(growthadj, na.rm = TRUE),
            sd = sd(growthadj, na.rm = TRUE),
            n = sum(!is.na(growthadj)),
            .groups = "drop") %>%
  pivot_wider(names_from = pH, values_from = c(mean, sd, n)) %>%
  mutate(site = "all sites",
         pct_change = round(100 * (`mean_low pH` - `mean_control pH`) / `mean_control pH`, 2))

pct_change <- bind_rows(pct_change_site, pct_change_species) %>%
  mutate(site = factor(site, levels = c("Emerald", "Rainbow", "Star", "MacN", "all sites"))) %>%
  arrange(species, site)
pct_change

write.csv(pct_change, file = "../../outputs/growth/urban bw pct change.csv", row.names = FALSE)
```


# Plots

```{r plots}
fill.color<-c("#018571","#80cdc1","#dfc27d","#a6611a","grey") # custom color palette

# creating dummy variables for stats labels on plot
stats_ofav_ph <- data.frame(pH = "control pH", growthadj = 0.6, lab = "pH: F1,82 = 0.02, p = 0.9")
stats_ofav_site <- data.frame(pH = "control pH", growthadj = 0.6, lab = "site: F3,82 = 4.3, p = 0.02*")

stats_ssid_ph <- data.frame(pH = "control pH", growthadj = 0.6, lab = "pH: F1,79 = 0.7, p = 0.5")
stats_ssid_site <- data.frame(pH = "control pH", growthadj = 0.6, lab = "site: F3,79 = 2.8, p = 0.07")

ofav <- ggboxplot(bw_ofav,
                  x = "site",
                  y = "growthadj",
                  fill = "site",
                  width = 0.7,
                  size = 0.75,
                  title = "Orbicella faveolata",
                  legend = "none") +
  facet_grid(~pH) +
  xlab(NULL) +
  ylab("Growth Rate (g cm-2 yr-1)") +
  scale_fill_manual(values = fill.color) +
  geom_hline(yintercept=0, linetype="dashed", color = "black") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_blank()) +
  geom_text(data = stats_ofav_ph,label = stats_ofav_ph$lab, aes(x=2, y=0.7)) +
  geom_text(data = stats_ofav_site,label = stats_ofav_site$lab, aes(x=2, y=0.65)) +
  geom_text(data=cld_ofav, aes(x = site, y=-0.13, label=.group)) +
  ylim(-0.13, 0.7)
ofav

ssid <- ggboxplot(bw_ssid,
                  x = "site",
                  y = "growthadj",
                  fill = "site",
                  width = 0.7,
                  size = 0.75,
                  title = "Siderastrea siderea",
                  legend = "none") +
  facet_grid(~pH) +
  xlab(NULL) +
  ylab("Growth Rate (g cm-2 yr-1)") +
  scale_fill_manual(values = fill.color) +
  geom_hline(yintercept=0, linetype="dashed", color = "black") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(data = stats_ssid_ph,label = stats_ssid_ph$lab, aes(x=2, y=0.7)) +
  geom_text(data = stats_ssid_site,label = stats_ssid_site$lab, aes(x=2, y=0.65)) +
  # geom_text(data=cld_ssid, aes(x = site, y=-0.13, label=.group)) +
  ylim(-0.13, 0.7)
ssid

# both panels
bw_species<-ggarrange(ofav,
                      ssid,
                      heights = c(3.9,4),
                      widths = c(4,4),
                      ncol = 1,
                      nrow = 2)
bw_species

ggsave("../../outputs/growth/urban bw.pdf", bw_species, width=8, height=8,dpi = 300)
```
