---
title: "Coral Growth (Buoyant Weight) — Linear Mixed-Effects Analysis"
author: "Michael Studivan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# -----------------------------------------------------------------------------
# Global chunk options
# echo = TRUE:     show code in rendered output
# warning = FALSE: suppress warnings in rendered document
#                  (warnings still print to console during knitting)
# message = FALSE: suppress package loading messages
# -----------------------------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 300
)

# save_plot(): writes a figure to PDF (outputs folder) and shows the same figure in the knitted HTML.
# The plotting code is run once for the PDF and once for a PNG copy at the same size, which is embedded in
# the HTML; when run interactively (not knitting) the plot is also drawn in the plot pane.
save_plot <- function(file, width, height, expr) {
  expr <- substitute(expr); env <- parent.frame()
  pdf(file = file, width = width, height = height)
  tryCatch(eval(expr, env), finally = dev.off())
  if (!isTRUE(getOption("knitr.in.progress"))) { # interactive: also draw in the plot pane (skipped if the pane is too small)
    ok <- try(eval(expr, env), silent = TRUE)
    if (inherits(ok, "try-error")) try(dev.off(), silent = TRUE) # reset a device left in a broken state
    return(invisible(file)) }
  png_file <- file.path(tempdir(), sub("\\.pdf$", ".png", basename(file)))
  png(png_file, width = width, height = height, units = "in", res = 150)
  tryCatch(eval(expr, env), finally = dev.off())
  knitr::include_graphics(png_file, rel_path = FALSE, dpi = 150) # displayed at its true size
}
```


# Packages

```{r packages}
library(ggplot2)
library(ggpubr)
library(stringr)
library(tidyr)
library(dplyr)
library(lme4)
library(lmerTest)
library(emmeans)
library(multcomp)
library(MuMIn)
```


# Data import

```{r data-import}
bw <- read.csv(file="../../data/growth/urban buoyant weight.csv", head=T)

# fragments removed from the survivorship analysis are excluded from all analyses (reasons in the survivorship
# data file's `exclude` column): 31 O. faveolata with partial mortality unrelated to heat stress (mostly a ciliate
# infection; they grew less than the other fragments in their tanks even before tissue loss was recorded) and one
# S. siderea (Ms3-1) with no health observations. Rows stay in the data file.
surv_exclude <- read.csv("../../data/survivorship/urban survivorship.csv") %>%
  filter(exclude != "")
bw %>% filter(sampleID %in% surv_exclude$sample) %>% count(species, treatment) # fragments excluded
bw <- bw %>% filter(!sampleID %in% surv_exclude$sample)

# growth rates are calculated here from dry skeletal mass (mass1, mass2; g) and surface area (cm2), annualized from
# the actual interval: buoyant weights were measured on 26 Feb (initial) and 3 Apr 2021 (post-OA), 36 days apart.
# (The master spreadsheet annualized by x12, assuming a one-month interval; those columns were removed from the data file.)
bw_interval_days <- as.numeric(as.Date("2021-04-03") - as.Date("2021-02-26"))
bw %>%
  mutate(growth = (mass2 - mass1) / surfacearea / bw_interval_days * 365) -> bw # g cm-2 yr-1

# dead skeleton controls: mean control growth for each pH treatment (both species) is subtracted from every fragment
bw %>%
  filter(species == "control") %>%
  group_by(pH) %>%
  summarise(control_growth = mean(growth), n_control = n(), .groups = "drop") -> bw_controls
bw_controls

bw %>%
  left_join(dplyr::select(bw_controls, pH, control_growth), by = "pH") %>%
  mutate(growthadj = growth - control_growth) %>%
  dplyr::select(-control_growth) -> bw

bw$tank <- as.factor(bw$tank)
bw$pH <- as.factor(bw$pH)
bw$site=factor(bw$site, levels=c("Emerald", "Rainbow", "Star", "MacN", "control")) 
bw$genotype <- as.factor(bw$genotype)

str(bw)
head(bw)
```


# Data normality and assumption testing

```{r data-normality-and-assumption-testing}
bw_subset <- subset(bw, site!="control")

bw_ofav <- bw_subset[bw_subset$species == "Ofav",] # filtering dataset by species
bw_ssid <- bw_subset[bw_subset$species == "Ssid",]
str(bw_ofav)
str(bw_ssid)
```


## Outlier removal

```{r outliers}
# identifying outliers with IQR > 1.5
outliers_ofav <- boxplot(bw_ofav$growthadj, plot=FALSE)$out # 1 outlier
bw_ofav <- bw_ofav[!bw_ofav$growthadj %in% outliers_ofav,] # safe when there are no outliers

outliers_ssid <- boxplot(bw_ssid$growthadj, plot=FALSE)$out # no outliers
```


## Raw data normality

```{r raw-normality}
# Shapiro test, p-values below 0.05 indicate violations of normality assumptions
shapiro.test(bw_ofav$growthadj) # meets normality assumptions

shapiro.test(bw_ssid$growthadj) # violates normality assumptions

save_plot("../../outputs/growth/urban bw normality.pdf", width = 7, height = 7, {
par(mfrow=c(2,2))
hist(bw_ofav$growthadj)
hist(bw_ssid$growthadj)
qqnorm(bw_ofav$growthadj)
qqline(bw_ofav$growthadj)
qqnorm(bw_ssid$growthadj)
qqline(bw_ssid$growthadj)
}) # despite violation of SW assumptions in ssid, both species data appear normally distributed
```


# Linear mixed-effects models


## *O. faveolata*


### Model fitting and interaction test

```{r models-ofav}
# full model, tank nested within pH, genotype nested within site
lm_ofav <- lmer(growthadj ~ pH * site + 
                  (1 | pH:tank) + 
                  (1 | site:genotype),
                data = bw_ofav, REML = TRUE)

# reduced model, no interaction of pH and site
lm_ofav_noint <- lmer(growthadj ~ pH + site + 
                        (1 | pH:tank) + 
                        (1 | site:genotype),
                      data = bw_ofav, REML = TRUE)

# LRT: interaction effect
lrt_int_ofav <- anova(update(lm_ofav, REML = FALSE), update(lm_ofav_noint, REML = FALSE))
lrt_int_ofav # nonsignificant p value, but AIC lower for noint model, so proceed without interaction terms
```


### Random effect tests

```{r random-effects-ofav}
# reduced model, dropped genotype as factor
lm_ofav_noG <- lmer(growthadj ~ pH + site + 
                      (1 | pH:tank),
                    data = bw_ofav, REML = TRUE)

# reduced model, dropped tank as factor
lm_ofav_noT <- lmer(growthadj ~ pH + site + 
                      (1 | site:genotype),
                    data = bw_ofav, REML = TRUE)

# LRT: genotype effect
lrt_geno_ofav <- anova(update(lm_ofav_noint, REML = FALSE), update(lm_ofav_noG, REML = FALSE))
lrt_geno_ofav # p value significant, and AIC much lower for noint model, so proceed with genotype term

# LRT: tank effect
lrt_tank_ofav <- anova(update(lm_ofav_noint, REML = FALSE), update(lm_ofav_noT, REML = FALSE))
lrt_tank_ofav # nonsignificant after the survivorship exclusions (the tank variance came from the excluded T1/T4/T8 fragments); tank kept as the unit of pH replication
```


### Tank-level check of pH effect

```{r tank-check-ofav}
# sanity check: is there a pH effect when looking at growth just across tanks?
tank_means_ofav <- aggregate(growthadj ~ pH + tank, bw_ofav, mean)
summary(lm(growthadj ~ pH, data = tank_means_ofav)) # nonsignificant; matches mixed-model pH test, confirming tank is the unit of replication for pH
```


### Variance components and R²

```{r variance-ofav}
# Random effects variance components
# NOTE: pct_nonfixed_var = % of random + residual variance only (fixed-effect variance excluded)
vc_ofav <- as.data.frame(VarCorr(lm_ofav_noint))
total_var_ofav <- sum(vc_ofav$vcov)
vc_ofav$pct_nonfixed_var <- round(100 * vc_ofav$vcov / total_var_ofav, 2)
colnames(vc_ofav)[colnames(vc_ofav) == "grp"] <- "component"

# Fixed effects R²
r2_ofav <- r.squaredGLMM(lm_ofav_noint)

# Individual fixed effect contributions
lm_ofav_ml      <- update(lm_ofav_noint, REML = FALSE)
lm_ofav_no_pH   <- update(lm_ofav_ml, . ~ . - pH)
lm_ofav_no_site <- update(lm_ofav_ml, . ~ . - site)

r2_ofav_full    <- r.squaredGLMM(lm_ofav_ml)[, "R2m"]
delta_pH_ofav   <- round(100 * (r2_ofav_full - r.squaredGLMM(lm_ofav_no_pH)[, "R2m"]), 2)
delta_site_ofav <- round(100 * (r2_ofav_full - r.squaredGLMM(lm_ofav_no_site)[, "R2m"]), 2)
```


### Model summaries

```{r summary-ofav}
# --- Console output ---
summary(lm_ofav_noint)
anova(lm_ofav_noint)
print(vc_ofav[, c("component", "vcov", "pct_nonfixed_var")])
cat("Marginal R² (fixed effects):", round(100 * r2_ofav[, "R2m"], 2), "%\n")
cat("Conditional R² (fixed + random):", round(100 * r2_ofav[, "R2c"], 2), "%\n")
cat("Delta R² pH:", delta_pH_ofav, "%\n")
cat("Delta R² site:", delta_site_ofav, "%\n")
```


### Export model outputs

```{r export-ofav}
# --- txt output ---
capture.output({
  cat("=== LRT: pH x Site Interaction (Fixed) ===\n")
  print(lrt_int_ofav)
  
  cat("\n=== LRT: Tank Effect (pH:tank) ===\n")
  print(lrt_tank_ofav)
  
  cat("\n=== LRT: Genotype Effect (site:genotype) ===\n")
  print(lrt_geno_ofav)
  
  cat("\n=== ANOVA (Fixed Effects) ===\n")
  print(anova(lm_ofav_noint))
  
  cat("\n=== Variance Components (Random Effects) ===\n")
  cat("pct_nonfixed_var = % of random + residual variance (excludes fixed-effect variance)\n")
  print(vc_ofav[, c("component", "vcov", "pct_nonfixed_var")])
  
  cat("\n=== R² (Fixed Effects) ===\n")
  cat("Marginal R² (fixed effects only):", round(100 * r2_ofav[, "R2m"], 2), "%\n")
  cat("Conditional R² (fixed + random):", round(100 * r2_ofav[, "R2c"], 2), "%\n")
  cat("Delta R² pH:", delta_pH_ofav, "%\n")
  cat("Delta R² site:", delta_site_ofav, "%\n")
  
}, file = "../../outputs/growth/urban bw ofav lme.txt")
```


### Pairwise site comparisons

```{r pairwise-ofav}
# pairwise site tests
emm_ofav <- emmeans(lm_ofav_noint, ~ site) # no pH x site interaction in model, so site contrasts are identical across pH levels
pairs(emm_ofav, adjust = "tukey")
capture.output(pairs(emm_ofav, adjust = "tukey"), file = "../../outputs/growth/urban bw ofav pairwise.txt")

# Create letters indicating significant differences for plot
cld_ofav <- cld(emm_ofav, adjust = "tukey", Letters = letters, alpha = 0.05)
cld_ofav
```


## *S. siderea*


### Model fitting and interaction test

```{r models-ssid}
# full model, tank nested within pH, genotype nested within site
lm_ssid <- lmer(growthadj ~ pH * site + 
                  (1 | pH:tank) + 
                  (1 | site:genotype),
                data = bw_ssid, REML = TRUE)

# reduced model, no interaction of pH and site
lm_ssid_noint <- lmer(growthadj ~ pH + site + 
                        (1 | pH:tank) + 
                        (1 | site:genotype),
                      data = bw_ssid, REML = TRUE)

# LRT: interaction effect
lrt_int_ssid <- anova(update(lm_ssid, REML = FALSE), update(lm_ssid_noint, REML = FALSE))
lrt_int_ssid # significant after the survivorship exclusions (Ms3-1; p = 0.005) and AIC lower with the interaction, so the final ssid model keeps the pH x site interaction (lm_ssid)
```


### Random effect tests

```{r random-effects-ssid}
# reduced model, dropped genotype as factor
lm_ssid_noG <- lmer(growthadj ~ pH * site + 
                      (1 | pH:tank),
                    data = bw_ssid, REML = TRUE)

# reduced model, dropped tank as factor
lm_ssid_noT <- lmer(growthadj ~ pH * site + 
                      (1 | site:genotype),
                    data = bw_ssid, REML = TRUE)

# LRT: genotype effect
lrt_geno_ssid <- anova(update(lm_ssid, REML = FALSE), update(lm_ssid_noG, REML = FALSE))
lrt_geno_ssid # significant p value in the final (interaction) model, so proceed with genotype term

# LRT: tank effect
lrt_tank_ssid <- anova(update(lm_ssid, REML = FALSE), update(lm_ssid_noT, REML = FALSE))
lrt_tank_ssid # significant p value in the final (interaction) model, so proceed with tank
```


### Tank-level check of pH effect

```{r tank-check-ssid}
# sanity check: is there a pH effect when looking at growth just across tanks?
tank_means_ssid <- aggregate(growthadj ~ pH + tank, bw_ssid, mean)
summary(lm(growthadj ~ pH, data = tank_means_ssid)) # nonsignificant; matches mixed-model pH test, confirming tank is the unit of replication for pH
```


### Variance components and R²

```{r variance-ssid}
# Random effects variance components
# NOTE: pct_nonfixed_var = % of random + residual variance only (fixed-effect variance excluded)
vc_ssid <- as.data.frame(VarCorr(lm_ssid))
total_var_ssid <- sum(vc_ssid$vcov)
vc_ssid$pct_nonfixed_var <- round(100 * vc_ssid$vcov / total_var_ssid, 2)
colnames(vc_ssid)[colnames(vc_ssid) == "grp"] <- "component"

# Fixed effects R²
r2_ssid <- r.squaredGLMM(lm_ssid)

# Individual fixed effect contributions
# interaction: R² of the final model minus R² without the interaction; main effects: dropped from the
# no-interaction model (marginality: a main effect can't be removed while its interaction stays in)
lm_ssid_ml       <- update(lm_ssid, REML = FALSE)
lm_ssid_noint_ml <- update(lm_ssid_noint, REML = FALSE)
lm_ssid_no_pH    <- update(lm_ssid_noint_ml, . ~ . - pH)
lm_ssid_no_site  <- update(lm_ssid_noint_ml, . ~ . - site)

r2_ssid_full    <- r.squaredGLMM(lm_ssid_ml)[, "R2m"]
r2_ssid_noint   <- r.squaredGLMM(lm_ssid_noint_ml)[, "R2m"]
delta_int_ssid  <- round(100 * (r2_ssid_full - r2_ssid_noint), 2)
delta_pH_ssid   <- round(100 * (r2_ssid_noint - r.squaredGLMM(lm_ssid_no_pH)[, "R2m"]), 2)
delta_site_ssid <- round(100 * (r2_ssid_noint - r.squaredGLMM(lm_ssid_no_site)[, "R2m"]), 2)
```


### Model summaries

```{r summary-ssid}
# --- Console output ---
summary(lm_ssid)
anova(lm_ssid)
print(vc_ssid[, c("component", "vcov", "pct_nonfixed_var")])
cat("Marginal R² (fixed effects):", round(100 * r2_ssid[, "R2m"], 2), "%\n")
cat("Conditional R² (fixed + random):", round(100 * r2_ssid[, "R2c"], 2), "%\n")
cat("Delta R² pH:", delta_pH_ssid, "%\n")
cat("Delta R² site:", delta_site_ssid, "%\n")
cat("Delta R² pH x site:", delta_int_ssid, "%\n")
```


### Export model outputs

```{r export-ssid}
# --- txt output ---
capture.output({
  cat("=== LRT: pH x Site Interaction (Fixed) ===\n")
  print(lrt_int_ssid)
  
  cat("\n=== LRT: Tank Effect (pH:tank) ===\n")
  print(lrt_tank_ssid)
  
  cat("\n=== LRT: Genotype Effect (site:genotype) ===\n")
  print(lrt_geno_ssid)
  
  cat("\n=== ANOVA (Fixed Effects) ===\n")
  print(anova(lm_ssid))
  
  cat("\n=== Variance Components (Random Effects) ===\n")
  cat("pct_nonfixed_var = % of random + residual variance (excludes fixed-effect variance)\n")
  print(vc_ssid[, c("component", "vcov", "pct_nonfixed_var")])
  
  cat("\n=== R² (Fixed Effects) ===\n")
  cat("Marginal R² (fixed effects only):", round(100 * r2_ssid[, "R2m"], 2), "%\n")
  cat("Conditional R² (fixed + random):", round(100 * r2_ssid[, "R2c"], 2), "%\n")
  cat("Delta R² pH:", delta_pH_ssid, "%\n")
  cat("Delta R² site:", delta_site_ssid, "%\n")
  cat("Delta R² pH x site:", delta_int_ssid, "%\n")
  
}, file = "../../outputs/growth/urban bw ssid lme.txt")
```


### Pairwise site comparisons

```{r pairwise-ssid}
# pairwise site tests within each pH treatment (final model includes the pH x site interaction)
emm_ssid <- emmeans(lm_ssid, ~ site | pH)
pairs(emm_ssid, adjust = "tukey")
capture.output(pairs(emm_ssid, adjust = "tukey"), file = "../../outputs/growth/urban bw ssid pairwise.txt")

# Create letters indicating significant differences for plot (separately for each pH panel)
cld_ssid <- cld(emm_ssid, adjust = "tukey", Letters = letters, alpha = 0.05)
cld_ssid

# pH effect within each site
emm_ssid_ph <- emmeans(lm_ssid, ~ pH | site)
pairs(emm_ssid_ph)
capture.output(pairs(emm_ssid_ph), file = "../../outputs/growth/urban bw ssid pH by site.txt")
```


# Urban vs reef contrasts

```{r urban-reef-contrasts}
# planned contrast from each final model: mean of the urban sites (Star Island, MacArthur North) minus mean of the
# reef sites (Emerald, Rainbow), averaged over pH. Tests the urban vs reef question with site kept in the model.
urban_reef_contrast <- list(`urban - reef` = c(-0.5, -0.5, 0.5, 0.5)) # site order: Emerald, Rainbow, Star, MacN
urban_reef_ofav <- contrast(emmeans(lm_ofav_noint, ~ site), method = urban_reef_contrast)
urban_reef_ssid <- contrast(emmeans(lm_ssid, ~ site), method = urban_reef_contrast) # averaged over pH
urban_reef_ssid_ph <- contrast(emmeans(lm_ssid, ~ site | pH), method = urban_reef_contrast) # within each pH (interaction model)
urban_reef_ofav
urban_reef_ssid
urban_reef_ssid_ph

capture.output({
  cat("=== O. faveolata: urban (Star, MacN) - reef (Emerald, Rainbow) ===\n")
  print(urban_reef_ofav)
  cat("\n=== S. siderea: urban (Star, MacN) - reef (Emerald, Rainbow) ===\n")
  print(urban_reef_ssid)
  cat("\n=== S. siderea: urban - reef within each pH treatment (pH x site interaction in the final model) ===\n")
  print(urban_reef_ssid_ph)
}, file = "../../outputs/growth/urban bw urban reef contrasts.txt")
```


# Model residual diagnostics

```{r model-residual-diagnostics}
# normality of the final models' residuals and random effects (the assumptions lmer actually makes),
# rather than of the raw response
resid_ofav <- resid(lm_ofav_noint)
resid_ssid <- resid(lm_ssid)

# Shapiro test on model residuals, p-values below 0.05 indicate violations of normality assumptions
shapiro.test(resid_ofav)
shapiro.test(resid_ssid)

# random effect estimates (BLUPs), few levels so interpret visually
re_ofav <- ranef(lm_ofav_noint)
re_ssid <- ranef(lm_ssid)

save_plot("../../outputs/growth/urban bw model residuals.pdf", width = 10, height = 8, {
par(mfrow = c(2, 4))
final_models <- list(ofav = lm_ofav_noint, ssid = lm_ssid)
for (sp in c("ofav", "ssid")) {
  m <- final_models[[sp]]
  r <- resid(m)
  re <- ranef(m)
  # residual normality
  qqnorm(r, main = paste(sp, "residuals Q-Q")); qqline(r)
  # homogeneity of variance
  plot(fitted(m), r, xlab = "Fitted values", ylab = "Residuals", main = paste(sp, "residuals vs fitted"))
  abline(h = 0, lty = 2)
  # random effect normality
  qqnorm(re$`site:genotype`[, 1], main = paste(sp, "genotype effects Q-Q")); qqline(re$`site:genotype`[, 1])
  qqnorm(re$`pH:tank`[, 1], main = paste(sp, "tank effects Q-Q")); qqline(re$`pH:tank`[, 1])
}
})

capture.output({
  cat("=== Shapiro-Wilk test on model residuals ===\n")
  cat("\nO. faveolata (lm_ofav_noint):\n")
  print(shapiro.test(resid_ofav))
  cat("S. siderea (lm_ssid, with pH x site interaction):\n")
  print(shapiro.test(resid_ssid))
}, file = "../../outputs/growth/urban bw model residuals.txt")
```


# Percent change in growth

```{r percent-change-in-growth}
# % change in mean growth under low pH relative to control pH: (low - control) / control * 100
# raw means from outlier-filtered data (bw_ofav, bw_ssid)
bw_models <- rbind(bw_ofav, bw_ssid)

growth_means <- bw_models %>%
  group_by(species, site, pH) %>%
  summarise(mean = mean(growthadj, na.rm = TRUE),
            sd = sd(growthadj, na.rm = TRUE),
            n = sum(!is.na(growthadj)),
            .groups = "drop")

# by species and site
pct_change_site <- growth_means %>%
  pivot_wider(names_from = pH, values_from = c(mean, sd, n)) %>%
  mutate(pct_change = round(100 * (`mean_low pH` - `mean_control pH`) / `mean_control pH`, 2))

# by species, pooled across sites
pct_change_species <- bw_models %>%
  group_by(species, pH) %>%
  summarise(mean = mean(growthadj, na.rm = TRUE),
            sd = sd(growthadj, na.rm = TRUE),
            n = sum(!is.na(growthadj)),
            .groups = "drop") %>%
  pivot_wider(names_from = pH, values_from = c(mean, sd, n)) %>%
  mutate(site = "all sites",
         pct_change = round(100 * (`mean_low pH` - `mean_control pH`) / `mean_control pH`, 2))

# by species and habitat (reef: Emerald, Rainbow; urban: Star, MacN); used where per-site low-pH n is small
pct_change_habitat <- bw_models %>%
  mutate(site = if_else(site %in% c("Star", "MacN"), "urban sites", "reef sites")) %>%
  group_by(species, site, pH) %>%
  summarise(mean = mean(growthadj, na.rm = TRUE),
            sd = sd(growthadj, na.rm = TRUE),
            n = sum(!is.na(growthadj)),
            .groups = "drop") %>%
  pivot_wider(names_from = pH, values_from = c(mean, sd, n)) %>%
  mutate(pct_change = round(100 * (`mean_low pH` - `mean_control pH`) / `mean_control pH`, 2))

pct_change <- bind_rows(mutate(pct_change_site, site = as.character(site)), pct_change_habitat, pct_change_species) %>%
  mutate(site = factor(site, levels = c("Emerald", "Rainbow", "Star", "MacN", "reef sites", "urban sites", "all sites"))) %>%
  arrange(species, site)
pct_change

write.csv(pct_change, file = "../../outputs/growth/urban bw pct change.csv", row.names = FALSE)

# % difference in mean growth of urban (Star Island, MacArthur North) vs reef (Emerald, Rainbow) fragments:
# (urban - reef) / reef * 100, fragment-weighted means from outlier-filtered data, by species and pH and pooled across pH
urban_reef <- function(dat) {
  dat %>%
    mutate(origin = if_else(site %in% c("Star", "MacN"), "urban", "reef")) %>%
    group_by(origin) %>%
    summarise(mean = mean(growthadj), n = n(), .groups = "drop") %>%
    pivot_wider(names_from = origin, values_from = c(mean, n)) %>%
    mutate(pct_urban_vs_reef = round(100 * (mean_urban - mean_reef) / mean_reef, 2))
}
pct_urban_reef <- bind_rows(
  bw_models %>% group_by(species, pH) %>% group_modify(~ urban_reef(.x)) %>% ungroup(),
  bw_models %>% group_by(species) %>% group_modify(~ urban_reef(.x)) %>% ungroup() %>% mutate(pH = "both pH")) %>%
  arrange(species, pH)
pct_urban_reef

write.csv(pct_urban_reef, file = "../../outputs/growth/urban bw pct urban reef.csv", row.names = FALSE)
```


# Plots

```{r plots}
fill.color<-c("#018571","#80cdc1","#dfc27d","#a6611a","grey") # custom color palette

# stats labels built from the final models so they stay in sync with the model outputs
# format: F(numerator df, total df = n - 1), p rounded to 1 significant figure (2 if it would round to 0.05), * if p < 0.05
stat_label <- function(model, term, label) {
  a <- anova(model)
  p <- a[term, "Pr(>F)"]
  p_digits <- if (signif(p, 1) == 0.05) 2 else 1
  p_txt <- if (p < 0.001) "p < 0.001" else paste0("p = ", signif(p, p_digits))
  paste0(label, ": F", a[term, "NumDF"], ",", nobs(model) - 1, " = ", sprintf("%.1f", a[term, "F value"]), ", ", p_txt, ifelse(p < 0.05, "*", ""))
}
# F values below 0.1 are shown with 2 decimals so they don't print as 0.0
stat_label2 <- function(model, term, label) {
  lab <- stat_label(model, term, label)
  f <- anova(model)[term, "F value"]
  if (f < 0.1) lab <- sub(sprintf("= %.1f,", f), sprintf("= %.2f,", f), lab, fixed = TRUE)
  lab
}

stats_ofav_ph <- data.frame(pH = "control pH", growthadj = 0.6, lab = stat_label2(lm_ofav_noint, "pH", "pH"))
stats_ofav_site <- data.frame(pH = "control pH", growthadj = 0.6, lab = stat_label2(lm_ofav_noint, "site", "site"))

stats_ssid_ph <- data.frame(pH = "control pH", growthadj = 0.6, lab = stat_label2(lm_ssid, "pH", "pH"))
stats_ssid_site <- data.frame(pH = "control pH", growthadj = 0.6, lab = stat_label2(lm_ssid, "site", "site"))
stats_ssid_int <- data.frame(pH = "control pH", growthadj = 0.6, lab = stat_label2(lm_ssid, "pH:site", "pH x site"))

ofav <- ggboxplot(bw_ofav,
                  x = "site",
                  y = "growthadj",
                  fill = "site",
                  width = 0.7,
                  size = 0.75,
                  title = "Orbicella faveolata",
                  legend = "none") +
  facet_grid(~pH) +
  xlab(NULL) +
  ylab("Growth Rate (g cm-2 yr-1)") +
  scale_fill_manual(values = fill.color) +
  geom_hline(yintercept=0, linetype="dashed", color = "black") +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_blank()) +
  geom_text(data = stats_ofav_ph,label = stats_ofav_ph$lab, aes(x=2, y=0.7)) +
  geom_text(data = stats_ofav_site,label = stats_ofav_site$lab, aes(x=2, y=0.65)) +
  geom_text(data=cld_ofav, aes(x = site, y=-0.2, label=.group)) +
  ylim(-0.2, 0.7) # lower limit covers Mo1-3 (-0.155); ylim() drops points outside it before the boxplot stats
ofav

ssid <- ggboxplot(bw_ssid,
                  x = "site",
                  y = "growthadj",
                  fill = "site",
                  width = 0.7,
                  size = 0.75,
                  title = "Siderastrea siderea",
                  legend = "none") +
  facet_grid(~pH) +
  xlab(NULL) +
  ylab("Growth Rate (g cm-2 yr-1)") +
  scale_fill_manual(values = fill.color) +
  geom_hline(yintercept=0, linetype="dashed", color = "black") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(data = stats_ssid_ph,label = stats_ssid_ph$lab, aes(x=2, y=0.7)) +
  geom_text(data = stats_ssid_site,label = stats_ssid_site$lab, aes(x=2, y=0.65)) +
  geom_text(data = stats_ssid_int,label = stats_ssid_int$lab, aes(x=2, y=0.6)) +
  geom_text(data=cld_ssid, aes(x = site, y=-0.2, label=.group)) +
  ylim(-0.2, 0.7) # lower limit covers Mo1-3 (-0.155); ylim() drops points outside it before the boxplot stats
ssid

# both panels
bw_species<-ggarrange(ofav,
                      ssid,
                      heights = c(3.9,4),
                      widths = c(4,4),
                      ncol = 1,
                      nrow = 2)
bw_species

ggsave("../../outputs/growth/urban bw.pdf", bw_species, width=8, height=8,dpi = 300)
```
